---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r readme-01, echo = FALSE}
knitr::opts_chunk$set(
  fig.asp = 2/3,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# ggpmisc #

[![cran version](https://www.r-pkg.org/badges/version/ggpmisc)](https://cran.r-project.org/package=ggpmisc) 

## Purpose

Package '**ggpmisc**' (Miscellaneous Extensions to 'ggplot2') is a set of
extensions to R package 'ggplot2' (>= 3.0.0) with emphasis on annotations and
highlighting related to fitted models and data summaries. Data summaries shown
as text, tables or equations are implemented. New geoms support insets in
ggplots. The location of fit summaries and graphical insets within the plotting
area needs usually to be set independently of the `x` and `y` scales. The
"natural" coordinates to use in such cases are expressed in 'grid' "npc" units
in the range [0..1] for which new aesthetics and their scales are made
available.

## ggplot methods

Being `ggplot()` defined as a generic method in 'ggplot2' makes it possible to
define specializations, and we provide two for time series stored in objects of
classes `ts` and `xts` which automatically convert these objects into tibbles
and set the as default the aesthetic mappings for `x` and `y`. A companion
function `try_tibble()` is also exported.

## Geometries

Geometries `geom_table()`, `geom_plot()` and `geom_grob()` make it possible to
add inset tables, inset plots, and arbitrary 'grid' graphical objects as layers
to a ggplot using native coordinates for `x` and `y`. 

Geometries `geom_text_npc()`, `geom_label_npc()`, `geom_table_npc()`,
`geom_plot_npc()` and `geom_grob_npc()`, `geom_text_npc()` and
`geom_label_npc()` are versions of geometries that interpret positions on _x_
and _y_ axes using aesthetics `npcx` and `npcy` values expressed in "npc" units.

Geometries `geom_x_margin_arrow()`, `geom_y_margin_arrow()`,
`geom_x_margin_grob()`, `geom_y_margin_grob()`, `geom_x_margin_point()` and
`geom_y_margin_point()` make it possible to add marks along the _x_ and _y_
axes. `geom_vhlines()` and `geom_quadrant_lines()` draw vertical and horizontal
reference lines within a single layer.

## Statistics

Statistic `stat_fmt_tb()` helps with the formatting of tables to be plotted
with `geom_table()`. 

Statistics `stat_peaks()` and `stat_valleys()` can be used to highlight and/or
label maxima and minima in a plot.

Statistics that help with reporting the results of model fits are
`stat_poly_eq()`, `stat_fit_residuals()`, `stat_fit_deviations()`,
`stat_fit_glance()`, `stat_fit_augment()`, `stat_fit_tidy()` and
`stat_fit_tb()`.

Two statistics, `stat_dens2d_filter()` and `stat_dens2d_label()`, implement
tagging or selective labelling of observations based on the local 2D density of
observations. These two stats are designed to work well together with
`geom_text_repel()` and `geom_label_repel()` from package 'ggrepel'.

A summary statistic using special grouping for quadrants
`stat_quadrant_counts()` can be used to automate labelling with the number of
observations.

The statistics `stat_apply_panel()` and `stat_apply_group()` can be useful for
applying arbitrary functions returning numeric vectors. They are specially
useful with functions lime `cumsum()`, `cummax()` and `diff()`.

## Aesthetics and scales

Scales `scale_npcx_continuous()` and  `scale_npcy_continuous()` and the
corresponding new aesthetics `npcx` and `npcy` make it possible to add graphic
elements and text to plots using coordinates expressed in `npc` units for the
location within the plotting area, improving support for annotations, most
notably when using facets.

Scales `scale_x_logFC()` and `scale_y_logFC()` are suitable for plotting of log
fold change data. Scales `scale_x_Pvalue()`, `scale_y_Pvalue()`, `scale_x_FDR()`
and `scale_y_FDR()` are suitable for plotting _p_-values and adjusted _p_-values
or false discovery rate (FDR). Default arguments are suitable for volcano and
quadrant plots as used for transcriptomics, metabolomics and similar data. 

Scales `scale_colour_outcome()`, `scale_fill_outcome()` and 
`scale_shape_outcome()` and functions `outome2factor()`, `threshold2factor()`,
`xy_outcomes2factor()` and `xy_thresholds2factor()` used together make it 
easy to map ternary numeric outputs and logical binary outcomes to colour,
fill and shape aesthetics. Default arguments are suitable for volcano,
quadrant and other plots as used for genomics, metabolomics and similar data.

## MIGRATED

Functions for the manipulation of layers in ggplot objects and statistics and
geometries that echo their data input to the R console, earlier included in this
package are now in package 'gginnards'.
[![cran version](https://www.r-pkg.org/badges/version/gginnards)](https://cran.r-project.org/package=gginnards) 

## Examples

```{r readme-02, message=FALSE}
library(ggpmisc)
library(ggrepel)
```

In the first example we plot a time series using the specialized version of
`ggplot()` that converts the time series into a tibble and maps the `x`
and `y` aesthetics automatically. We also highlight and label the peaks
using `stat_peaks`.

```{r readme-03}
ggplot(lynx, as.numeric = FALSE) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", angle = 66,
             hjust = -0.1, x.label.fmt = "%Y") +
  stat_peaks(geom = "rug", colour = "red", sides = "b") +
  expand_limits(y = 8000)
```

In the second example we add the equation for a fitted polynomial plus the
adjusted coefficient of determination to a plot showing the observations plus
the fitted curve, deviations and confidence band. We use `stat_poly_eq()`.

```{r readme-04}
formula <- y ~ x + I(x^2)
ggplot(cars, aes(speed, dist)) +
  geom_point() +
  stat_fit_deviations(method = "lm", formula = formula, colour = "red") +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label =  paste(stat(eq.label), stat(adj.rr.label), sep = "*\", \"*")),
               formula = formula, parse = TRUE)
```

The same figure as in the second example but this time annotated with the
ANOVA table for the model fit. We use `stat_fit_tb()` which can be used to add
ANOVA or summary tables.

```{r readme-05}
formula <- y ~ x + I(x^2)
ggplot(cars, aes(speed, dist)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_tb(method = "lm",
              method.args = list(formula = formula),
              tb.type = "fit.anova",
              tb.vars = c(Effect = "term", 
                          "df",
                          "M.S." = "meansq", 
                          "italic(F)" = "statistic", 
                          "italic(P)" = "p.value"),
              label.y.npc = "top", label.x.npc = "left",
              size = 2.5,
              parse = TRUE)
```

A plot with an inset plot.

```{r, readme-06}
library(tibble)
p <- ggplot(mtcars, aes(factor(cyl), mpg, colour = factor(cyl))) +
  stat_boxplot() +
  labs(y = NULL) +
  theme_bw(9) + theme(legend.position = "none")
df <- tibble(x = 0.01, y = 0.015, plot = list(p))
ggplot(mtcars, aes(wt, mpg, colour = factor(cyl))) +
  geom_point() +
  geom_plot_npc(data = df, mapping = aes(npcx = x, npcy = y, label = plot),
                vjust = 0, hjust = 0) +
  expand_limits(y = 0, x = 0)
```

A quadrant plot with counts and labels, using `geom_text_repel()` from package
'ggrepel'.

```{r}
ggplot(quadrant_example.df, aes(logFC.x, logFC.y)) +
  geom_point(alpha = 0.3) +
  geom_quadrant_lines() +
  stat_quadrant_counts() +
  stat_dens2d_filter(color = "red", keep.fraction = 0.03) +
  stat_dens2d_labels(aes(label = gene), keep.fraction = 0.03, 
                     geom = "text_repel", size = 2, colour = "red") +
  scale_x_logFC(name = "Transcript abundance after A%unit") +
  scale_y_logFC(name = "Transcript abundance after B%unit")
```

## Installation

Installation of the most recent stable version from CRAN:

```{r cran-instalaltion, eval=FALSE}
install.packages("ggpmisc")
```

Installation of the current unstable version from Bitbucket:

```{r bb-instalaltion, eval=FALSE}
# install.packages("devtools")
devtools::install_bitbucket("aphalo/ggpmisc")
```

## Documentation 

HTML documentation is available at (https://docs.r4photobiology.info/ggpmisc/), including a _User Guide_.

News about updates are regularly posted at (https://www.r4photobiology.info/).

## Contributing

Please report bugs and request new features at (https://bitbucket.org/aphalo/ggpmisc/issues). Pull requests are welcome at (https://bitbucket.org/aphalo/ggpmisc).

## Citation

If you use this package to produce scientific or commercial publications, please cite according to:

```{r}
citation("ggpmisc")
```

## License

© 2016-2020 Pedro J. Aphalo (pedro.aphalo@helsinki.fi). Released under the GPL, version 2 or greater. This software carries no warranty of any kind.
