---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.asp = 2/3,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# ggpmisc #

[![](http://www.r-pkg.org/badges/version/ggpmisc)](https://cran.r-project.org/package=ggpmisc) 
[![](http://cranlogs.r-pkg.org/badges/ggpmisc)](https://cran.r-project.org/package=ggpmisc) 
[![](http://cranlogs.r-pkg.org/badges/grand-total/ggpmisc)](https://cran.r-project.org/package=ggpmisc)

Package '**ggpmisc**' (Miscellaneous Extensions to 'ggplot2') is a set of
extensions to R package 'ggplot2' (>= 3.0.0) with emphasis on annotations and
highlighting related to fitted models and data summaries. To complement these,
the widely useful `geom_table()` and `stat_fmt_tb()` are defined as well as
`ggplot` constructors for time series objects. The provided `ggplot.ts()` and
`ggplot.xts()` use `try_tibble()` which is also exported and accepts objects of
additional classes as input.

Statistics useful for highlighting and/or annotating individual data points 
in regions of plot panels with high/low densities of observations. These stats are designed to work
well together with `geom_text_repel()` and `geom_label_repel()` from package 
'ggrepel'.

Functions for the manipulation of layers in ggplot objects and statistics and
geometries that echo their data input to the R console, earlier included in this
package are now in package 'gginnards'.

## Examples

This is a basic example which shows plotting using the most automatic approach.

```{r example, message=FALSE}
library(photobiology)
library(ggpmisc)
```

```{r}
ggplot(lynx, as.numeric = FALSE) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", angle = 66,
             hjust = -0.1, x.label.fmt = "%Y") +
  ylim(NA, 8000)
```


```{r}
formula <- y ~ poly(x, 2, raw = TRUE)
ggplot(cars, aes(speed, dist)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label =  paste(..eq.label.., ..adj.rr.label.., sep = "~~~~")),
               formula = formula, parse = TRUE)
```

```{r}
formula <- y ~ x + I(x^2)
ggplot(cars, aes(speed, dist)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_tb(method = "lm",
              method.args = list(formula = formula),
              tb.type = "fit.anova",
              tb.vars = c(Effect = "term", 
                          "df",
                          "M.S." = "meansq", 
                          "italic(F)" = "statistic", 
                          "italic(P)" = "p.value"),
               label.y.npc = "top", label.x.npc = "left",
              parse = TRUE)
```

## Installation

Installation of the most recent stable version from CRAN:

```{r cran-instalaltion, eval=FALSE}
install.packages("ggspectra")
```

Installation of the current unstable version from Bitbucket:

```{r bb-instalaltion, eval=FALSE}
# install.packages("devtools")
devtools::install_bitbucket("aphalo/ggspectra")
```

## Documentation 

HTML documentation is available at (http://docs.r4photobiology.info/ggpmisc/), including a _User Guide_.

News on updates to the different packages of the 'r4photobiology' suite are regularly posted at (https://www.r4photobiology.info/).

## Contributing

Please report bugs and request new features at (https://bitbucket.org/aphalo/ggpmisc/issues). Pull requests are welcome at (https://bitbucket.org/aphalo/ggpmisc).

Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

## Citation

If you use this package to produce scientific or commercial publications, please cite according to:

```{r}
citation("ggpmisc")
```

## License

© 2016-2018 Pedro J. Aphalo (pedro.aphalo@helsinki.fi). Released under the GPL, version 2 or greater. This software carries no warranty of any kind.
