---
title: "Annotations are not data labels"
subtitle: "On how 'ggpmisc' extends the grammar of graphics"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Annotations are not data labels}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = 'center', 
                      fig.show = 'hold', fig.width = 7, fig.height = 4)
```

## Data labels, annotations and insets

*Data labels* add textual information directly related to individual
data points (shown as glyphs). Text position in this case is dependent
on the scales used to represent data points. Text is usually displaced
so that it does not occlude the glyph representing the data point and
when the link to the data point is unclear, this link is signaled with a
line segment or arrow. Data labels are distinct from annotations but
instead contribute to the representation of data on a plot or map.

*References* are lines, shading or marks used to help the reading of a
plot. These elements are used to highlight specific values on an axis or
a region in a plot. They, like data labels, are positioned relative to
the scales used for data. The position of data labels and lines, glyphs
or shading used as reference cannot be altered by the designer of a
plot, as the position conveys information. I will use the term *data
labels* irrespective if the "labels" are textual or graphical, like
icons and small plots and simple tables linked to data points or map
features.

According to Koponen and Hildén (2019), "[in a statistical chart]
annotations can be used to draw reader attention to relevant detail".
These authors use as an example a text box in a plot to highlight a data
point that is off-scale and has been "squeezed" to a position
immediately outside the plotting area.

*Annotations* differ from data labels, in that their position is
decoupled from their meaning. Insets can be thought as larger, but still
self-contained annotations. In most cases the reading of inset tables
and plots depends only weakly on the plot or map in which they are
included.

In the case of annotations and insets the designer of a data
visualization has the freedom to locate them anywhere, as long they do
not occlude features used to describe data. I will use the term
*annotation* irrespective if the "labels" are textual or graphical.
Insets are similar to annotations, but the term *inset* is used would
the inset graphical or textual element is complex, and usually lager
than an annotation. Insets can be either positioned within the main
plotting area, or adjacent to it, usually as a smaller panel, without
any loss of meaning .

That the position of annotations and insets is independent of the
plotted data cannot be expressed using the grammar of graphics (GG) as
implemented in package 'ggplot2'.

## Annotations in 'ggplot2'

In *The Layered Grammar of Graphics* (Wickham 2000) as implemented in
package 'ggplot2' (Wickham et al. 2016) annotations are "second class",
but still data-dependent features. As layers they behave differently
than data layers and ignore faceting into panels. So `annotate()`
disconnects the resulting plot elements from the data source, but not
from the scales used to graphically map the data values.

In our view, the main practical difference between data labels and
annotations is the scales used to position them. Instead of using an
`annotate()` function to implement annotations, we could retain the use
of GG but based on native plot coordinates (npc). Support of annotations
done in this way would allow "annotation layers" to behave almost
identically to "data layers". The *X* and *y* position aesthetics used
for data could be supplemented with pseudo-aesthetics without any
translation relative to the native plotting coordinates of the plotting
area or viewport. Doing so, would allow the graphic design flexibility
conceptually inherent to annotations.

## Extending 'ggplot2'

Based on this insight, a new approach to adding annotations and insets
was implemented in package 'ggpmisc' through two new *x* and *y*
pseudo-aesthetics, *npcx* and *npcy*, and corresponding dumb scales and
various geometries that make use of them. These scales and geometries
were implemented as an extension to package 'ggplot2'.

Before showing how this works in practice, we load the package and set an
uncluttered theme as default.

```{r}
library(ggpmisc)
theme_set(theme_bw() + theme(panel.grid = element_blank()))
```

Native plot coordinates have range 0..1 but the "npc" geoms also recognize some
positions by name with the same text strings as used for text justification in
'ggplot2'. The default justification is `"inward"`.

```{r}
p1 <- ggplot(mtcars, aes(factor(cyl), mpg)) +
  geom_point() +
  geom_text_npc(data = data.frame(x = c("left", "left"),
                                  y = c("top", "bottom"),
                                  label = c("Most\nefficient",
                                            "Least\nefficient")),
                mapping = aes(npcx = x, npcy = y, label = label),
                size = 3)
p1

```

The advantage of this approach becomes apparent when we modify the limits of the
_y_ scale. We do not show it here, but when a script or user defined plotting
function sets the scale limits based on the input data, in the absence of the
extensions proposed here, setting annotations consistently within the plotting
area becomes rather complex.

```{r}
p1 + expand_limits(y = 0)
```

So as to support the existing syntax for annotations using the new geometries,
function `ggplot2::annotate()` gets overridden when package 'ggpmisc' is loaded.
The new definition adds support for the new pseudo-aesthetics *npcx* and *npcy*
retaining the old behaviour otherwise.

```{r}
ggplot(mtcars, aes(factor(cyl), mpg)) +
  geom_point() +
  annotate(geom = "text_npc",
           npcx = c("left", "left"),
           npcy = c("top", "bottom"),
           label = c("Most\nefficient",
                     "Least\nefficient"),
           size = 3)
```

Inset plots can be added with the same syntax using `geom_plot_npc()`. They can
be thought also an being annotations.

```{r}
p2 <- ggplot(mtcars, aes(factor(cyl), mpg, colour = factor(cyl))) +
  stat_boxplot() +
  labs(y = NULL) +
  theme_bw(9) + 
  theme(legend.position = "none",
        panel.grid = element_blank())

ggplot(mtcars, aes(wt, mpg, colour = factor(cyl))) +
  geom_point() +
  annotate("plot_npc", npcx = "left", npcy = "bottom", label = p2) +
  expand_limits(y = 0, x = 0)

```

An example with facets, labeling of panels in a more traditional way.
Panel tags within the plotting are at same "npc" location with free
scale limits in panels.

```{r}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  geom_text_npc(data = data.frame(cyl = levels(factor(mtcars$cyl)),
                                  label = LETTERS[seq_along(levels(factor(mtcars$cyl)))],
                                  x = 0.90,
                                  y = 0.95),
                 mapping = aes(npcx = x, npcy = y, label = label),
                size = 4) +
  facet_wrap(~factor(cyl), scales = "free") +
  theme(strip.background = element_blank(),
        strip.text = element_blank())
```

This approach was first implemented in 'ggpmisc' version 0.3.1 released
in April 2919. As of version >= 0.3.8, the implementation can be
considered fairly stable. However, this implementation is to an extent
dependent on undocumented behaviour of 'ggplot2' functions, which means
that future updates to 'ggplot2' can break this functionality.

## When to use annotations and insets

Package 'ggpmisc' adds support for various plot annotations and reference
guides. It also adds support for some data labels related to data features.
While developing these statistics, it became clear that expanding GG's support
for annotations would simplify the code of these statistics considerably and
also more cleanly separate the computations on data from the positioning of
annotations.

Geometries useful for data labels and annotations.

| Data labels (data coordinates) | Annotations (npc)  | Mapped to `label` aesthetic |
|--------------------------------|--------------------|---------------------------|
| `ggplot2::geom_text()`         | `geom_text_npc()`  | `character` vector        |
| `ggplot2::geom_label()`        | `geom_label_npc()` | `character` vector        |
| `geom_plot()`                  | `geom_plot_npc()`  | `list()` of `ggplot2::gg` |
| `geom_table()`                 | `geom_table_npc()` | `list()` of `data.frame`  |
| `geom_grob()`                  | `geom_grob_npc()`  | `list()` of `grid::grob`  |

To decide on the best approach, consider if the location of an element of a plot
is more "naturally" expressed in the original data units or as position relative
to the edges or centre of the plotting area. In the second case, prefer the
"npc" geoms as you are dealing with _annotations_, otherwise, use the ordinary
geometries as you are dealing with data labels or data points.

## Thoughts for the future

Although we have discussed only scales for *x* and *y* aesthetics, the
same consideration applies to other aesthetics like *colour*. It seems
generally useful to allow aesthetics to use a different scale for
annotations than for data. Defining pseudo-aesthetics for annotations
instead of allowing multiple scales for each aesthetic in a plot would
add flexibility and still keep the guarantee that the meaning of
aesthetics values remains consistent across all plot elements
representing data.

## List of references

Koponen, J; Hildén, J. (2019) *Data visualization handbook*. Aalto ARTS
books, Espoo. ISBN 978-952-60-7449-8.

Wickham H. (2010) A Layered Grammar of Graphics. *Journal of
Computational and Graphical Statistics* **19:** 3--28.

Wickham H. (2016) *ggplot2*. Springer International Publishing. ISBN
978-3-319-24275-0.
