---
title: "Annotations are not data labels"
subtitle: "On how 'ggpmisc' extends the grammar of graphics"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Annotations are not data labels}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = 'center', 
                      fig.show = 'hold', fig.width = 7, fig.height = 4)
```

## Data labels, annotations and insets

*Data labels* add textual information directly related to individual
data points (shown as glyphs). Text position in this case is dependent
on the scales used to represent data points. Text is usually displaced
so that it does not occlude the glyph representing the data point and
when the link to the data point is unclear, this link is signaled with a
line segment or arrow. Data labels are distinct from annotations but
instead contribute to the representation of data on a plot or map.

*References* are lines, shading or marks used to help the reading of a
plot. These elements are used to highlight specific values on an axis or
a region in a plot. They, like data labels, are positioned relative to
the scales used for data. The position of data labels and lines, glyphs
or shading used as reference cannot be altered by the designer of a
plot, as the position conveys information. I will use the term *data
labels* irrespective if the "labels" are textual or graphical, like
icons and small plots and simple tables linked to data points or map
features.

According to Koponen and Hildén (2019), "[in a statistical chart]
annotations can be used to draw reader attention to relevant detail".
These authors use as an example a text box in a plot to highlight a data
point that is off-scale and has been "squeezed" to a position
immediately outside the plotting area.

*Annotations* differ from data labels, in that their position is
decoupled from their meaning. Insets can be thought as larger, but still
self-contained annotations. In most cases the reading of inset tables
and plots depends only weakly on the plot or map in which they are
included.

In the case of annotations and insets the designer of a data visualization has
the freedom to locate them anywhere, as long they do not occlude features used
to describe data. I will use the term *annotation* irrespective if the "labels"
are textual or graphical. Insets are similar to annotations, but the term
*inset* is used when an annotation's graphical or textual element is complex and
occupies more space within the plotting area. Insets can be moved from within
the main plotting area to being adjacent to it, e.g., as a smaller panel,
without any loss of meaning .

That the position of annotations and insets is independent of the
plotted data cannot be expressed using the grammar of graphics (GG) as
implemented in package 'ggplot2'.

## Annotations in 'ggplot2'

In *The Layered Grammar of Graphics* (Wickham 2000) as implemented in package
'ggplot2' (Wickham et al. 2016) annotations are "second class" features. As
layers they behave differently than data layers: Only constant values can be
mapped to aesthetics and do no support faceting into panels. In escence
`annotate()` disconnects the resulting plot elements from the data source and
facetting, but not from the scales used to graphically display the data values.

From the data visualization perspective the main practical and conceptual
difference between data labels and annotations is in the scales used to position
them within the plotting area. Instead of using an `annotate()` function that
deviates from the grammar of graphics to implement annotations, we could retain
the use of the grammar of graphics for annotations but add support of native
plot coordinates (npc). Support of annotations done in this way would allow
"annotation layers" to behave almost identically to "data layers", and use the
same grammar. The *x* and *y* position aesthetics used for data could be
supplemented with pseudo-aesthetics without any translation relative to the
native plotting coordinates of the plotting area or viewport. Doing so, would
allow the graphic design flexibility conceptually inherent to annotations within
a user-friendly syntax.

## Extending 'ggplot2'

Based on this insight, a new approach to adding annotations and insets
was implemented in package 'ggpmisc' (>= 0.3.1) through two new *x* and *y*
pseudo-aesthetics, *npcx* and *npcy*, and corresponding dumb scales and
various geometries that make use of them. These scales and geometries
were implemented as an extension to package 'ggplot2'.

Before showing how this works in practice, we load the package and set an
uncluttered theme as default.

```{r, message=FALSE}
library(ggpmisc)
theme_set(theme_bw() + theme(panel.grid = element_blank()))
```

Native plot coordinates have range 0..1. However, the "npc" geoms also recognize
some positions by name with the same text strings as used for text justification
in 'ggplot2'. The default justification is `"inward"`, as this protects from
clipping at the edges the plotting area irrespective the arguments passed to the
`witdth` and `height` parameters of R's graphic devices.

```{r}
p1 <- ggplot(mtcars, aes(factor(cyl), mpg)) +
  geom_point() +
  geom_text_npc(data = data.frame(x = c("left", "left"),
                                  y = c("top", "bottom"),
                                  label = c("Most\nefficient",
                                            "Least\nefficient")),
                mapping = aes(npcx = x, npcy = y, label = label),
                size = 3)
p1

```

The advantage of this approach becomes apparent when the limits of the _y_ scale
are unknown or vary. When a script or user defined plotting function sets the
scale limits based on the input data, in the absence of the extensions proposed
here, setting annotations consistently within the plotting area becomes
laborious. The example below, shows how the annotations remain at the desired
position when the _y_ scale limts are expanded.

```{r}
p1 + expand_limits(y = 0)
```

To support the existing syntax for annotations using the new geometries,
function `ggplot2::annotate()` is overridden when package 'ggpmisc' is loaded.
The new definition adds support for the new pseudo-aesthetics *npcx* and *npcy*
retaining its original 'ggplot2' behaviour in all other respects.

```{r}
ggplot(mtcars, aes(factor(cyl), mpg)) +
  geom_point() +
  annotate(geom = "text_npc",
           npcx = c("left", "left"),
           npcy = c("top", "bottom"),
           label = c("Most\nefficient",
                     "Least\nefficient"),
           size = 3)
```

Inset plots can be added with the same syntax using `geom_plot_npc()`. They can
be thought also as being annotations. Here we use `annotate()` but `geom_plot()`
can be also used directly, in which case the inset plots can be different for
each panel.

```{r}
p2 <- ggplot(mtcars, aes(factor(cyl), mpg, colour = factor(cyl))) +
  stat_boxplot() +
  labs(y = NULL) +
  theme_bw(9) + 
  theme(legend.position = "none",
        panel.grid = element_blank())

ggplot(mtcars, aes(wt, mpg, colour = factor(cyl))) +
  geom_point() +
  annotate("plot_npc", npcx = "left", npcy = "bottom", label = p2) +
  expand_limits(y = 0, x = 0)

```

A simple example with facets, labeling of panels in a traditional way as
requiered by some book and journal styles. In this case panel tags within the
plotting are at same "npc" location with free scale limits in panels.

```{r}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  geom_text_npc(data = data.frame(cyl = levels(factor(mtcars$cyl)),
                                  label = LETTERS[seq_along(levels(factor(mtcars$cyl)))],
                                  x = 0.90,
                                  y = 0.95),
                mapping = aes(npcx = x, npcy = y, label = label),
                size = 4) +
  facet_wrap(~factor(cyl), scales = "free") +
  theme(strip.background = element_blank(),
        strip.text = element_blank())
```

This approach was first implemented in 'ggpmisc' version 0.3.1 released
in April 2919. As of version >= 0.3.8, the implementation can be
considered fairly stable. However, this implementation is to an extent
dependent on undocumented behaviour of 'ggplot2' functions, which means
that future updates to 'ggplot2' can break this functionality.

## When to use annotations and insets

Package 'ggpmisc' adds support for various plot annotations and reference
guides. It also adds support for some data labels related to data features.
While developing these statistics, it became clear that expanding the grammar of
graphics's support for annotations would simplify the code of these statistics
considerably and also more cleanly separate the computations on data from the
positioning of annotations.

**Table**. Geometries useful for data labels and annotations. Currently 
implemented _ordinary_ geometries and their _npc_ versions. The rightmost
column shows the expected class of the objects mapped to the `label` aesthetic.
When using `annotate()` to add a single plot, table or grob as an inset, 
enclosing them in a list is allowed, but not a requirement.

| Data labels (data coordinates) | Annotations (npc)  | Mapped to `label` aesthetic |
|--------------------------------|--------------------|---------------------------|
| `ggplot2::geom_text()`         | `geom_text_npc()`  | `character` vector        |
| `ggplot2::geom_label()`        | `geom_label_npc()` | `character` vector        |
| `geom_plot()`                  | `geom_plot_npc()`  | `list()` of `ggplot2::gg` |
| `geom_table()`                 | `geom_table_npc()` | `list()` of `data.frame`  |
| `geom_grob()`                  | `geom_grob_npc()`  | `list()` of `grid::grob`  |

When adding an informative element to a plot, assess whether it is an
_annotation_ or a _data label_. To decide on the best approach, consider if the
location of the element is more "naturally" expressed in the original data units
or as position relative to the edges or centre of the plotting area. In the
second case, prefer the "npc" geoms as you are dealing with _annotations_,
otherwise, use the ordinary geometries as you are dealing with data labels or
data points.

## Thoughts for the future

Although we have discussed only scales for *x* and *y* aesthetics, the
same consideration applies to other aesthetics like *colour*. It seems
generally useful to allow aesthetics to use a different scale for
annotations than for data. Defining pseudo-aesthetics for annotations
instead of allowing multiple scales for each aesthetic in a plot would
add flexibility while still keeping the guarantee that the meaning of
aesthetics values remains consistent across all plot elements
representing data.

## List of references

Koponen, J; Hildén, J. (2019) *Data visualization handbook*. Aalto ARTS
books, Espoo. ISBN 978-952-60-7449-8.

Wickham H. (2010) A Layered Grammar of Graphics. *Journal of
Computational and Graphical Statistics* **19:** 3--28.

Wickham H. (2016) *ggplot2*. Springer International Publishing. ISBN
978-3-319-24275-0.
