---
title: "Debugging ggplots"
subtitle: "`ggpmisc` `r packageVersion('ggpmisc')`"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Debugging ggplots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path = 'figure/debug-pos-', fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4)
options(warnPartialMatchArgs = FALSE)
```

## Preliminaries

```{r}
library(ggpmisc)
# library(ggplot2)
library(tibble)
```

We generate some artificial data.

```{r}
set.seed(4321)
# generate artificial data
x <- 1:100
y <- (x + x^2 + x^3) + rnorm(length(x), mean = 0, sd = mean(x^3) / 4)
my.data <- data.frame(x, 
                      y, 
                      group = c("A", "B"), 
                      y2 = y * c(0.5,2),
                      block = c("a", "a", "b", "b"))
```

## Introduction

The motivation for writing these stats and geoms is that at the moment it is in
many cases not possible to set breakpoints inside the code of stats and geoms.
This can make it tedious to see how these functions work, as one may need to add
`print` statements to their source code to achieve this. I wrote these functions
as tools to help in the development of this package itself, and as a way of
learning myself how data are passed around within the different components of a
`ggplot` object when it is printed.

The stats described in this vignette are very simple and print a summary of 
their `data` input to the console. In addition they also return a data frame
containing labels suitable for plotting with geom "text" or geom "label".
However, starting from version 0.2.7 of the package the default geom is "null".
The values are listed to the console at the time when the `ggplot` object is
printed.

As shown here, no other _geom_ or _stat_ is required, however in the remaining
examples we include `geom_point()` to make the data on the plot visible.


```{r}
ggplot(my.data, aes(x, y)) + stat_debug_group()
```
 
In the absence of facets or groups we get just get the summary from one data 
frame.

```{r}
ggplot(my.data, aes(x, y)) + geom_point() + stat_debug_group()
```

```{r}
ggplot(my.data, aes(x, y)) + geom_point() + stat_debug_panel()
```

In the case of grouping then one data frame is summarized for each group
in the ggplot object.

```{r}
ggplot(my.data, aes(x, y, colour = group)) + geom_point() + 
  stat_debug_group()
```

Without facets, we still have only one panel.

```{r}
ggplot(my.data, aes(x, y, colour = group)) + geom_point() + 
  stat_debug_panel()
```

The data are similar, except for the column named after the aesthetic, for the aesthetics used for grouping.

```{r}
ggplot(my.data, aes(x, y, shape = group)) + geom_point() + 
  stat_debug_group()
```

If we use as _geom_ `"label"` or `"text"` a _debug_ summary is added
to the plot itself, we can use other arguments valid for the _geom_ used,
in this case `vjust`.

```{r}
ggplot(my.data, aes(x, y, shape = group)) + geom_point() + 
  stat_debug_group(geom = "label", vjust = c(-0.5,1.5))
```

The summary function can be a user defined one, which allows lots of flexibility.

```{r}
ggplot(my.data, aes(x, y)) + geom_point() + 
  stat_debug_group(summary.fun = summary)
```

```{r}
ggplot(my.data, aes(x, y)) + geom_point() + 
  stat_debug_group(summary.fun = head)
```

```{r}
ggplot(my.data, aes(x, y)) + geom_point() + 
  stat_debug_group(summary.fun = nrow)
```

The default.

```{r}
ggplot(my.data, aes(x, y)) + geom_point() + 
  stat_debug_group(summary.fun = as_data_frame)
```

```{r}
ggplot(my.data, aes(x, y)) + geom_point() + 
  stat_debug_group(summary.fun = head, summary.fun.args = list(n = 3))
```

This next chunk showing how to print the whole data frame is not run as its output is more than 100 lines long as the data set contains 100 observations.

```{r, eval=FALSE}
ggplot(my.data, aes(x, y)) + geom_point() + 
  stat_debug_group(summary.fun = function(x) {x})
```

With grouping, for each _group_ the `compute_group()` function is called with a subset of the data.

```{r}
ggplot(my.data, aes(x, y, colour = group)) + geom_point() + 
  stat_debug_group(summary.fun = head, summary.fun.args = list(n = 3))
```

In this example with grouping and facets, within each _panel_ the `compute_group()` function is called for each group, in total four times.

```{r}
ggplot(my.data, aes(x, y, colour = group)) + geom_point() + 
  stat_debug_group(summary.fun = nrow) +
  facet_wrap(~block)
```

With facets, for each _panel_ the `compute_panel()` function is called with a subset of the data that is not split by groups. For our example, it is called twice.

```{r}
ggplot(my.data, aes(x, y, colour = group)) + geom_point() + 
  stat_debug_panel(summary.fun = nrow) +
  facet_wrap(~block)
```

Finally we show how `geom_debug()` can be used. First to print to the console the data as passed to geoms.

```{r}
ggplot(my.data, aes(x, y, colour = group)) + geom_point() + 
  geom_debug(summary.fun = head)
```

And also to print to the console the data returned by a stat. 

```{r}
ggplot(my.data, aes(x, y, colour = group)) + geom_point() + 
  stat_smooth(method = "lm",
             geom = "debug", 
             summary.fun = as_data_frame, 
             summary.fun.args = list())
```

```{r}
ggplot(my.data, aes(x, y, colour = group)) + geom_point() + 
  stat_peaks(span = NULL,
             geom = "debug", 
             summary.fun = as_data_frame, 
             summary.fun.args = list())
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  stat_fit_residuals(formula = formula, 
                     geom = "debug",
                     summary.fun = as_data_frame, 
                     summary.fun.args = list())
```

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_fit_augment(method = "lm", 
                   method.args = list(formula = formula),
                   geom = "debug",
                   summary.fun = tibble::as_data_frame, 
                   summary.fun.args = list()) +
  stat_fit_augment(method = "lm", 
                   method.args = list(formula = formula),
                   geom = "smooth")
```

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y2, colour = group)) +
  geom_point() +
  stat_fit_augment(method = "lm", 
                   method.args = list(formula = formula),
                   geom = "debug",
                   summary.fun = tibble::as_data_frame, 
                   summary.fun.args = list()) +
  stat_fit_augment(method = "lm", 
                   method.args = list(formula = formula),
                   geom = "smooth")
```

The package also defines a `"null"` geom, which is used as default by the _debug_
_stats_ described above. Currently this geom is similar to the recently added 
`ggplot2::geom_blank()`.

```{r}
ggplot(my.data, aes(x, y, colour = group)) + geom_null()
```
