---
title: "`ggpmisc` `r packageVersion('ggpmisc')`<br>Additional examples"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Additional Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path = 'figure/pos-', fig.align = 'center', fig.show = 'hold',
               fig.width = 7, fig.height = 4)
options(warnPartialMatchArgs = FALSE)
```

## Preliminaries

```{r}
library(ggplot2)
library(ggpmisc)
library(xts)
```

## `try_data_frame()`

### Time series

Several different formats for storing time series data are used in R. Here we use in the examples objects of class `ts` but several other classes are supported as `try.xts()` is used internally.

```{r}
class(austres)
austres.df <- try_data_frame(austres)
class(austres.df)
lapply(austres.df, "class")
head(austres.df, 4)
```

```{r}
austres.df <- try_data_frame(austres, "month")
lapply(austres.df, "class")
head(austres.df, 4)
```

```{r}
austres.df <- try_data_frame(austres, as.numeric = TRUE)
lapply(austres.df, "class")
head(austres.df, 4)
```

```{r}
class(lynx)
lynx.df <- try_data_frame(lynx)
class(lynx.df)
lapply(lynx.df, "class")
head(lynx.df, 3)
```

```{r}
lynx_n.df <- try_data_frame(lynx, "year", as.numeric = TRUE)
lapply(lynx_n.df, "class")
head(lynx_n.df, 3)
```

```{r}
lynx <- try.xts(lynx)
class(lynx)
lynx.df <- try_data_frame(lynx)
lapply(lynx.df, "class")
head(lynx.df, 3)
```

### Other classes

`try_data_frame()` attempts to handle gracefully objects that are not time series.

```{r}
try_data_frame(1:5)
```

```{r}
try_data_frame(letters[1:5])
```

```{r}
try_data_frame(factor(letters[1:5]))
```

```{r}
try_data_frame(list(x = rep(1,5), y = 1:5))
```

```{r}
try_data_frame(data.frame(x = rep(1,5), y = 1:5))
```

```{r}
try_data_frame(matrix(1:10, ncol = 2))
```

## `stat_peaks()` and `stat_valleys()`

```{r}
ggplot(lynx.df, aes(time, V.lynx)) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", vjust = -0.5) +
  ylim(-100, 7300)
```

```{r}
ggplot(lynx_n.df, aes(time, V.lynx)) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", vjust = -0.5) +
  ylim(-100, 7300)
```

```{r}
ggplot(lynx.df, aes(time, V.lynx)) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", vjust = -0.5, x.label.fmt = "%Y") +
  stat_valleys(colour = "blue") +
  stat_valleys(geom = "text", colour = "blue", vjust = 1.5, x.label.fmt = "%Y") +
  ylim(-100, 7300)
```

```{r}
ggplot(lynx_n.df, aes(time, V.lynx)) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", vjust = -0.5, x.label.fmt = "%4.0f") +
  stat_valleys(colour = "blue") +
  stat_valleys(geom = "text", colour = "blue", vjust = 1.5, x.label.fmt = "%4.0f") +
  ylim(-100, 7300)
```

```{r}
ggplot(lynx.df, aes(time, V.lynx)) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", vjust = -0.5, x.label.fmt = "%Y") +
  stat_valleys(colour = "blue") +
  stat_valleys(geom = "text", colour = "blue", vjust = 1.5, x.label.fmt = "%Y") +
  ylim(-100, 7300)
```

```{r}
ggplot(lynx.df, aes(time, V.lynx)) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "rug", colour = "red") +
  stat_peaks(geom = "text", colour = "red", angle = 66,
             hjust = -0.1, x.label.fmt = "%Y") +
  ylim(NA, 7300)
```

```{r}
ggplot(lynx.df, aes(time, V.lynx)) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "rug", colour = "red") +
  stat_valleys(colour = "blue") +
  stat_valleys(geom = "rug", colour = "blue")
```

## `stat_poly_eq()`

```{r}
set.seed(4321)
# generate artificial data
x <- 1:100
y <- (x + x^2 + x^3) + rnorm(length(x), mean = 0, sd = mean(x^3) / 4)
my.data <- data.frame(x, y, group = c("A", "B"), y2 = y * c(0.5,2))

```

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(formula = formula, parse = TRUE)
```

```{r}
formula <- y ~ x + I(x^2) + I(x^3) - 1
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(formula = formula, parse = TRUE)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(formula = formula, parse = TRUE)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..adj.rr.label..), formula = formula, parse = TRUE)
```

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), formula = formula, 
               parse = TRUE)
```

```{r}
formula <- y ~ x + I(x^2) + I(x^3) - 1
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), formula = formula, parse = TRUE)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), formula = formula, parse = TRUE)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label =  paste(..eq.label.., ..adj.rr.label.., sep = "~~~~")), 
               formula = formula, parse = TRUE)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), size = rel(2.8),
               formula = formula, parse = TRUE) +
  facet_wrap(~group)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), size = rel(2.8),
               formula = formula, parse = TRUE) +
  facet_wrap(~group, scales = "free_y")
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, colour = group)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), vjust = c(-8, 0),
               formula = formula, parse = TRUE) +
  theme_bw()
```

```{r}
formula <- y ~ poly(x, 1, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), formula = formula, parse = TRUE)
```

```{r}
formula <- y ~ poly(x, 2, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), formula = formula, parse = TRUE)
```

```{r}
formula <- y ~ poly(x, 5, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), formula = formula, parse = TRUE)
```

## `stat_debug()`

```{r}
ggplot(my.data, aes(x, y)) + geom_point() + stat_debug_group()
```

```{r}
ggplot(my.data, aes(x, y)) + geom_point() + stat_debug_panel()
```

```{r}
ggplot(my.data, aes(x, y, colour = group)) + geom_point() + 
  stat_debug_group(vjust = c(-0.5,1.5))
```

```{r}
ggplot(my.data, aes(x, y, colour = group)) + geom_point() + 
  stat_debug_panel()
```

```{r}
ggplot(my.data, aes(x, y, shape = group)) + geom_point() + 
  stat_debug_group(vjust = c(-0.5,1.5))
```

```{r}
ggplot(my.data, aes(x, y)) + geom_point() + 
  facet_wrap(~group) + stat_debug_group()
```

```{r}
ggplot(my.data, aes(x, y)) + geom_point() + 
  facet_wrap(~group) + stat_debug_panel()
```

```{r}
ggplot(my.data, aes(x, y, colour = group)) + geom_point() + 
  facet_wrap(~group) + stat_debug_group()
```

```{r}
ggplot(my.data, aes(x, y, colour = group, shape = group)) + geom_point() + 
  facet_wrap(~group) + stat_debug_group()
```

## Datetime variables and ggplot statistics

Data frame `lynx_n.df` contains `time` as numeric,

```{r}
lynx_n.df <- try_data_frame(lynx, "year", as.numeric = TRUE)
lapply(lynx_n.df, "class")
head(lynx_n.df, 3)
```

and the compute function in the stat receives `x` as numeric.

```{r}
ggplot(lynx_n.df, aes(time, V.lynx)) + geom_line() +
  stat_debug_panel()
```

Data frame `lynx_t.df` contains `time` as `POSIXct`,

```{r}
lynx_t.df <- try_data_frame(lynx, "year", as.numeric = FALSE)
lapply(lynx_t.df, "class")
head(lynx_t.df, 3)
```

and the compute function in the stat still receives `x` as numeric. When writing
new stats in this case one has to use the `scales` argument received by the
`compute_group` or `compute_panel` functios to work-out when needed the scale
in use.

```{r}
ggplot(lynx_t.df, aes(time, V.lynx)) + geom_line() +
  scale_x_datetime() +
  stat_debug_panel()
```
