---
title: "`ggpmisc` `r packageVersion('ggpmisc')`<br>Density filtering"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Density filtering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path = 'figure/dens2d-pos-', fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4)
options(warnPartialMatchArgs = FALSE)
```

## Introduction

This vignette demonstrates the performance of the default arguments with 
artificial data sets of different sizes and drawn from different theoretical 
distributions. It used initially for testing but it also shows what to expect
under different situations.

## Preliminaries

```{r}
library(ggpmisc)
library(ggplot2)
library(tibble)
```

We define functions to simplify the generation of random data sets.

```{r}
make_data_tbl <- function(nrow = 100, rfun = rnorm, ...) {
  if (nrow %% 2) {
    nrow <- nrow + 1
  }
  
  set.seed(1001)
  
  tibble::tibble(
    x = rfun(nrow, ...),
    y = rfun(nrow, ...),
    group = rep(c("A", "B"), c(nrow / 2, nrow / 2))
  )
}
```

## Tests with different data sets

### Number of observations

By default the fraction of observations kept is 1/10.

```{r}
ggplot(data = make_data_tbl(6), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red")
```

```{r}
ggplot(data = make_data_tbl(6), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red", keep.fraction = 1/2)
```

```{r}
ggplot(data = make_data_tbl(20), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red")
```

```{r}
ggplot(data = make_data_tbl(100), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red")
```

```{r}
ggplot(data = make_data_tbl(500), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red")
```

```{r}
ggplot(data = make_data_tbl(2000), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red")
```

```{r}
ggplot(data = make_data_tbl(2000), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red", keep.fraction = 0.01)
```

```{r}
ggplot(data = make_data_tbl(2000), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red", 
                     keep.sparse = FALSE)
```

```{r}
ggplot(data = make_data_tbl(2000), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red", 
                     keep.sparse = FALSE)+
  stat_dens2d_filter(color = "blue")
```

```{r}
ggplot(data = make_data_tbl(2000), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red", 
                     keep.fraction = 0.01,
                     keep.sparse = FALSE)
```

```{r, eval=FALSE}
ggplot(data = make_data_tbl(10000), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red")
```

```{r, eval=FALSE}
ggplot(data = make_data_tbl(10000), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red", keep.fraction = 0.01)
```

### Random draws from different theoretical distributions

```{r}
ggplot(data = make_data_tbl(1000, rfun = runif), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red", keep.fraction = 0.1)
```

```{r}
ggplot(data = make_data_tbl(1000, rfun = rgamma, shape = 2), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red", keep.fraction = 0.1)
```

```{r}
ggplot(data = make_data_tbl(1000, rfun = rgamma, shape = 6), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red", keep.fraction = 0.1)
```

```{r}
ggplot(data = make_data_tbl(1000, rfun = rbeta, shape1 = 3, shape2 = 12), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red", keep.fraction = 0.1)
```

### Transformed scales

```{r}
ggplot(data = make_data_tbl(1000, rfun = rbeta, shape1 = 3, shape2 = 12), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red", keep.fraction = 0.1) +
  scale_y_log10()
```
