---
title: "Model-Based Plot Annotations"
subtitle: "'ggpmisc' `r packageVersion('ggpmisc')`"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Model-Based Plot Annotations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4)
options(warnPartialMatchArgs = FALSE,
        tibble.print.max = 4,
        tibble.print.min = 4,
        dplyr.summarise.inform = FALSE)
eval_flag <- TRUE # evaluate all code chunks
```

## Preliminaries

We load all the packages used in the examples.

```{r, message=FALSE}
library(ggpmisc)
library(xts)
library(lubridate)
library(tibble)
library(dplyr)
library(quantreg)

eval_nlme <-  requireNamespace("nlme", quietly = TRUE)
if (eval_nlme) library(nlme)
eval_broom <-  requireNamespace("broom", quietly = TRUE)
if (eval_broom) library(broom)
eval_broom_mixed <-  requireNamespace("broom.mixed", quietly = TRUE)
if (eval_broom_mixed) library(broom.mixed)
```

As we will use text and labels on the plotting area we change the default theme
to an uncluttered one.

```{r}
old_theme <- theme_set(theme_bw())
```

# Introduction to model-based annotations

The most common annotations on plots based on model fits are model equations,
tests of significance and various indicators of goodness of fit. What annotations are most useful depend on whether both _x_ and _y_ are mapped to
continuous variables, or _y_ to a continuous variable, and _x_ to a factor.
In some cases it may be desirable to add an ANOVA table as an inset, or a 
summary table.

The annotations for pairwise comparisons are currently not supported by 'ggpmisc' as package 'ggsignif' supports them.

In the case of transcriptomics and metabolomics data it is frequent to use 
ternary outcomes from tests of significance. As points can be very numerous it
it is also common to annotate plots with the number of observations in each quadrant.

These are all fairly specialized uses, that do not extend the grammar of 
graphics per se, but help by streamlining/automating its use for specific 
cases.

## Statistics

Statistics that help with reporting the results of model fits are
`stat_poly_eq()`, `stat_fit_residuals()`, `stat_fit_deviations()`,
`stat_fit_glance()`, `stat_fit_augment()`, `stat_fit_tidy()` and
`stat_fit_tb()`.

## Aesthetics and scales

Scales `scale_x_logFC()` and `scale_y_logFC()` are suitable for plotting of log
fold change data. Scales `scale_x_Pvalue()`, `scale_y_Pvalue()`, `scale_x_FDR()`
and `scale_y_FDR()` are suitable for plotting _p_-values and adjusted _p_-values
or false discovery rate (FDR). Default arguments are suitable for volcano and
quadrant plots as used for transcriptomics, metabolomics and similar data. 

Scales `scale_colour_outcome()`, `scale_fill_outcome()` and 
`scale_shape_outcome()` and functions `outome2factor()`, `threshold2factor()`,
`xy_outcomes2factor()` and `xy_thresholds2factor()` used together make it 
easy to map ternary numeric outputs and logical binary outcomes to color,
fill and shape aesthetics. Default arguments are suitable for volcano,
quadrant and other plots as used for genomics, metabolomics and similar data.

# Examples

## Statistics

### stat_poly_eq() and stat_poly_line()

A frequently used annotation in plots showing fitted lines is the display of the
parameters estimates from the model fit as an equation. `stat_poly_eq()`
automates this for models of _y_ on _x_ or _x_ on _y_ fitted with function
`lm()`, and `MASS:rlm()` or a user-supplied model fitting function and
arguments. By default this _statistic_ behaves similarly as
`ggplot2::stat_smooth()` with `method = "lm"` for _y_ on _x_ fits but adds
support for model formulas of _x_ on _y_ (in addition to the use of `orientation
= "y"`). Its default behaviour is to generate labels as character strings using
R's _expression_ syntax. This stat can also output equations using different
mark-up languages, including $\LaTeX$ and Markdown, or just numeric values
selected by the argument passed to parameter `output.type`. Nevertheless, all
examples below use expression syntax.

The statistic `stat_poly_line()` adds a layer identical to that from
`ggplot2::stat_smooth()` but has the same user interface and default arguments
as `stat_poly_eq()`.

Other _statistics_ described in later sections of this vignette are even more
flexible allowing various annotations for model fit functions supported by
package 'broom' or 'broom.mixed'.

We first generate a set of artificial data suitable for the plotting examples.

```{r}
set.seed(4321)
# generate artificial data
x <- 1:100
y <- (x + x^2 + x^3) + rnorm(length(x), mean = 0, sd = mean(x^3) / 4)
my.data <- data.frame(x, 
                      y, 
                      group = c("A", "B"), 
                      y2 = y * c(0.5,2),
                      block = c("a", "a", "b", "b"),
                      wt = sqrt(x))
```

First one example using defaults. The best practice, ensuring that the formula
used in both _statistics_ is the same is assign the formula to a variable, as
shown here. This is because the same model is fit twice to the same `data`, once
in `stat_smooth` and once in `stat_poly_eq`.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(formula = formula)
```

A ready formatted equation is also returned as a string that needs to be parsed
into an _expression_ for display.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = stat(eq.label)), formula = formula)
```

`stat_poly_eq()` makes available several character strings in the returned data
frame in separate columns: `eq.label`, `rr.label`, `adj.rr.label`, `AIC.label`,
`BIC.label`, `f.value.label`, `p.value.label` and `n`. One of these, `rr.label`,
is used by default, but `aes()` can be used to map a different one to the
`label` aesthetic. Here we show the adjusted coefficient of determination and
AIC. We call `stat_poly_eq()` twice to be able to separately control the
position and size of each label.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = stat(adj.rr.label)), formula = formula) +
  stat_poly_eq(aes(label = stat(AIC.label)),
               label.x = "right", label.y = "bottom", size = 3,     
               formula = formula)

```

Within `aes()` it is also possible to _compute_ new labels based on those
returned plus "arbitrary" text. The labels returned by default are meant to be
_parsed_ into expressions, so any text added should be valid for a string that
will be parsed. Inserting a comma plus white space in an expression requires
some trickery in the argument passed to `sep`. Do note the need to _escape_ the
embedded quotation marks as `\"`. Combining the labels in this way ensures
correct alignment. To insert only white space `sep = "~~~~~"` can be used, with
each tilde character, `~`, adding a rather small amount of white space. We show
only here, not to clutter the remaining examples, how to change the axis labels
to ensure consistency with the typesetting of the equation.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label =  paste(stat(eq.label), stat(adj.rr.label), 
                                  sep = "*\", \"*")),
               formula = formula) +
  labs(x = expression(italic(x)), y = expression(italic(y)))
```

Above we combined two character-string labels, but it is possible to add
additional ones and even character strings constants. In this example we use
several labels instead of just two, and separate them with various character
strings. We also change the size of the text in the label.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label =  paste(stat(eq.label), "*\" with \"*", 
                                  stat(rr.label), "*\", \"*", 
                                  stat(f.value.label), "*\", and \"*",
                                  stat(p.value.label), "*\".\"",
                                  sep = "")),
               formula = formula, size = 3)
```

It is also possible to format the text using `plain()`, `italic()`, `bold()` or
`bolditalic()` as described in `plotmath`.

```{r, eval=eval_flag}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label =  paste(stat(eq.label), stat(adj.rr.label), 
                                  sep = "~~italic(\"with\")~~")),
               formula = formula)
```

As these are expressions, to include two lines of text, we need either to add
`stat_poly_eq()` twice, passing an argument to `label.y` to reposition one of
the labels (as shown above) or use (as shown here) `atop()` within the
expression to create a single label with two lines of text.

```{r, eval=eval_flag}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = paste("atop(", stat(AIC.label), ",", stat(BIC.label), ")", sep = "")), 
               formula = formula)
```

Next, one example of how to remove the left hand side (_lhs_).

```{r, eval=eval_flag}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = stat(eq.label)),
               eq.with.lhs = FALSE,
               formula = formula)
```

Replacing the _lhs_.

```{r, eval=eval_flag}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = stat(eq.label)),
               eq.with.lhs = "italic(hat(y))~`=`~",
               formula = formula)
```

And a final example replacing both the  _lhs_ and the variable symbol used on
the _rhs_.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = stat(eq.label)),
               eq.with.lhs = "italic(h)~`=`~",
               eq.x.rhs = "~italic(z)",
               formula = formula) +
  labs(x = expression(italic(z)), y = expression(italic(h)))
```

As any valid R expression can be used, Greek letters are also supported, as well
as the inclusion in the label of variable transformations used in the model
formula or applied to the data. In addition, in the next example we insert white
space in between the parameter estimates and the variable symbol in the
equation.

```{r, eval=eval_flag}
formula <- y ~ poly(x, 2, raw = TRUE)
ggplot(my.data, aes(x, log10(y + 1e6))) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = stat(eq.label)),
               eq.with.lhs = "plain(log)[10](italic(delta)+10^6)~`=`~",
               eq.x.rhs = "~Omega",
               formula = formula) +
  labs(y = expression(plain(log)[10](italic(delta)+10^6)), x = expression(Omega))
```

A couple of additional examples of polynomials of different orders, and
specified in different ways.

Higher order polynomial. When using function `poly()` it is necessary to
set `raw = TRUE`.

```{r}
formula <- y ~ poly(x, 5, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = stat(eq.label)), formula = formula)
```

Intercept forced to zero.

```{r, eval=eval_flag}
formula <- y ~ x + I(x^2) + I(x^3) - 1
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = stat(eq.label)), formula = formula)
```
Even when labels are returned, numeric values are also returned for `r.square`,
`p.value` and `n`. This can be used in `aes()` for example to show the equation
only if a certain condition like a minimum value for `r.square` has been reached.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label =  ifelse(stat(adj.r.squared > 0.3),
                                   paste(stat(eq.label), stat(adj.rr.label), 
                                         sep = "*\", \"*"),
                                   stat(adj.rr.label))),
               formula = formula) +
  labs(x = expression(italic(x)), y = expression(italic(y)))
```

We give below several examples to demonstrate how other components of the
`ggplot` object affect the behaviour of this statistic.

Facets work as expected either with fixed or free scales. Although below we had
to adjust the size of the font used for the equation.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = stat(eq.label)), size = 3,
               formula = formula) +
  facet_wrap(~group)
```

```{r, eval=eval_flag}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = stat(eq.label)), size = 3,
               formula = formula) +
  facet_wrap(~group, scales = "free_y")
```

Grouping, in this example using the colour aesthetic also works as expected. We
can use justification and supply an absolute location for the equation, but the
default frequently works well as in the example below.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, colour = group)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = stat(eq.label)), formula = formula, vstep = 0.06)
```

To add a label to the equation for each group, we map it to a pseudo-aesthetic
named `grp.label`. (Equation labelling uses a "back door" in 'ggplot2' that may
be closed in future versions, meaning that this approach to labelling may stop
working in the future.)

```{r, eval=eval_flag}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, colour = group, grp.label = group)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = stat(paste("bold(", grp.label, "*\":\")~~", 
                                      eq.label, sep = ""))),
               formula = formula)
```

Being able to label equations allows us to dispense with the use of color,
which in many cases is needed for printed output.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, linetype = group, grp.label = group)) +
  geom_point() +
  stat_poly_line(formula = formula, color = "black") +
  stat_poly_eq(aes(label = stat(paste("bold(", grp.label, "*':')~~~", 
                                      eq.label, sep = ""))),
               formula = formula)
```

User supplied label positions relative to the ranges of the _x_ and _y_ scales
are also supported, both through string constants and numeric values in the
range 0 to 1, when using the default `geom_text_npc()`.

```{r, eval=eval_flag}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, colour = group)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = stat(eq.label)),
               formula = formula,
               label.x = "centre")
```

The default locations are now based on normalized coordinates, and consequently
these defaults work even when the range of the _x_ and _y_ scales varies from
panel to panel.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, fill = block)) +
  geom_point(shape = 21, size = 3) +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = stat(rr.label)), size = 3,
               geom = "label_npc", alpha = 0.33,
               formula = formula) +
  facet_wrap(~group, scales = "free_y")
```

If grouping is not the same within each panel created by faceting, the automatic
location of labels results in "holes" for the factor levels not present in a
given panel. In this case, consistent positioning requires passing explicitly
the positions for each individual group. In the plot below the simultaneous
mapping to both color and fill aesthetics creates four groups, two with data in
panel A and the other two in panel B.

```{r, eval=eval_flag}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, colour = group, fill = block)) +
  geom_point(shape = 21, size = 3) +
  stat_poly_line(formula = formula) +
  stat_poly_eq(aes(label = stat(rr.label)), size = 3, alpha = 0.2,
               geom = "label_npc", label.y = c(0.95, 0.85, 0.95, 0.85),
               formula = formula) +
  facet_wrap(~group, scales = "free_y")
```

It is possible to use `geom_text()` and `geom_label()` but in this case, if
label coordinates are given explicitly they should be expressed in native data
coordinates. When multiple labels need to be positioned a vector of coordinates
can be used as shown here for  `label.x` and `label.y`.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, colour = group)) +
  geom_point() +
  stat_poly_line(formula = formula) +
  stat_poly_eq(geom = "text", aes(label = stat(eq.label)),
               label.x = c(100, 90), label.y = c(-1e4, 2.1e6), hjust = "inward",
               formula = formula)
```

**Note:** Automatic positioning using `geom_text()` and `geom_label()` is not
supported when faceting uses free scales. In this case `geom_text_npc()` and/or 
`geom_label_npc()` must be used.

Occasionally, possibly as a demonstration in teaching, we may want to
compare a fit of _x_ on _y_ and one of _y_ on _x_ in the same plot. As is the
case for model fitting functions themselves this is simply indicated by the
model passed as argument to `formula`, or consistently with `stat_smooth()`
through parameter `orientation`.

```{r}
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_poly_line(formula = y ~ x, color = "blue") +
  stat_poly_eq(aes(label = stat(eq.label)), color = "blue") +
  stat_poly_line(formula = y ~ x, color = "red", orientation = "y") +
  stat_poly_eq(aes(label = stat(eq.label)), color = "red", orientation = "y",
               label.y = 0.9)
```

### stat_quant_eq(), stat_quant_line() and stat_quant_band()

Package 'quantreg' implements quantile regression for multiple types of models:
linear models with function `rq()`, non-linear models with function `nlreg()`, 
and additive models with function `rqss()`. Each of these functions supports
different methods of estimation. As with the statistics for polynomials 
described in the previous section, the statistics described in the current
section _support only linear models_.

In the case of quantile regression it is frequently the case that multiple
quantile regressions are fitted. This case is not supported by `stat_poly_eq()`
and we need to use `stat_quant_eq()`. This statistics is very similar to
`stat_poly_eq()` both in behaviour and parameters. It supports quantile
regression of _y_ on _x_ or _x_ on _y_ fitted with function `quantreg::rq`.

The statistic `stat_quant_line()` adds a layer similar to that created by
`ggplot2::stat_smooth()` including confidence bands for quantile regression, by
default for _p_ = 0.95. However, it has the same user interface and default
arguments as `stat_quant_eq()`, and supports both _x_ and _y_ as explanatory
variable in `formula` and also `orientation`. Confidence intervals for the fits
to individual quantiles fulfil the same purpose as in OLS fits: providing a
boundary for the _estimated line_ computed for each individual quantile. They
are not shown by default, unless a single quantile is passed, but can be enabled
with `se = TRUE` and disabled with `se = FALSE`.

The statistic `stat_quant_band()` creates a line plus band plot that is, by
default equivalent to the box of a box plot showing regressions as a band for
quartiles plus a line for the median media. In this case the 
band informs about the spread of the _observations_. Which three quantiles
are fitted can be set through parameter `quantiles` but in contrast to
`stat_quant_line()` and `stat_quant_eq()` only a vector of three quantiles is
accepted.

Similarly to `ggplot2::stat_quantiles()`, `stat_quant_eq()` and
`stat_quant_line()` accept a `numeric` vector as argument to `quantiles` and
output one equation or one line per group and quantile. Stat `stat_quant_eq()`
also returns column `grp.label` containing by default a character string
indicating the value of the quantile.

Here we use the default value for quantiles, `c(0.25, 0.50, 0.75)`, equivalent
to those used for the box of box plots, and we obtain a plot with a line for
median regression and a band highlighting the space between the quartiles.

```{r, warning=FALSE}
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_quant_band(formula = y ~ poly(x, 2))
```

Overriding the values for `quantiles` we find a possible boundary containing 90% of the observations.

```{r, warning=FALSE}
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_quant_line(formula = y ~ poly(x, 2), quantiles = c(0.1, 0.9))
```

A line for the median regression plus a confidence band for it is the default
for a single quantile. (This is similar to the behaviour of `ggplot2::stat_smooth()`.)

```{r, warning=FALSE}
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_quant_line(formula = y ~ poly(x, 2), quantiles = 0.5)
```

With `stat_quant_eq()` we add the fitted equations to the first example in this
section (we also override the use of color).

```{r, warning=FALSE}
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_quant_band(formula = formula, color = "black", fill = "grey60") +
  stat_quant_eq(aes(label = paste(stat(grp.label), "*\": \"*",
                                  stat(eq.label), sep = "")),
                formula = formula) +
  theme_classic()
```

Grouping and faceting are supported by `stat_quant_eq()`, `stat_quant_line()`
and `stat_quant_band()`. Below are some examples of grouping.

```{r, warning=FALSE}
ggplot(my.data, aes(x, y, color = group)) +
  geom_point() +
  stat_quant_line(formula = formula) +
  stat_quant_eq(aes(label = paste(stat(grp.label), "*\": \"*",
                                  stat(eq.label), sep = "")),
               formula = formula)
```

This can be combined with a group label by mapping in `aes()` a character 
variable to `grp.label`.

```{r, warning=FALSE}
ggplot(my.data, aes(x, y, group = group, linetype = group, 
                    shape = group, grp.label = group)) +
  geom_point() +
  stat_quantile(formula = formula, color = "black") +
  stat_quant_eq(aes(label = paste(stat(grp.label), "*\": \"*",
                                  stat(eq.label), sep = "")),
                formula = formula, quantiles = c(0.05, 0.95)) +
  theme_classic()
```

In some cases _double quantile regression_ is an informative method to assess
reciprocal constraints. For this a fit of _x_ on _y_ and one of _y_ on _x_ in
the same plot are computed for a large quantile. As is the case for model
fitting functions themselves this is simply indicated by the model passed as
argument to `formula`, or consistently with `ggplot2::stat_smooth()` through
parameter `orientation`.

```{r, warning=FALSE}
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_quant_line(formula = y ~ x, color = "blue", quantiles = 0.05) +
  stat_quant_eq(aes(label = stat(eq.label)), formula = y ~ x, color = "blue",
                quantiles = 0.05) +
  stat_quant_line(formula = x ~ y, color = "red", quantiles = 0.95) +
  stat_quant_eq(aes(label = stat(eq.label)), formula = x ~ y, color = "red", 
                quantiles = 0.95, label.y = 0.9)
```

Please, see the section on `stat_poly_eq()` for additional examples, as here
we have highlighted mainly the differences between these two statistics. 

### stat_fit_residuals

Sometimes we need to quickly plot residuals matching fits plotted with
`geom_smooth()` using grouping and facets, so a new (simple) statistic was born.
In teaching it is specially important to avoid distractions and plots residuals
from different types of models consistently.

`stat_fit_residuals()` can be used to plot the residuals form models of _y_ on
_x_ or _x_ on _y_ fitted with function `lm`, `MASS:rlm`, `quantreg::rq` (only
median regression) or a user-supplied model fitting function and arguments. In 
the returned values the response variable is replaced by the residuals from
the fit. The default geom is `"point"`.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y, colour = group)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  stat_fit_residuals(formula = formula)
```

### stat_fit_deviations

When teaching we may need to highlight residuals in plots used in slides and notes.
Statistic `stat_fit_deviations` makes it easy to
achieve this for the residuals from models of _y_ on
_x_ or _x_ on _y_ fitted with function `lm`, `MASS:rlm`, `quantreg::rq` (only
median regression) or a user-supplied model fitting function and arguments. The
default geom is `"segment"` and the values of the the observations and the
fitted values are mapped to aesthetics so that each deviation is displayed as
a segment.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_deviations(formula = formula, colour = "red") +
  geom_point()
```

As the geometry used by default is `geom_segment()`, additional aesthetics
can be mapped or set to constant values. Here we add arrowheads.

```{r, eval=eval_flag}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_smooth(method = "lm", formula = formula) +
  geom_point() +
  stat_fit_deviations(formula = formula, colour = "red",
                      arrow = arrow(length = unit(0.015, "npc"), 
                                   ends = "both"))
```

When weights are available, either suplied by the user, or computed as part of
the fit, they are returned in `data`. Having weights available allows encoding
them using colour. We here use a robust regression fit with `MASS::rlm`.

```{r}
my.data.outlier <- my.data
my.data.outlier[6, "y"] <- my.data.outlier[6, "y"] * 5
ggplot(my.data.outlier, aes(x, y)) +
  stat_smooth(method = MASS::rlm, formula = formula) +
  stat_fit_deviations(formula = formula, method = "rlm",
                      mapping = aes(colour = after_stat(weights)),
                      show.legend = TRUE) +
  scale_color_gradient(low = "red", high = "blue", limits = c(0, 1)) +
  geom_point()
```

### stat_fit_glance

Package 'broom' provides consistently formatted output from different model
fitting functions. These made it possible to write a model-annotation statistic
that is very flexible but that requires additional input from the user to
generate the character strings to be mapped to the `label` aesthetic.

As we have above given some simple examples, we here exemplify this statistic
in a plot with grouping, and assemble a label for the _P_-value using a string
parsed into a expression. We also change the default position of the labels. 

```{r}
# formula <- y ~ poly(x, 3, raw = TRUE)
# broom::augment does not handle poly() correctly!
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y, colour = group)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_glance(method = "lm", 
                  method.args = list(formula = formula),
                  label.x = "right",
                  label.y = "bottom",
                  aes(label = paste("italic(P)*\"-value = \"*", 
                                    signif(stat(p.value), digits = 4),
                                    sep = "")),
                  parse = TRUE)
```

It is also possible to fit a non-linear model with `method = "nls"`, and any
other model for which a `glance()` method exists. Do consult the documentation 
for package 'broom'. Here we fit the Michaelis-Menten equation to reaction
rate versus concentration data.

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  geom_smooth(method = "nls", 
              formula = micmen.formula,
              se = FALSE) +
  stat_fit_glance(method = "nls", 
                  method.args = list(formula = micmen.formula),
                  aes(label = paste("AIC = ", signif(stat(AIC), digits = 3), 
                                    ", BIC = ", signif(stat(BIC), digits = 3),
                                    sep = "")),
                  label.x = "centre", label.y = "bottom")
```

### stat_fit_tidy

This stat makes it possible to add the equation for any fitted model for which
`broom::tidy()` is implemented. Alternatively, individual values such as
estimates for the fitted parameters, standard errors, or _P_-values can be added
to a plot. However, the user has to explicitly construct the labels within
`aes()`. This statistic respects grouping based on aesthetics, and reshapes the
output of `tidy()` so that the values for a given group are in a single row in
the returned `data`.

As first example we fit a non-linear model, the Michaelis-Menten equation of
reaction kinetics, using `nls()`. We use the self-starting function `SSmicmen()`
available in R.

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  geom_smooth(method = "nls", 
              formula = micmen.formula,
              se = FALSE) +
  stat_fit_tidy(method = "nls", 
                method.args = list(formula = micmen.formula),
                label.x = "right",
                label.y = "bottom",
                aes(label = paste("V[m]~`=`~", signif(stat(Vm_estimate), digits = 3),
                                  "%+-%", signif(stat(Vm_se), digits = 2),
                                  "~~~~K~`=`~", signif(stat(K_estimate), digits = 3),
                                  "%+-%", signif(stat(K_se), digits = 2),
                                  sep = "")),
                parse = TRUE)
```

Using paste we can build a string that can be parsed into an R expression, in
this case for a non-linear equation.

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  geom_smooth(method = "nls", 
              formula = micmen.formula,
              se = FALSE) +
  stat_fit_tidy(method = "nls", 
                method.args = list(formula = micmen.formula),
                size = 3,
                label.x = "center",
                label.y = "bottom",
                vstep = 0.12,
                aes(label = paste("V~`=`~frac(", signif(stat(Vm_estimate), digits = 2), "~C,",
                                  signif(stat(K_estimate), digits = 2), "+C)",
                                  sep = "")),
                parse = TRUE) +
  labs(x = "C", y = "V")
```

What if we would need a more specific _statistic_, similar to `stat_poly_eq()`?
We can use `stat_fit_tidy()` as the basis for its definition.

```{r}
stat_micmen_eq <- function(vstep = 0.12,
                           size = 3,
                           ...) {
  stat_fit_tidy(method = "nls", 
                method.args = list(formula = micmen.formula),
                aes(label = paste("V~`=`~frac(", signif(stat(Vm_estimate), digits = 2), "~C,",
                                  signif(stat(K_estimate), digits = 2), "+C)",
                                  sep = "")),
                parse = TRUE,
                vstep = vstep,
                size = size,
                ...)
}
```

The code for the figure is now simpler, and still produces the same figure (not shown).

```{r, eval=eval_flag}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  geom_smooth(method = "nls", 
              formula = micmen.formula,
              se = FALSE) +
  stat_micmen_eq(label.x = "center",
                label.y = "bottom") +
  labs(x = "C", y = "V")
```

As a second example we show a quantile regression fit using function `rq()` 
from package 'quantreg'.

```{r}
my_formula <- y ~ x

ggplot(mpg, aes(displ, 1 / hwy)) +
  geom_point() +
  geom_quantile(quantiles = 0.5, formula = my_formula) +
  stat_fit_tidy(method = "rq",
                method.args = list(formula = y ~ x, tau = 0.5), 
                tidy.args = list(se.type = "nid"),
                mapping = aes(label = sprintf('y~"="~%.3g+%.3g~x*", with "*italic(P)~"="~%.3f',
                                              after_stat(Intercept_estimate), 
                                              after_stat(x_estimate),
                                              after_stat(x_p.value))),
                parse = TRUE)
```

We can define a `stat_rq_eq()` if we need to add similar equations to
several plots. In this example we retain the ability of the user to
override most of the default default arguments.

```{r}
stat_rq_eqn <- 
  function(formula = y ~ x, 
           tau = 0.5,
           method = "br",
           mapping = aes(label = sprintf('y~"="~%.3g+%.3g~x*", with "*italic(P)~"="~%.3f',
                                         after_stat(Intercept_estimate), 
                                         after_stat(x_estimate),
                                         after_stat(x_p.value))),
           parse = TRUE,
           ...) {
    method.args <- list(formula = formula, tau = tau, method = method)
    stat_fit_tidy(method = "rq",
                  method.args = method.args, 
                  tidy.args = list(se.type = "nid"),
                  mapping = mapping,
                  parse = parse,
                  ...)
  }
```

And the code of the figure now as simple as. Figure not shown, as is identical
to the one above.

```{r, eval=eval_flag}
ggplot(mpg, aes(displ, 1 / hwy)) +
  geom_point() +
  geom_quantile(quantiles = 0.5, formula = my_formula) +
  stat_rq_eqn(tau = 0.5, formula = my_formula)
```

### stat_fit_tb

This stat makes it possible to add summary or ANOVA tables for any fitted model
for which `broom::tidy()` is implemented. The output from `tidy()` is embedded
as a single list value within the returned `data`, an object of class `tibble`.
This statistic **ignores grouping** based on aesthetics. This allows fitting
models when `x` or `y` is a factor (as in such cases `ggplot` splits the data
into groups, one for each level of the factor, which is needed for example for
`stat_summary()` to work as expected). By default, the `"table"` geometry is
used. The use of `geom_table()` is described in a separate section of this User
Guide. Table themes and mapped aesthetics are supported.

The default output of `stat_fit_tb` is the default output from `tidy(mf)` where
`mf` is the fitted model.

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_tb(method = "lm",
              method.args = list(formula = formula),
              tb.vars = c(Parameter = "term", 
                          Estimate = "estimate", 
                          "s.e." = "std.error", 
                          "italic(t)" = "statistic", 
                          "italic(P)" = "p.value"),
              label.y = "top", label.x = "left",
              parse = TRUE)
```

When `tb.type = "fit.anova"` the output returned is that from `tidy(anova(mf))`
where `mf` is the fitted model. Here we also show how to replace names of
columns and terms, and exclude one column, in this case, the mean squares.

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_tb(method = "lm",
              method.args = list(formula = formula),
              tb.type = "fit.anova",
              tb.vars = c(Effect = "term", 
                          df = "df",
                          "italic(F)" = "statistic", 
                          "italic(P)" = "p.value"),
              tb.params = c(x = 1, "x^2" = 2, "x^3" = 3, Resid = 4),
              label.y = "top", label.x = "left",
              parse = TRUE)
```

When `tb.type = "fit.coefs"` the output returned is that of `tidy(mf)` after
selecting the `term` and `estimate` columns.

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_tb(method = "lm",
              method.args = list(formula = formula),
              tb.type = "fit.coefs", parse = TRUE,
              label.y = "center", label.x = "left")
```

Faceting works as expected, but grouping is ignored as mentioned above. In this
case, the color aesthetic is not applied to the text of the tables.
Furthermore, if `label.x.npc` or `label.y.npc` are passed numeric vectors of
length > 1, the corresponding values are obeyed by the different panels.

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K)
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  facet_wrap(~state) +
  geom_point() +
  geom_smooth(method = "nls",
              formula = micmen.formula,
              se = FALSE) +
  stat_fit_tb(method = "nls",
              method.args = list(formula = micmen.formula),
              tb.type = "fit.coefs",
              label.x = 0.9,
              label.y = c(0.75, 0.2)) +
  theme(legend.position = "none") +
  labs(x = "C", y = "V")
```

The data in the example below are split by `ggplot` into six groups based on the
levels of the `feed` factor. However, as `stat_fit_tb()` ignores groupings, we
can still fit a linear model to all the data in the panel.

```{r}
ggplot(chickwts, aes(factor(feed), weight)) +
  stat_summary(fun.data = "mean_se") +
  stat_fit_tb(tb.type = "fit.anova",
              label.x = "center",
              label.y = "bottom") +
  expand_limits(y = 0)
```

We can flip the system of coordinates, if desired.

```{r}
ggplot(chickwts, aes(factor(feed), weight)) +
  stat_summary(fun.data = "mean_se") +
  stat_fit_tb(tb.type = "fit.anova", label.x = "left", size = 3) +
  scale_x_discrete(expand = expansion(mult = c(0.2, 0.5))) +
  coord_flip()
```

It is also possible to rotate the table using `angle`. Here we also show how to
replace the column headers with strings to be parsed into R expressions.

```{r, eval=eval_flag}
ggplot(chickwts, aes(factor(feed), weight)) +
  stat_summary(fun.data = "mean_se") +
  stat_fit_tb(tb.type = "fit.anova",
              angle = 90, size = 3,
              label.x = "right", label.y = "center",
              hjust = 0.5, vjust = 0,
              tb.vars = c(Effect = "term", 
                          "df",
                          "M.S." = "meansq", 
                          "italic(F)" = "statistic", 
                          "italic(P)" = "p.value"),
              parse = TRUE) +
  scale_x_discrete(expand = expansion(mult = c(0.1, 0.35))) +
  expand_limits(y = 0)
```

### stat_fit_augment

**Experimental!** 
Use `ggplot2::stat_smooth` instead of `stat_fit_augment` if possible.

For a single panel and no grouping, there is little advantage in using this
statistic compared to the examples in the documentation of package 'broom'. With
grouping and faceting `stat_fit_augment` may occasionally be more convenient
than `ggplot2::stat_smooth` because of its additional flexibility.

```{r}
# formula <- y ~ poly(x, 3, raw = TRUE)
# broom::augment does not handle poly correctly!
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_fit_augment(method = "lm",
                   method.args = list(formula = formula))
```

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y, colour = group)) +
  geom_point() +
  stat_fit_augment(method = "lm", 
                   method.args = list(formula = formula))
```

We can override the variable returned as `y` to be any of the variables in
the data frame returned by `broom::augment` while still preserving the original
`y` values.

```{r, eval=eval_flag}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y)) +
  stat_fit_augment(method = "lm",
                   method.args = list(formula = formula),
                   geom = "point",
                   y.out = ".resid")
```

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y, colour = group)) +
  stat_fit_augment(method = "lm",
                   method.args = list(formula = formula),
                   geom = "point",
                   y.out = ".std.resid")
```

We can use any model fitting method for which `augment` is implemented.

```{r, eval=eval_flag}
args <- list(formula = y ~ k * e ^ x,
             start = list(k = 1, e = 2))
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  stat_fit_augment(method = "nls",
                   method.args = args)
```

```{r, eval=eval_flag}
args <- list(formula = y ~ k * e ^ x,
             start = list(k = 1, e = 2))
ggplot(mtcars, aes(wt, mpg)) +
  stat_fit_augment(method = "nls",
                   method.args = args,
                   geom = "point",
                   y.out = ".resid")
```

**Note:** The tidiers for mixed models have moved to package 'broom.mixed'!

```{r, eval=(eval_nlme && eval_broom_mixed)}
args <- list(model = y ~ SSlogis(x, Asym, xmid, scal),
             fixed = Asym + xmid + scal ~1,
             random = Asym ~1 | group,
             start = c(Asym = 200, xmid = 725, scal = 350))
ggplot(Orange, aes(age, circumference, colour = Tree)) +
  geom_point() +
  stat_fit_augment(method = "nlme",
                   method.args = args,
                   augment.args = list(data = quote(data)))
```


## Scales

Volcano and quadrant plots are scatter plots with some peculiarities. Creating
these plots from scratch using 'ggplot2' can be time consuming, but allows
flexibility in design. Rather than providing a 'canned' function to produce
volcano plots, package 'ggpmisc' provides several building blocks that
facilitate the production of volcano and quadrant plots. These are wrappers on
_scales_ from packages 'scales' and 'ggplot2' with suitable defaults plus helper
functions to create factors from numeric outcomes.

### scale_color_outcome, scale_fill_outcome

Manual scales for color and fill aesthetics with defaults suitable for the three
way outcomes from statistical tests.

### scale_x_FDR, scale_y_FDR, scale_x_logFC, scale_y_logFC, scale_x_Pvalue, scale_y_Pvalue

Scales for _x_ or _y_ aesthetics mapped to _P_-values, FDR (false discovery
rate) or log FC (logarithm of fold-change) as used in volcano and quadrant plots
of transcriptomics and metabolomics data.

### symmetric_limits

A simple function to expand scale limits to be symmetric around zero. Can be
passed as argument to parameter limits of continuous scales from packages
'ggpmisc', 'ggplot2' or 'scales' (and extensions).

### Volcano-plot examples

Volcano plots are frequently used for transcriptomics and metabolomics. They are used to show _P_-values or FDR (false discovery rate) as a function of effect size, which can be either an increase or a decrease. Effect sizes are expressed as fold-changes compared to a control or reference condition. Colors are frequently used to highlight significant responses. Counts of significant increases and decreases are frequently also added.

Outcomes encoded as -1, 0 or 1, as seen in the tibble below need to be converted into factors with suitable labels for levels. This can be easily achieved with function `outcome2factor()`.

```{r}
head(volcano_example.df) 
```

Most frequently fold-change data is available log-transformed, using either
2 or 10 as base. In general it is more informative to use tick labels expressed
as un-transformed fold-change. 

By default `scale_x_logFC()` and `scale_y_logFC()` expect the logFC data
log2-transformed, but this can be overridden. The default use of untransformed
fold-change for tick labels can also be overridden. Scale `scale_x_logFC()` in
addition by default expands the scale limits to make them symmetric around zero.
If `%unit` is included in the name character string, suitable units are appended
as shown in the example below.

Scales `scale_y_Pvalue()` and `scale_x_Pvalue()` have suitable defaults for name
and labels, while `scale_colour_outcome()` provides suitable defaults for the
colors mapped to the outcomes. To change the labels and title of the `key` or
`guide` pass suitable arguments to parameters `name` and `labels` of these
scales. These x and y scales by default _squish_ off-limits (out-of-bounds)
observations towards the limits.

```{r}
volcano_example.df %>%
  mutate(., outcome.fct = outcome2factor(outcome)) %>%
  ggplot(., aes(logFC, PValue, colour = outcome.fct)) +
  geom_point() +
  scale_x_logFC(name = "Transcript abundance%unit") +
  scale_y_Pvalue() +
  scale_colour_outcome() +
  stat_quadrant_counts(data = . %>% filter(outcome != 0))
```

By default `outcome2factor()` creates a factor with three levels as in the
example above, but this default can be overridden as shown below.

```{r}
volcano_example.df %>%
  mutate(., outcome.fct = outcome2factor(outcome, n.levels = 2)) %>%
  ggplot(., aes(logFC, PValue, colour = outcome.fct)) +
  geom_point() +
  scale_x_logFC(name = "Transcript abundance%unit") +
  scale_y_Pvalue() +
  scale_colour_outcome() +
  stat_quadrant_counts(data = . %>% filter(outcome != 0))
```

### Quadrant-plot examples

Quadrant plots are scatter plots with comparable variables on both axes and usually include the four quadrants. When used for transcriptomics and metabolomics, they are used to compare responses expressed as fold-change to two different conditions or treatments. They are useful when many responses are measured as in transcriptomics (many different genes) or metabolomics (many different metabolites). 

A single panel quadrant plot is as easy to produce as a volcano plot using scales `scale_x_logFC()` and `scale_y_logFC()`. The data includes two different outcomes and two different log fold-change variables. See the previous section on volcano plots for details. In this example we use as shape a filled circle and map one of the outcomes to color and the other to fill, using the two matched scales `scale_colour_outcome()` and `scale_fill_outcome()`.

```{r}
head(quadrant_example.df)
```

In this plot we do not include those genes whose change in transcript abundance is uncertain under both x and y conditions.

```{r}
quadrant_example.df %>%
  mutate(.,
         outcome.x.fct = outcome2factor(outcome.x),
         outcome.y.fct = outcome2factor(outcome.y),
         outcome.xy.fct = xy_outcomes2factor(outcome.x, outcome.y)) %>%
         filter(outcome.xy.fct != "none") %>%
  ggplot(., aes(logFC.x, logFC.y, colour = outcome.x.fct, fill = outcome.y.fct)) +
  geom_quadrant_lines(linetype = "dotted") +
  stat_quadrant_counts(size = 3, colour = "white") +
  geom_point(shape = "circle filled") +
  scale_x_logFC(name = "Transcript abundance for x%unit") +
  scale_y_logFC(name = "Transcript abundance for y%unit") +
  scale_colour_outcome() +
  scale_fill_outcome() +
  theme_dark()
```

To plot in separate panels those observations that are significant along both _x_ and _y_ axes, _x_ axis, _y_ axis, or none, with quadrants merged takes more effort. We first define two helper functions to add counts and quadrant lines to each of the four panels.

```{r}
all_quadrant_counts <- function(...) {
  list(  
    stat_quadrant_counts(data = . %>% filter(outcome.xy.fct == "xy"), ...),
    stat_quadrant_counts(data = . %>% filter(outcome.xy.fct == "x"), pool.along = "y", ...),
    stat_quadrant_counts(data = . %>% filter(outcome.xy.fct == "y"), pool.along = "x", ...),
    stat_quadrant_counts(data = . %>% filter(outcome.xy.fct == "none"), quadrants = 0L, ...)
  )
}
```


```{r}
all_quadrant_lines <- function(...) { 
  list(
    geom_hline(data =  data.frame(outcome.xy.fct = factor(c("xy", "x", "y", "none"),
                                                          levels = c("xy", "x", "y", "none")),
                                  yintercept = c(0, NA, 0, NA)),
               aes_(yintercept = ~yintercept),
               na.rm = TRUE,
               ...),
    geom_vline(data =  data.frame(outcome.xy.fct = factor(c("xy", "x", "y", "none"),
                                                          levels = c("xy", "x", "y", "none")),
                                  xintercept = c(0, 0, NA, NA)),
               aes_(xintercept = ~xintercept),
               na.rm = TRUE,
               ...)
  )
}
```

And use these functions to build the final plot, in this case including all genes.

```{r}
quadrant_example.df %>%
  mutate(.,
         outcome.x.fct = outcome2factor(outcome.x),
         outcome.y.fct = outcome2factor(outcome.y),
         outcome.xy.fct = xy_outcomes2factor(outcome.x, outcome.y)) %>%
  ggplot(., aes(logFC.x, logFC.y, colour = outcome.x.fct, fill = outcome.y.fct)) +
  geom_point(shape = 21) +
  all_quadrant_lines(linetype = "dotted") +
  all_quadrant_counts(size = 3, colour = "white") +
  scale_x_logFC(name = "Transcript abundance for x%unit") +
  scale_y_logFC(name = "Transcript abundance for y%unit") +
  scale_colour_outcome() +
  scale_fill_outcome() +
  facet_wrap(~outcome.xy.fct) +
  theme_dark()
```

