---
title: "User Guide"
subtitle: "'ggpmisc' `r packageVersion('ggpmisc')`"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4)
options(warnPartialMatchArgs = FALSE,
        tibble.print.max = 4,
        tibble.print.min = 4)
```

## Preliminaries

We load all the packages used in the examples.

```{r}
library(ggpmisc)
library(ggrepel)
library(xts)
library(lubridate)
library(tibble)
library(nlme)
```
As we will use text and labels on the plotting area we change the default theme
to an uncluttered one.

```{r}
old_theme <- theme_set(theme_bw())
```

## Introduction

Many of the functions, including ggplot statistics and geoms, included in
package 'ggpmisc' had their origin in my need to produce plots for use in
teaching. Some of them are more generally useful, such as `stat_poly_eq()`, but
others like `stat_fit_deviations()` are squarely aimed and producing learning
material. Finally, several statistics for debugging and learning how ggplot
statistics and geoms interact with each other will be of use only to developers
of new statistics and geoms. Function `try_tibble()` opens the door to easily
converting time series stored in objects of several different classes into data
frames. This, for example, useful for plotting time series data with `ggplot()`.
New `ggplot()` methods for classes `ts` and `xts` make the call to
`try_tibble()` and conversion automatic.

## ggplot methods for time series

`ggplot()` methods for classes `"ts"` and `"xts"` automate plotting of time
series data. For plotting time series data stored in objects of other classes,
see the conversion functions `try_tibble()` and `try_data_frame()` in the last
section of this vignette.

```{r}
ggplot(lynx) + geom_line()
```

```{r}
ggplot(lynx, as.numeric = FALSE) + geom_line()
```

```{r}
ggplot(AirPassengers) + geom_line()
```

```{r}
ggplot(AirPassengers, as.numeric = FALSE) + geom_line()
```

## stat_peaks and stat_valleys

Using POSIXct for `time` and the default formatting of labels.

```{r}
ggplot(lynx, as.numeric = FALSE) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", vjust = -0.5) +
  ylim(-100, 7300)
```

Using numeric values for `time` and the default formatting of labels.

```{r}
ggplot(lynx) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", vjust = -0.5) +
  ylim(-100, 7300)
```

Using POSIXct for `time` but supplying a format string. In addition marking both
peaks and valleys.

```{r}
ggplot(lynx, as.numeric = FALSE) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", vjust = -0.5, x.label.fmt = "%Y") +
  stat_valleys(colour = "blue") +
  stat_valleys(geom = "text", colour = "blue", vjust = 1.5, x.label.fmt = "%Y") +
  ylim(-100, 7300)
```

Using numeric for `time` but supplying a format string. In addition marking both peaks and valleys.

```{r}
ggplot(lynx) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", vjust = -0.5, x.label.fmt = "%4.0f") +
  stat_valleys(colour = "blue") +
  stat_valleys(geom = "text", colour = "blue", vjust = 1.5, x.label.fmt = "%4.0f") +
  ylim(-100, 7300)
```

Rotating the labels.

```{r}
ggplot(lynx, as.numeric = FALSE) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", angle = 66,
             hjust = -0.1, x.label.fmt = "%Y") +
  ylim(NA, 8000)
```

Using `geom_rug` for the peaks and valleys.

```{r}
ggplot(lynx, as.numeric = FALSE) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "rug", colour = "red") +
  stat_valleys(colour = "blue") +
  stat_valleys(geom = "rug", colour = "blue")
```

## stat_quadrant_counts

This statistic automates the annotation of plots with number of observations,
either by quadrant, by pairs of quadrants or the four quadrants taken together.

We generate some artificial data.

```{r}
set.seed(4321)
# generate artificial data
x <- -99:100
y <- x + rnorm(length(x), mean = 0, sd = abs(x))
my.data <- data.frame(x, 
                      y, 
                      group = c("A", "B"))
```

Using defaults except for colour.

```{r}
ggplot(my.data, aes(x, y)) +
  geom_hline(yintercept = 0, colour = "red") +
  geom_vline(xintercept = 0, colour = "red") +
  geom_point() +
  stat_quadrant_counts(colour = "red")
```

Pooling quadrants along the _x_-axis.

```{r}
ggplot(my.data, aes(x, y)) +
  geom_hline(yintercept = 0, colour = "red") +
  geom_point() +
  stat_quadrant_counts(colour = "red", pool.along = "x")
```

Manual positioning of the text annotations and pooling of all four quadrants.

```{r}
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_quadrant_counts(quadrants = 0L, labels.range.x = c(-90, -90), hjust = 0)
```

Annotation of only specific quadrants.

```{r}
ggplot(my.data, aes(x, y)) +
  geom_hline(yintercept = 0, colour = "red") +
  geom_vline(xintercept = 0, colour = "red") +
  geom_point() +
  stat_quadrant_counts(colour = "red", quadrants = c(2, 4))
```
```{r}
ggplot(my.data, aes(x, y)) +
  geom_hline(yintercept = 0, colour = "red") +
  geom_vline(xintercept = 0, colour = "red") +
  geom_point() +
  stat_quadrant_counts(colour = "red") +
  facet_wrap(~group)
```

## stat_poly_eq

We generate some artificial data.

```{r}
set.seed(4321)
# generate artificial data
x <- 1:100
y <- (x + x^2 + x^3) + rnorm(length(x), mean = 0, sd = mean(x^3) / 4)
my.data <- data.frame(x, 
                      y, 
                      group = c("A", "B"), 
                      y2 = y * c(0.5,2),
                      block = c("a", "a", "b", "b"),
                      wt = sqrt(x))
```

First one example using defaults.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(formula = formula, parse = TRUE)
```

`stat_poly_eq()` makes available three different labels in the returned data
frame. One of these is used by default, but `aes()` can be used to select a
different one.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..adj.rr.label..), formula = formula, 
               parse = TRUE)
```

BIC and AIC labels are also returned.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..AIC.label..), 
               formula = formula, 
               parse = TRUE)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), formula = formula, 
               parse = TRUE)
```

Within `aes()` it is possible to _compute_ new labels based on those returned
plus "arbitrary" text. The supplied labels are meant to be _parsed_ into
expressions, so any text added should be valid for a string that will be parsed.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label =  paste(..eq.label.., ..adj.rr.label.., sep = "~~~~")),
               formula = formula, parse = TRUE)
```

Above we inserted spaces between the labels, but it is possible to add also
other characters, even character strings. In this example we use several labels
instead of just two, and separate them with comma followed by a space. Do note
the need to _escape_ the embedded quotation marks.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label =  paste(..eq.label.., ..rr.label.., 
                                  ..AIC.label.., ..BIC.label..,
                                  sep = "*\",\"~~")),
               formula = formula, parse = TRUE)
```

It is also possible to insert arbitrary text, and format it, using `plain()`, `italic()`, `bold()` or `bolditalic()`.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label =  paste(..eq.label.., ..adj.rr.label.., 
                                  sep = "~~italic(\"with\")~~")),
               formula = formula, parse = TRUE)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = paste("atop(", ..AIC.label.., ",", ..BIC.label.., ")", sep = "")), 
               formula = formula, 
               parse = TRUE)
```

Two examples of removing and modifying the default _lhs_ and/or _rhs_ of the equation.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..),
               eq.with.lhs = FALSE,
               formula = formula, parse = TRUE)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..),
               eq.with.lhs = "italic(hat(y))~`=`~",
               formula = formula, parse = TRUE)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  labs(x = expression(italic(z)), y = expression(italic(h)) ) + 
  stat_poly_eq(aes(label = ..eq.label..),
               eq.with.lhs = "italic(h)~`=`~",
               eq.x.rhs = "~italic(z)",
               formula = formula, parse = TRUE)
```

As any valid R expression can be used, Greek letters are also supported, as well
as the inclusion in the label of variable transformations used in the model
formula.

```{r}
formula <- y ~ poly(x, 2, raw = TRUE)
ggplot(my.data, aes(x, log10(y + 1e6))) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..),
               eq.with.lhs = "plain(log)[10](italic(y)+10^6)~`=`~",
               formula = formula, parse = TRUE)
```

A couple of additional examples of polynomials of different orders, and
specified in different ways.

Higher order polynomial.

```{r}
formula <- y ~ poly(x, 5, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), formula = formula, parse = TRUE)
```

Intercept forced to zero.

```{r}
formula <- y ~ x + I(x^2) + I(x^3) - 1
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), formula = formula, 
               parse = TRUE)
```

We give below several examples to demonstrate how other components of the
`ggplot` object affect the behaviour of this statistic.

Facets work as expected either with fixed or free scales. Although bellow we had
to adjust the size of the font used for the equation. In addition to we manually
position the equation label by supplying coordinates.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), size = 3,
               formula = formula, parse = TRUE) +
  facet_wrap(~group)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), size = 3,
               formula = formula, parse = TRUE) +
  facet_wrap(~group, scales = "free_y")
```

Grouping, in this example using the colour aesthetic also works as expected. We
can use justification and supply an absolute location for the equation.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, colour = group)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..),
               formula = formula, parse = TRUE)
```

Label positions relative to the ranges of the _x_ and _y_ scales are also
supported, both through string constants and numeric values in the range
0 to 1.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, colour = group)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..),
               formula = formula, parse = TRUE, label.y.npc = "center")
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, colour = group)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..),
               formula = formula, parse = TRUE, label.y.npc = 0.75)
```

The default locations are now based on normalized coordinates, and consequently
these defaults work even when the range of the _x_ and _y_ scales varies from
panel to panel.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, fill = block)) +
  geom_point(shape = 21, size = 3) +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..rr.label..), size = 3,
               geom = "label", alpha = 0.33,
               formula = formula, parse = TRUE) +
  facet_wrap(~group, scales = "free_y")
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, colour = group, fill = block)) +
  geom_point(shape = 21, size = 3) +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..rr.label..), size = 3,
               geom = "label", alpha = 0.2,
               formula = formula, parse = TRUE,
               label.y.npc = 0.66) +
  facet_wrap(~group, scales = "free_y")
```

## stat_fit_residuals

I had the need to quickly plot residuals matching fits plotted with `geom_smooth()`
using grouping and facets, so a new (simple) statistic was born.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  stat_fit_residuals(formula = formula)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  stat_fit_residuals(formula = formula, resid.type = "working")
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y, color = group)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  stat_fit_residuals(formula = formula)
```

## stat_fit_deviations

As I also had the need to highlight residuals in slides and notes to be used in
teaching, another statistic was born.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_deviations(formula = formula, color = "red") +
  geom_point()
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y, color = group)) +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_deviations(formula = formula) +
  geom_point()
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_deviations(formula = formula, color = "red",
                      arrow = arrow(length = unit(0.015, "npc"), 
                                   ends = "both")) +
  geom_point()
```

## stat_fit_glance

```{r}
# formula <- y ~ poly(x, 3, raw = TRUE)
# broom::augment does not handle poly correctly!
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_glance(method = "lm", 
                  method.args = list(formula = formula),
                  geom = "text",
                  aes(label = signif(..p.value.., digits = 4)))
```

```{r}
# formula <- y ~ poly(x, 3, raw = TRUE)
# broom::augment does not handle poly() correctly!
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y, color = group)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_glance(method = "lm", 
                  method.args = list(formula = formula),
                  geom = "text", 
                  aes(label = paste("P-value = ", signif(..p.value.., digits = 4), sep = "")))
```

```{r}
# formula <- y ~ poly(x, 3, raw = TRUE)
# broom::augment does not handle poly correctly!
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y, color = group)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_glance(method = "lm", 
                  method.args = list(formula = formula),
                  label.x.npc = "right",
                  label.y.npc = "bottom",
                  geom = "text", 
                  aes(label = paste("P-value = ", signif(..p.value.., digits = 4), sep = "")))
```

It is also possible to fit a non-linear model with `method = "nls"`, and any
other model for which a `glance()` method exists. Do consult the documentation 
for package 'broom'. Here we fit the Michaelis-Menten equation to reaction
rate versus concentration data.

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  geom_smooth(method = "nls", 
              formula = micmen.formula,
              se = FALSE) +
  stat_fit_glance(method = "nls", 
                  method.args = list(formula = micmen.formula),
                  geom = "text",
                  aes(label = paste("AIC = ", signif(..AIC.., digits = 3), 
                                    ", BIC = ", signif(..BIC.., digits = 3),
                                    sep = "")))
```

## stat_fit_tidy

This stat makes it possible to add the equation for any fitted model for which
`broom::tidy()` is implemented. Alternatively, individual values such as
estimates for the fitted parameters, standard errors, or _P_-values can be added
to a plot. However, the user has to explicitly construct the labels within
`aes()`. This statistic respects grouping based on aesthetics, and reshapes the
output of `tidy()` so that the values for a given group are in a single row in
the returned `data`.

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  geom_smooth(method = "nls", 
              formula = micmen.formula,
              se = FALSE) +
  stat_fit_tidy(method = "nls", 
                method.args = list(formula = micmen.formula),
                geom = "text",
                label.x.npc = "right",
                label.y.npc = "bottom",
                aes(label = paste("V[m]~`=`~", signif(..Vm_estimate.., digits = 3),
                                  "%+-%", signif(..Vm_se.., digits = 2),
                                  "~~~~K~`=`~", signif(..K_estimate.., digits = 3),
                                  "%+-%", signif(..K_se.., digits = 2),
                                  sep = "")),
                parse = TRUE)
```

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  geom_smooth(method = "nls", 
              formula = micmen.formula,
              se = FALSE) +
  stat_fit_tidy(method = "nls", 
                method.args = list(formula = micmen.formula),
                geom = "text",
                label.x.npc = 0.9,
                label.y.npc = 0.3,
                aes(label = paste("V~`=`~frac(", signif(..Vm_estimate.., digits = 2), "~C,",
                                  signif(..K_estimate.., digits = 2), "+C)",
                                  sep = "")),
                parse = TRUE) +
  labs(x = "C", y = "V")
```

## stat_fit_tb

This stat makes it possible to add summary or ANOVA tables for any fitted model
for which `broom::tidy()` is implemented. The output from `tidy()` is embedded
as a single list value within the returned `data`, an object of class `tibble`.
This statistic **ignores grouping** based on aesthetics. This allows fitting
models when `x` or `y` is a factor (as in such cases `ggplot` splits the data
into groups, one for each level of the factor, which is needed for example for
`stat_summary()` to work as expected). By default, the `"table"` geometry is
used. The use of `geom_table()` is described in a separate section of this User
Guide.

The default output of `stat_fit_tb` is the default output from `tidy(mf)` where
`mf` is the fitted model.

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_tb(method = "lm",
              method.args = list(formula = formula),
              tb.vars = c(Parameter = "term", 
                          Estimate = "estimate", 
                          "s.e." = "std.error", 
                          "italic(t)" = "statistic", 
                          "italic(P)" = "p.value"),
              label.y.npc = "top", label.x.npc = "left",
              parse = TRUE)
```

When `tb.type = "fit.anova"` the output returned is that from `tidy(anova(mf))`
where `mf` is the fitted model.

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_tb(method = "lm",
              method.args = list(formula = formula),
              tb.type = "fit.anova",
              tb.vars = c(Effect = "term", 
                          "df",
                          "M.S." = "meansq", 
                          "italic(F)" = "statistic", 
                          "italic(P)" = "p.value"),
               label.y.npc = "top", label.x.npc = "left",
              parse = TRUE)
```

When `tb.type = "fit.coefs"` the output returned is that of `tidy(mf)` after
selecting the `term` and `estimate` columns.

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_tb(method = "lm",
              method.args = list(formula = formula),
              tb.type = "fit.coefs",
              label.y.npc = "center", label.x.npc = "left")
```

Faceting works as expected, but grouping is ignored as mentioned above. In this
case, the colour aesthetic is not applied to the text of the tables.
Furthermore, if `label.x.npc` or `label.y.npc` are passed numeric vectors of
length > 1, the values are obeyed by the different panels.

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K)
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  facet_wrap(~state) +
  geom_point() +
  geom_smooth(method = "nls",
              formula = micmen.formula,
              se = FALSE) +
  stat_fit_tb(method = "nls",
              method.args = list(formula = micmen.formula),
              tb.type = "fit.coefs",
              label.x.npc = 0.67,
              label.y.npc = c(0.6, 0.3)) +
  theme(legend.position = "none") +
  labs(x = "C", y = "V")
```

The data in the example below are split by `ggplot` into six groups based on the
levels of the `feed` factor. However, as `stat_fit_tb()` ignores groupings, we
can still fit a linear model to all the data in the panel.

```{r}
ggplot(chickwts, aes(factor(feed), weight)) +
  stat_summary(fun.data = "mean_se") +
  stat_fit_tb(tb.type = "fit.anova",
              label.x.npc = 1.07, hjust = 1,
              label.y.npc = "bottom") +
  expand_limits(y = 0)
```

We can flip the system of coordinates, if desired. Be aware that in this case
the specification of the position and justification of the table is no longer
intuitive.

```{r}
ggplot(chickwts, aes(factor(feed), weight)) +
  stat_summary(fun.data = "mean_se") +
  stat_fit_tb(tb.type = "fit.anova",
              label.x.npc = 1.1, hjust = 0,
              label.y.npc = 0, vjust = 1) +
  expand_limits(y = 0) +
  coord_flip()
```

It is also possible to rotate the table using `angle`.

```{r}
ggplot(chickwts, aes(factor(feed), weight)) +
  stat_summary(fun.data = "mean_se") +
  stat_fit_tb(tb.type = "fit.anova",
              angle = 90,
              tb.vars = c(Effect = "term", 
                          "df",
                          "M.S." = "meansq", 
                          "italic(F)" = "statistic", 
                          "italic(P)" = "p.value"),
              label.x.npc = 0.5, hjust = 0.5,
              label.y.npc = 0, vjust = 1,
              parse = TRUE) +
  expand_limits(y = 0) +
  coord_flip()
```

## stat_fit_augment

**Experimental!** 
Use `ggplot2::stat_smooth` instead of `stat_fit_augment` if possible.

For a single panel and no grouping, there is little advantage in using this
statistic compared to the examples in the documentation of package 'broom'. With
grouping and faceting `stat_fit_augment` may occasionally be more convenient
than `ggplot2::stat_smooth` because of its flexibility.

```{r}
# formula <- y ~ poly(x, 3, raw = TRUE)
# broom::augment does not handle poly correctly!
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_fit_augment(method = "lm",
                   method.args = list(formula = formula))
```

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y, color = group)) +
  geom_point() +
  stat_fit_augment(method = "lm", 
                   method.args = list(formula = formula))
```

We can override the variable returned as `y` to be any of the variables in
the data frame returned by `broom::augment` while still preserving the original
`y` values.

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y)) +
  stat_fit_augment(method = "lm",
                   method.args = list(formula = formula),
                   geom = "point",
                   y.out = ".resid")
```

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y, color = group)) +
  stat_fit_augment(method = "lm",
                   method.args = list(formula = formula),
                   geom = "point",
                   y.out = ".std.resid")
```

We can use any model fitting method for which `augment` is implemented.

```{r}
args <- list(formula = y ~ k * e ^ x,
             start = list(k = 1, e = 2))
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  stat_fit_augment(method = "nls",
                   method.args = args)
```

```{r}
args <- list(formula = y ~ k * e ^ x,
             start = list(k = 1, e = 2))
ggplot(mtcars, aes(wt, mpg)) +
  stat_fit_augment(method = "nls",
                   method.args = args,
                   geom = "point",
                   y.out = ".resid")
```

```{r}
args <- list(model = y ~ SSlogis(x, Asym, xmid, scal),
             fixed = Asym + xmid + scal ~1,
             random = Asym ~1 | group,
             start = c(Asym = 200, xmid = 725, scal = 350))
ggplot(Orange, aes(age, circumference, color = Tree)) +
  geom_point() +
  stat_fit_augment(method = "nlme",
                   method.args = args,
                   augment.args = list(data = quote(data)))
```

## stat_dens2d_labels and stat_dens2d_filter

These stats had their origin in an enhancement suggestion for 'ggrepel' from 
Hadley Wickham and discussion with Kamil Slowikowski (ggrepel's author) and 
others. In fact the code is based on code Kamil gave during the discussion,
but simplified and taking a few further ideas from `ggplot::stat_dens2d`.

**Warning!** Which observations are selected by the algorithm used, based on 
`MASS:kde2d`, depends strongly on the values of parameters `h` and `n`. You may need to alter the defaults by passing explicit arguments to these stats. Beware, though, that what are good values, may depend on individual data sets even if they include the same number of observations. For the selection of observations to work cleanly, the argument for `n` must create a dense grid. Much larger values of `n` than in the examples in the documentation of `MASS::kde2d` and `ggplot2::stat_dens2d` will be needed in most cases.

Some random data with random labels.

```{r}
random_string <- function(len = 6) {
paste(sample(letters, len, replace = TRUE), collapse = "")
}

# Make random data.
set.seed(1001)
d <- tibble::tibble(
  x = rnorm(100),
  y = rnorm(100),
  group = rep(c("A", "B"), c(50, 50)),
  lab = replicate(100, { random_string() })
)
```

### stat_dens2d_filter

The stat `stat_dens2d_filter` _filters_ observations, in other words passes to 
the geom a subset of the data received as input. The default value for `geom` is
`"point"`.

Using defaults except for the color aesthetic. Highlight 1/10 of observations 
from lowest density areas of the plot panel.

```{r}
ggplot(data = d, aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red")
```

Highlighting  1/4 of the observations by under-plotting with larger black points.

```{r}
ggplot(data = d, aes(x, y, color = group)) +
   stat_dens2d_filter(keep.fraction = 0.25,
                      size = 3,
                      color = "black") +
   geom_point()
```

A different way of highlighting 1/4 of the observations, using over-plotting with
a 'hollow' shape. We also shift one group with respect to the other.

```{r}
ggplot(data = d, aes(x + rep(c(-2,2), rep(50,2)), 
                     y, color = group)) +
   geom_point() +
   stat_dens2d_filter(shape = 1, size = 3,
                      keep.fraction = 0.25)
```

Highlight 1/4 of observations from lowest density areas of the plot, with
density considered separately for each individual group. In this case grouping
is based on the color aesthetic.

```{r}
ggplot(data = d, aes(x + rep(c(-2,2), rep(50,2)), 
                     y, color = group)) +
   geom_point() +
   stat_dens2d_filter_g(shape = 1, size = 3,
                      keep.fraction = 0.25)
```

Add text labels to 1/10 of the observations. The "text_repel" geom sees only
these observations.

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
  geom_point() + 
  stat_dens2d_filter(geom = "text_repel")
```

Add text labels to 1/2 of the observations. 

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
  geom_point() +
  stat_dens2d_filter(geom = "text_repel", keep.fraction = 0.5)
```

### stat_dens2d_labels

The stat `stat_dens2d_labels` _replaces_ the values of the 
label (aesthetic) variable in data in the high density regions of the panel with the argument passed to `label.fill` before passing them to the geom. The default value for `geom` is `"text"`. The default value of `label.fill` is `""` which results in empty labels, while using `NA` as fill label results in observations being omitted. Using `NA` as `label.fill` is not too different from using `stat_dens2d_filter` as long as the geom used requires a `label` aesthetic.

Label 1/10 of observations from lowest density areas of the plot panels.

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
  geom_point() +
  stat_dens2d_labels()
```

Add text labels to 45% of the observations. 

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
  geom_point() +
  stat_dens2d_labels(keep.fraction = 0.45)
```

When using geom `"text"` we can statically adjust the positioning of labels, but
this is rarely enough even when keeping 1/4 of the labels.

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
  geom_point() +
  stat_dens2d_labels(keep.fraction = 0.25,
                     vjust = -0.3)
```

Using the geoms from package 'ggrepel' avoids clashes among labels or on top of
data points. This works with versions 0.6.0 and newer of 'ggrepel'. One example
with `geom_text_repel` follows.

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
  geom_point() +
  stat_dens2d_labels(geom = "text_repel", 
                     keep.fraction = 0.45)
```

With `geom_label_repel` one needs to use a smaller value for `keep.fracton`, or
a smaller `size`, as labels use more space on the plot than the test alone.

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
  geom_point() +
  stat_dens2d_labels(geom = "label_repel", 
                     keep.fraction = 0.25)
```

Additional arguments can be used to change the angle and position of the text,
but may give unexpected output when labels are long as the repulsion algorithm
"sees" always a rectangular bounding box that is not rotated. With short labels
or angles that are multiples of 90 degrees, there is no such problem.

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
geom_point() +
stat_dens2d_labels(geom = "text_repel",
keep.fraction = 0.25, angle = 90)
```

Using `NA` as fill makes the observations with labels set to `NA` to be skipped
completely, possibly leading to text overlapping the points corresponding to
unlabelled observations around the boundary of the regions where labels are kept
or discarded. We use here `alpha` so that the overlaps can be seen.

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
  geom_point() +
  stat_dens2d_labels(geom = "label_repel", 
                     keep.fraction = 0.35, 
                     alpha = 0.5,
                     label.fill = NA)
```

## geom_table and stat_fmt_table

The `geom_table()` _geometry_ plots a data frame or tibble, nested in a tibble passed as data
argument, using _easthetics_ `x` and `y` for positioning, and `label` for the
data frame containing the data for the table. The table is created as a 'grid'
grob and added as usual to the ggplot object.

```{r}
tb <- tibble(date = ymd(c("2017-05-21", "2007-05-21"), tz = "UCT"),
             value = 1:2 * 10)
data.tb <- tibble(x = 0, y = 8e5, tb = list(tb))
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_table(data = data.tb, aes(label = tb), hjust = 0, vjust = 0) +
  theme_bw()
```

Using `stat_fmt_tb()` we can rename the columns of the tibble as shown below,
and/or select a subset of columns or a slice of rows. 

```{r}
tb <- tibble(date = ymd(c("2017-05-21", "2007-05-21"), tz = "UCT"),
             value = 1:2 * 10)
data.tb <- tibble(x = 0, y = 8e5, tb = list(tb))
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_table(data = data.tb, aes(label = tb), hjust = 0, vjust = 0,
             stat = "fmt_tb", tb.vars = c(Date = "date", Amount = "value")) +
  theme_bw()
```

```{r}
tb <- tibble(date = ymd(c("2017-05-21", "2007-05-21"), tz = "UCT"),
             value = 1:2 * 10)
data.tb <- tibble(x = 25, y = 8e5, tb = list(tb))
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_table(data = data.tb, aes(label = tb), colour = "blue", size = 4) +
  theme_bw()
```

Parsed text, using plot math syntax is supported in the table, with fall-back to
plain text in case of parsing errors, on a cell by cell basis.

```{r}
tb.pm <- tibble(parameter = c("frac(beta[1], a^2)", "frac(beta[2], a^3)"),
             value = c("10^2.4", "10^3.532"))
data.tb <- tibble(x = 12.5, y = 8.5e5, tb = list(tb.pm))
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_table(data = data.tb, aes(label = tb), parse = TRUE) +
  theme_bw()
```

## geom_plot

The `geom_plot()` _geometry_ plots ggplot objects, nested in a tibble passed as data
argument, using _easthetics_ `x` and `y` for positioning, and `label` for the
ggplot object containing the definition of the plot to be nested. A Grob is 
created with 'ggplotGrob' and added as usual to the ggplot object.

```{r}
p <- ggplot(my.data, aes(x, y)) +
  geom_point()

plot.tb <- tibble(x = 25, y = 8e5, plot = list(p %+% theme_bw(8)))

p +
  geom_plot(data = plot.tb, aes(label = plot)) +
  theme_bw(12)
```

As implemented, there is no limitation to the number of insets, and faceting is
respected. If the base plot shows a map, multiple small plots could be
superimposed on different countries or regions. The size of the insets is set
relative to the main plot, so the combined plot can be scaled. A possible
unintuitive but useful feature, is that the theme is linked to each plot.

## geom_grob

The `geom_grob()` _geometry_ plots Grobs (graphical objects as created with
'grid'), nested in a tibble passed as data argument, using _easthetics_ `x` and
`y` for positioning, and `label` for the Grob object. While `geom_table()` and
`geom_plot()` take as values to plot tibbles or data frames, and ggplots,
respectively, and convert them into Grobs before adding them to the plot,
`geom_grob()` expects Grobs ready to be rendered.

```{r}
tiger <- magick::image_read('http://jeroen.github.io/images/tiger.svg')
grobs.tb <- tibble(x = c(0, 10, 20, 40), y = c(4e5, 6e5, 8e5, 9e5),
                   width = c(0.05, 0.05, 0.01, 0.25),
                   height =  c(0.05, 0.05, 0.01, 0.25),
                   grob = list(grid::circleGrob(), 
                               grid::rectGrob(), 
                               grid::textGrob("I am a Grob"),
                               grid::rasterGrob(image = tiger)))

p +
  geom_grob(data = grobs.tb, aes(label = grob, vp.width = width, vp.height = height)) +
  theme_bw(12)
```

## try_tibble

### Time series

Several different formats for storing time series data are used in R. Here we
use in the examples objects of class `ts` but several other classes are
supported as `try.xts()` is used internally. The first example is a quarterly
series.

```{r}
class(austres)
austres.df <- try_tibble(austres)
class(austres.df)
lapply(austres.df, "class")
head(austres.df, 4)
```

The next chunk demonstrates that numeric times are expressed as decimal years in
the returned data frame.

```{r}
austres.df <- try_tibble(austres, as.numeric = TRUE)
lapply(austres.df, "class")
head(austres.df, 4)
```

This second example is for a series of yearly values.

```{r}
class(lynx)
lynx.df <- try_tibble(lynx)
class(lynx.df)
lapply(lynx.df, "class")
head(lynx.df, 3)
```

Above there is a small rounding error of 1 s for these old dates. We can correct
this by rounding to year.

```{r}
lynx.df <- try_tibble(lynx, "year")
head(lynx.df, 3)
```

In addition we can convert the POSIXct values into numeric values in calendar
years plus a decimal fraction.

```{r}
lynx_n.df <- try_tibble(lynx, "year", as.numeric = TRUE)
lapply(lynx_n.df, "class")
head(lynx_n.df, 3)
```

### Other classes

`try_tibble()` attempts to handle gracefully objects that are not time series.

```{r}
try_tibble(1:5)
```

```{r}
try_tibble(letters[1:5])
```

```{r}
try_tibble(factor(letters[1:5]))
```

```{r}
try_tibble(list(x = rep(1,5), y = 1:5))
```

```{r}
try_tibble(data.frame(x = rep(1,5), y = 1:5))
```

```{r}
try_tibble(matrix(1:10, ncol = 2))
```

## Appendix: Additional examples of the use density filtering

We define a function to simplify the generation of random data sets.

```{r}
make_data_tbl <- function(nrow = 100, rfun = rnorm, ...) {
  if (nrow %% 2) {
    nrow <- nrow + 1
  }
  
  set.seed(1001)
  
  tibble::tibble(
    x = rfun(nrow, ...),
    y = rfun(nrow, ...),
    group = rep(c("A", "B"), c(nrow / 2, nrow / 2))
  )
}
```

As we will draw many points on the plotting area we change the default theme
to an uncluttered one.

```{r}
old_theme <- theme_set(theme_bw())
```

In all the examples in this vignette we use colours to demonstrate which data points are selected, but any other suitable aesthetic and discrete scale can be used instead. With `keep.sparse = FALSE` we keep 1/3 of the observations in the denser region of the plot. Although here we first plot all data points and later overplot the selected ones this is not necessary.

```{r}
ggplot(data = make_data_tbl(300), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red", 
                     keep.sparse = FALSE, 
                     keep.fraction = 1/3)
```

Here we highlight the observations split in three group equal groups, each of a
different density of observations.

```{r}
ggplot(data = make_data_tbl(300), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red", 
                     keep.sparse = FALSE, 
                     keep.fraction = 1/3)+
  stat_dens2d_filter(color = "blue", 
                     keep.fraction = 1/3)
```

The algorithm seems to work well also with other distributions, in this example the uniform distribution.

```{r}
ggplot(data = make_data_tbl(300, rfun = runif), aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red", keep.fraction = 1/2)
```

One example with the gamma distribution, which is asymmetric.

```{r}
ggplot(data = make_data_tbl(300, rfun = rgamma, shape = 2), 
       aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red", keep.fraction = 1/3)
```
