---
title: "`ggpmisc` `r packageVersion('ggpmisc')`<br>User Guide"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path = 'figure/guide-pos-', fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4)
options(warnPartialMatchArgs = FALSE)
```

## Preliminaries

```{r}
library(ggpmisc)
library(ggplot2)
library(ggrepel)
library(xts)
library(lubridate)
library(tibble)
library(nlme)
```

## Introduction

Many of the functions, including ggplot statistics and geoms, included in package 
'ggpmisc' had their origin in my need to produce plots for use in teaching.
Some of them are more generally useful, such as `stat_poly_eq()`, but others
like `stat_fit_deviations()` are squarely aimed and producing learning material.
Finally, several statistics for debugging and learning how ggplot statistics
and geoms interact with each other will be of use only to developers of new
statistics and geoms. Function `try_data_frame()` opens the door to easily
converting time series objects into data frames for plotting with `ggplot()`.

## `try_data_frame()`

### Time series

Several different formats for storing time series data are used in R. Here we use in the examples objects of class `ts` but several other classes are supported as `try.xts()` is used internally. The first example is a quarterly series.

```{r}
class(austres)
austres.df <- try_data_frame(austres)
class(austres.df)
lapply(austres.df, "class")
head(austres.df, 4)
```

The next chunk demonstrates that numeric times are expressed as decimal years in the returned data frame.

```{r}
austres.df <- try_data_frame(austres, as.numeric = TRUE)
lapply(austres.df, "class")
head(austres.df, 4)
```

This second example is for a series of yearly values.

```{r}
class(lynx)
lynx.df <- try_data_frame(lynx)
class(lynx.df)
lapply(lynx.df, "class")
head(lynx.df, 3)
```

Above there is a small rounding error of 1 s for these old dates. We can correct this by rounding to year.

```{r}
lynx.df <- try_data_frame(lynx, "year")
head(lynx.df, 3)
```

In addition we can convert the POSIXct values into numeric values in calendar years plus a decimal fraction.

```{r}
lynx_n.df <- try_data_frame(lynx, "year", as.numeric = TRUE)
lapply(lynx_n.df, "class")
head(lynx_n.df, 3)
```

### Other classes

`try_data_frame()` attempts to handle gracefully objects that are not time series.

```{r}
try_data_frame(1:5)
```

```{r}
try_data_frame(letters[1:5])
```

```{r}
try_data_frame(factor(letters[1:5]))
```

```{r}
try_data_frame(list(x = rep(1,5), y = 1:5))
```

```{r}
try_data_frame(data.frame(x = rep(1,5), y = 1:5))
```

```{r}
try_data_frame(matrix(1:10, ncol = 2))
```

## `stat_peaks()` and `stat_valleys()`

Using POSIXct for `time` and the default formatting of labels.

```{r}
ggplot(lynx.df, aes(time, V.lynx)) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", vjust = -0.5) +
  ylim(-100, 7300)
```

Using numeric values for `time` and the default formatting of labels.

```{r}
ggplot(lynx_n.df, aes(time, V.lynx)) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", vjust = -0.5) +
  ylim(-100, 7300)
```

Using POSIXct for `time` but supplying a format string. In addition marking both peaks and valleys.

```{r}
ggplot(lynx.df, aes(time, V.lynx)) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", vjust = -0.5, x.label.fmt = "%Y") +
  stat_valleys(colour = "blue") +
  stat_valleys(geom = "text", colour = "blue", vjust = 1.5, x.label.fmt = "%Y") +
  ylim(-100, 7300)
```

Using numeric for `time` but supplying a format string. In addition marking both peaks and valleys.

```{r}
ggplot(lynx_n.df, aes(time, V.lynx)) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", vjust = -0.5, x.label.fmt = "%4.0f") +
  stat_valleys(colour = "blue") +
  stat_valleys(geom = "text", colour = "blue", vjust = 1.5, x.label.fmt = "%4.0f") +
  ylim(-100, 7300)
```

Rotating the labels.

```{r}
ggplot(lynx.df, aes(time, V.lynx)) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", angle = 66,
             hjust = -0.1, x.label.fmt = "%Y") +
  ylim(NA, 7800)
```

Using `geom_rug` for the peaks and valleys.

```{r}
ggplot(lynx.df, aes(time, V.lynx)) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "rug", colour = "red") +
  stat_valleys(colour = "blue") +
  stat_valleys(geom = "rug", colour = "blue")
```

## `stat_poly_eq()`

We generate some artificial data.

```{r}
set.seed(4321)
# generate artificial data
x <- 1:100
y <- (x + x^2 + x^3) + rnorm(length(x), mean = 0, sd = mean(x^3) / 4)
my.data <- data.frame(x, 
                      y, 
                      group = c("A", "B"), 
                      y2 = y * c(0.5,2),
                      block = c("a", "a", "b", "b"))
```

First one example using defaults.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(formula = formula, parse = TRUE)
```

`stat_poly_eq()` makes available three different labels in the returned data frame. One of these is used by default, but `aes()` can be used to select a different one. 

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..adj.rr.label..), formula = formula, 
               parse = TRUE)
```

BIC and AIC labels are also returned.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..AIC.label..), 
               formula = formula, 
               parse = TRUE)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), formula = formula, 
               parse = TRUE)
```

Within `aes()` it is possible to _compute_ new labels based on those returned plus "arbitrary" text. The supplied labels are meant to be _parsed_ into expressions, so any text added should be valid for a string that will be parsed.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label =  paste(..eq.label.., ..adj.rr.label.., sep = "~~~~")),
               formula = formula, parse = TRUE)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = paste("atop(", ..AIC.label.., ",", ..BIC.label.., ")", sep = "")), 
               formula = formula, 
               parse = TRUE)
```

Two examples of removing and changing the _ohs_ and/or _rhus_ of the equation.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..),
               eq.with.lhs = FALSE,
               formula = formula, parse = TRUE)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..),
               eq.with.lhs = "italic(hat(y))~`=`~",
               formula = formula, parse = TRUE)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  labs(x = expression(italic(z)), y = expression(italic(h)) ) + 
  stat_poly_eq(aes(label = ..eq.label..),
               eq.with.lhs = "italic(h)~`=`~",
               eq.x.rhs = "~italic(z)",
               formula = formula, parse = TRUE)
```

As any valid R expression can be used, Greek letters are also supported, as well as the inclusion in the label of variable transformations used in the model formula.

```{r}
formula <- y ~ poly(x, 2, raw = TRUE)
ggplot(my.data, aes(x, log10(y + 1e6))) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..),
               eq.with.lhs = "plain(log)[10](italic(y)+10^6)~`=`~",
               formula = formula, parse = TRUE)
```

A couple of additional examples of polynomials of different orders, and specified in different ways.

Higher order polynomial.

```{r}
formula <- y ~ poly(x, 5, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), formula = formula, parse = TRUE)
```

Intercept forced to zero.

```{r}
formula <- y ~ x + I(x^2) + I(x^3) - 1
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), formula = formula, 
               parse = TRUE)
```

We give below several examples to demonstrate how other components of the `ggplot` object affect the behaviour of this statistic.

Facets work as expected either with fixed or free scales. Although bellow we had to adjust the size of the font used for the equation. In addition to we manually position the equation label by supplying coordinates.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), size = 3,
               formula = formula, parse = TRUE) +
  facet_wrap(~group)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..), size = 3,
               formula = formula, parse = TRUE) +
  facet_wrap(~group, scales = "free_y")
```

Grouping, in this example using the colour aesthetic also works as expected. We can use justification and supply an absolute location for the equation.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, colour = group)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..),
               formula = formula, parse = TRUE) +
  theme_bw()
```

Label positions relative to the ranges of the _x_ and _y_ scales are also
supported, both through string constants and numeric values in the range
0 to 1.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, colour = group)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..),
               formula = formula, parse = TRUE, label.y.npc = "center") +
  theme_bw()
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, colour = group)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..eq.label..),
               formula = formula, parse = TRUE, label.y.npc = 0.75) +
  theme_bw()
```

The default locations are now based on normalized coordinates, and consequently
these defaults work even when the range of the _x_ and _y_ scales varies from
panel to panel.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, fill = block)) +
  geom_point(shape = 21, size = 3) +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..rr.label..), size = 3,
               geom = "label", alpha = 0.33,
               formula = formula, parse = TRUE) +
  facet_wrap(~group, scales = "free_y") +
  theme_bw()
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y2, colour = group, fill = block)) +
  geom_point(shape = 21, size = 3) +
  geom_smooth(method = "lm", formula = formula) +
  stat_poly_eq(aes(label = ..rr.label..), size = 3,
               geom = "label", alpha = 0.2,
               formula = formula, parse = TRUE,
               label.y.npc = 0.66) +
  facet_wrap(~group, scales = "free_y") +
  theme_bw()
```


## `stat_fit_residuals`

I had the need to quickly plot residuals matching fits plotted with `geom_smooth()`
using grouping and facets, so a new (simple) statistic was born.

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  stat_fit_residuals(formula = formula)
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  stat_fit_residuals(formula = formula, resid.type = "working")
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y, color = group)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  stat_fit_residuals(formula = formula)
```

## `stat_fit_deviations`

As I also had the need to highlight residuals in slides and notes to be used in teaching, 
another statistic was born. 

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_deviations(formula = formula, color = "red") +
  geom_point()
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y, color = group)) +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_deviations(formula = formula) +
  geom_point()
```

```{r}
formula <- y ~ poly(x, 3, raw = TRUE)
ggplot(my.data, aes(x, y)) +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_deviations(formula = formula, color = "red",
                      arrow = arrow(length = unit(0.015, "npc"), 
                                   ends = "both")) +
  geom_point()
```

## `stat_fit_glance`

```{r}
# formula <- y ~ poly(x, 3, raw = TRUE)
# broom::augment does not handle poly correctly!
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_glance(method = "lm", 
                  method.args = list(formula = formula),
                  geom = "text",
                  aes(label = signif(..p.value.., digits = 4)))
```

```{r}
# formula <- y ~ poly(x, 3, raw = TRUE)
# broom::augment does not handle poly() correctly!
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y, color = group)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_glance(method = "lm", 
                  method.args = list(formula = formula),
                  geom = "text", 
                  aes(label = paste("P-value = ", signif(..p.value.., digits = 4), sep = "")))
```

```{r}
# formula <- y ~ poly(x, 3, raw = TRUE)
# broom::augment does not handle poly correctly!
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y, color = group)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula) +
  stat_fit_glance(method = "lm", 
                  method.args = list(formula = formula),
                  label.x.npc = "right",
                  label.y.npc = "bottom",
                  geom = "text", 
                  aes(label = paste("P-value = ", signif(..p.value.., digits = 4), sep = "")))
```

## `stat_fit_augment`

**Experimental!** 
Use `ggplot2::stat_smooth` instead of `stat_fit_augment` if possible.

For a single panel and no grouping, there is little advantage in using this statistic compared to the examples in the documentation of package 'broom'. With grouping and faceting `stat_fit_augment` may occasionally be more convenient than `ggplot2::stat_smooth` because of its flexibility.

```{r}
# formula <- y ~ poly(x, 3, raw = TRUE)
# broom::augment does not handle poly correctly!
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y)) +
  geom_point() +
  stat_fit_augment(method = "lm",
                   method.args = list(formula = formula))
```

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y, color = group)) +
  geom_point() +
  stat_fit_augment(method = "lm", 
                   method.args = list(formula = formula))
```

We can override the variable returned as \code{y} to be any of the variables in the data frame returned by `broom::augment` while still preserving the original _y_ values. 

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y)) +
  stat_fit_augment(method = "lm",
                   method.args = list(formula = formula),
                   geom = "point",
                   y.out = ".resid")
```

```{r}
formula <- y ~ x + I(x^2) + I(x^3)
ggplot(my.data, aes(x, y, color = group)) +
  stat_fit_augment(method = "lm",
                   method.args = list(formula = formula),
                   geom = "point",
                   y.out = ".std.resid")
```

We can use any model fitting method for which `augment` is implemented.

```{r}
args <- list(formula = y ~ k * e ^ x,
             start = list(k = 1, e = 2))
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  stat_fit_augment(method = "nls",
                   method.args = args)
```

```{r}
args <- list(formula = y ~ k * e ^ x,
             start = list(k = 1, e = 2))
ggplot(mtcars, aes(wt, mpg)) +
  stat_fit_augment(method = "nls",
                   method.args = args,
                   geom = "point",
                   y.out = ".resid")
```

```{r}
args <- list(model = y ~ SSlogis(x, Asym, xmid, scal),
             fixed = Asym + xmid + scal ~1,
             random = Asym ~1 | group,
             start = c(Asym = 200, xmid = 725, scal = 350))
ggplot(Orange, aes(age, circumference, color = Tree)) +
  geom_point() +
  stat_fit_augment(method = "nlme",
                   method.args = args,
                   augment.args = list(data = quote(data)))
```

## `stat_dens2d_labels` and `stat_dens2d_filter`

These stats had their origin in an enhancement suggestion for 'ggrepel' from 
Hadley Wickham and discussion with Kamil Slowikowski (ggrepel's author) and 
others. In fact the code is based on code Kamil gave during the discussion,
but simplified and taking a few further ideas from `ggplot::stat_dens2d`.

**Warning!** Which observations are selected by the algorithm used, based on 
`MASS:kde2d`, depends strongly on the values of parameters `h` and `n`. As these
stats have been tested only with a few data sets, you may need to alter the
defaults. Beware, though, that what are good values, may depend on individual 
data sets even if they include the same number of observations. For the
selection of observations to work cleanly, the argument for `n` must create a
dense grid. Much larger values of `n` than in the examples in the documentation
of `MASS::kde2d` and `ggplot2::stat_dens2d` will be needed in most cases.

Some random data with random labels.

```{r}
random_string <- function(len = 6) {
paste(sample(letters, len, replace = TRUE), collapse = "")
}

# Make random data.
set.seed(1001)
d <- tibble::tibble(
  x = rnorm(100),
  y = rnorm(100),
  group = rep(c("A", "B"), c(50, 50)),
  lab = replicate(100, { random_string() })
)
```

### `stat_dens2d_filter`

The stat `stat_dens2d_filter` _filters_ observations, in other words passes to 
the geom a subset of the data received as input. The default value for `geom` is
`"point"`.

Using defaults except for the color aesthetic. Highlight 1/10 of observations 
from lowest density areas of the plot panel.

```{r}
ggplot(data = d, aes(x, y)) +
  geom_point() +
  stat_dens2d_filter(color = "red")
```

Highlighting  1/4 of the observations by under-plotting with larger black points.

```{r}
ggplot(data = d, aes(x, y, color = group)) +
   stat_dens2d_filter(keep.fraction = 0.25,
                      size = 3,
                      color = "black") +
   geom_point()
```

A different way of highlighting 1/4 of the observations, using over-plotting with
a 'hollow' shape. We also shift one group with respect to the other.

```{r}
ggplot(data = d, aes(x + rep(c(-2,2), rep(50,2)), 
                     y, color = group)) +
   geom_point() +
   stat_dens2d_filter(shape = 1, size = 3,
                      keep.fraction = 0.25)
```

Highlight 1/4 of observations from lowest density areas of the plot, with
density considered separately for each individual group, in this case based on
the color aesthetic.

```{r}
ggplot(data = d, aes(x + rep(c(-2,2), rep(50,2)), 
                     y, color = group)) +
   geom_point() +
   stat_dens2d_filter_g(shape = 1, size = 3,
                      keep.fraction = 0.25)
```

Add text labels to 1/10 of the observations. The "text_repel" geom sees only
these observations.

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
  geom_point() + 
  stat_dens2d_filter(geom = "text_repel")
```

Add text labels to 1/2 of the observations. 

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
  geom_point() +
  stat_dens2d_filter(geom = "text_repel", keep.fraction = 0.5)
```

### `stat_dens2d_labels`

The stat `stat_dens2d_labels` _replaces_ with `label.fill` the values of the 
label (aesthetic) variable in data in the high density regions of the panel 
before passing them to the geom. The default value for `geom` is `"text"`. The
default value of `label.fill` is `""` which results in empty labels, while using
`NA` as fill label results in observations being omitted. Using `NA` as
`label.fill` is not too different from using `stat_dens2d_filter` as long as the
geom used requires a `label` aesthetic.

Label 1/10 of observations from lowest density areas of the plot panels.

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
  geom_point() +
  stat_dens2d_labels()
```

Add text labels to 45% of the observations. 

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
  geom_point() +
  stat_dens2d_labels(keep.fraction = 0.45)
```

When using geom `"text"` we can statically adjust the positioning of labels, but
this is rarely enough even when keeping 1/4 of the labels.

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
  geom_point() +
  stat_dens2d_labels(keep.fraction = 0.25,
                     vjust = -0.3)
```

Using the geoms from package 'ggrepel' avoids clashes among labels or on top of data points. This works with versions 0.6.0 and newer of 'ggrepel'. One example with `geom_text_repel` follows. 

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
  geom_point() +
  stat_dens2d_labels(geom = "text_repel", 
                     keep.fraction = 0.45)
```

With `geom_label_repel` one needs to use a smaller value for `keep.fracton`, or a smaller `size`, as labels use more space on the plot than the test alone.

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
  geom_point() +
  stat_dens2d_labels(geom = "label_repel", 
                     keep.fraction = 0.25)
```

Additional arguments can be used to change the angle and position of the text, but may give unexpected output when labels are long as the repulsion algorithm "sees" always a rectangular bounding box that is not rotated. With short labels or angles that are multiples of 90 degrees, there is no such problem.

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
geom_point() +
stat_dens2d_labels(geom = "text_repel",
keep.fraction = 0.25, angle = 90)
```

Using `NA` as fill makes the observations with labels set to `NA` to be skipped completely, possibly leading to text overlapping the points corresponding to unlabelled observations around the boundary of the regions where labels are kept or discarded. We use here `alpha` so that the overlaps can be seen.

```{r}
ggplot(data = d, aes(x, y, label = lab, color = group)) +
  geom_point() +
  stat_dens2d_labels(geom = "label_repel", 
                     keep.fraction = 0.35, 
                     alpha = 0.5,
                     label.fill = NA)
```
