---
title: "User Guide: 4 Manipulation of plot layers"
subtitle: "'ggpmisc' `r packageVersion('ggpmisc')`"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{User Guide: 4 Manipulation of plot layers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path = 'figure/debug-pos-', fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4)
options(warnPartialMatchArgs = FALSE)
```

## Preliminaries

```{r}
library(ggplot2)
library(ggpmisc)
library(tibble)
```

We generate some artificial data.

```{r}
set.seed(4321)
# generate artificial data
my.data <- data.frame(
  group = factor(rep(letters[1:4], each = 30)),
  panel = factor(rep(LETTERS[1:2], each = 60)),
  y = rnorm(40)
)
```

We change the default theme to an uncluttered one.

```{r}
old_theme <- theme_set(theme_bw())
```

## Introduction

The functions described here are not expected to be useful in everyday plotting
as one can simply change the order in which layers are added to a ggplot.
However, if one uses high level methods or functions that automatically produce
a full plot using 'ggplot2' internally, one may need to add, move or delete
layers so as to profit from such canned methods and retain enough flexibility.

In a `ggplot` object, layers reside in a list, and their positions in the list
determine the plotting order when generating the graphical output. The Grammar
of Graphics treats the list of layers as a _stack_ using only _push_ operations.
In other words, always the most recently added layer resides at the end of the
list, and over-plots all layers previously added. The functions described in
this vignette allow overriding the **normal** syntax at the cost of breaking the
expectations of the grammar. These functions are, as told above, to be used only
in exceptional cases. This notwithstanding, they are rather easy to use. The
user interface is consistent across all of them. Moreover, they are designed to
return objects that are identical to objects created using the normal syntax
rules of the Grammar of Graphics**.

Function | Use                     
-------- | -------------------------------------
`delete_layers()` | delete one or more layers 
`append_layers()` | append layers at a specific position 
`move_layers()`   | move layers to an absolute position
`shift_layers()`  | move layers to a relative position
`which_layers()`  | obtain the index positions of layers
`extract_layers()`   | extract matched or indexed layers
`num_layers()`    | obtain number of layers
`top_layer()`    | obtain position of top layer
`bottom_layer()`    | obtain position of bottom layer

Although their definitions does not rely on code internal to 'ggplot2', they 
rely on the internal structure of objects belonging to class `gg` and `ggplot`.
Consequently, long-term backwards compatibility cannot be guaranteed.

## Examples

### Preliminaries

We generate a plot to be used later to demonstrate the use of the functions.

```{r}
p <- ggplot(my.data, aes(group, y)) + 
  geom_point() +
  stat_summary(fun.data = mean_se, colour = "red", size = 1.3) +
  facet_wrap(~panel, scales = "free_x", labeller = label_both)
p
```

To display textual information about a `gg` object we use method `summary()`,
while methods `print()` and `plot()` will display the actual plot.

```{r}
summary(p)
```

Layers in a plot are stored in a list. The default `print()` method for a list
of layers displays only a small part of the information in a layer.

```{r}
p$layers
```

To see all the fields, one needs to use `str()`, which we use here for a single layer.

```{r}
str(p$layers[[1]])
```

### Manipulation of plot layers

We start by using `which_layers()` as it produces simply a vector of indexes into
the list of layers. The third statement is useless here, but demonstrates how
layers are selected in all the functions described in this document.

```{r}
which_layers(p, "GeomPoint")
which_layers(p, "StatSummary")
which_layers(p, idx = 1L)
```

We can also easily extract matching layers with `extract_layers()`. Here one layer
is returned, and displayed using the default `print()` method. Method `str()`
can be used as shown above.

```{r}
extract_layers(p, "GeomPoint")
```

With `delete_layers()` we can remove layers from a plot, selecting them using 
the match to a class, as shown here, or by a positional index. This was shown
above for `which_layers()`.

```{r}
delete_layers(p, "GeomPoint")
```

```{r}
delete_layers(p, idx = 1L)
```

```{r}
delete_layers(p, "StatSummary")
```

With `move_layers()` we can alter the stacking order of layers. The layers to
move are selected in the same way as in the examples above, while `position`
gives where to move the layers to. Two character strings, `"top"` and `"bottom"`
are accept as `position` argument, as well as `integer`s. In the later case, 
the layer(s) is/are appended after the supplied position with reference to the 
list of layers not being moved. 

```{r}
move_layers(p, "GeomPoint", position = "top")
```

Same operation but using a relative position. A positive value for `shift` is 
intertreted as an upward displacement and a negative one as downwards 
displacement. 

```{r}
shift_layers(p, "GeomPoint", shift = +1)
```

Here we show how to add a layer behind all other layers.

```{r}
append_layers(p, geom_line(colour = "orange", size = 1), position = "bottom")

```

It is also possible to append the new layer immediately above an arbitrary
existing layer using a numeric index, which as shown here can be also obtained
by matching to a class name. In this example we insert a new layer in-between
two layers already present in the plot. As with the `+` operator of the Grammar
of Graphics, `object` also accepts a list of layers as argument (no example
shown).

```{r}
append_layers(p, object = geom_line(colour = "orange", size = 1), 
              position = which_layers(p, "GeomPoint"))
```

Annotations add layers, so they can be manipulated in the same way as other layers.

```{r}
p1 <- p + 
  annotate("text", label = "text label", x = 1.1, y = 0, hjust = 0)
p1
```

```{r}
delete_layers(p1, "GeomText")
```

