---
title: "Multiple comparisons"
subtitle: "'ggpmisc' `r packageVersion('ggpmisc')`"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
---

```{r, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(fig.align = 'center', 
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = 'hold', fig.width = 7, fig.height = 5,
                      dev = "svg")

options(warnPartialMatchArgs = FALSE,
        tibble.print.max = 4,
        tibble.print.min = 4,
        dplyr.summarise.inform = FALSE)

eval_flag <- TRUE # evaluate all code chunks
```

## Aims of 'ggpmisc' and caveats

Package 'ggpmisc' makes it easier to add to plots created using 'ggplot2'
annotations based on fitted models and other statistics. It does this by
wrapping existing model fit and other functions. The same annotations can be
produced by calling the model fit functions, extracting the desired estimates
and adding them to plots. There are two advantages in wrapping these functions
in an extension to package 'ggplot2': 1) we ensure the coupling of graphical
elements and the annotations by building all elements of the plot using the same
data and a consistent grammar and 2) we make it easier to annotate plots to the
casual user of R, already familiar with the grammar of graphics.

_To avoid confusion it is good to make clear what may seem obvious to some: if
no plot is needed, then there is no reason to use this package. The values shown
as annotations are not computed by 'ggpmisc' but instead by the usual model-fit
and statistical functions from R and R packages. The same is true for model
predictions, residuals, etc. that some of the functions in 'ggpmisc' display as
lines, segments, or other graphical elements._

It is also important to remember that in most cases data analysis including
exploratory and other stages should take place before annotated plots for
publication are produced. Even though data analysis can benefit from combined
numerical and graphical representation of the results, the use I envision for
'ggpmisc' is mainly for the production of plots for publication or
communication. In either case, whether used for analysis or communication, it is
crucial that users cite and refer both to 'ggpmisc' and to the underlying R and
R packages when publishing plots created with functions and methods from
'ggpmisc'.

```{r}
print(citation(package = "ggpmisc", auto = TRUE), bibtex = FALSE)
```

# Multiple comparisons

In experimental designs with more than two groups or treatments, we face the
problem that testing every possible pair for significant difference, includes a
test of the smallest vs. the largest observed value. This means that the
probability of at least one pair being significant is no longer the one computed
in a _t_-test or _LSD_ (least significant difference). We need to distinguish
the probability of making the wrong decision in an individual comparison
(comparison-wise) from that for the experiment as a whole (experiment-wise, the
number of wrong decisions per experiment).

There are different approaches, that can be grouped into methods that attempt to
control the experiment-wise probability of false positive outcomes to a given
_P_-level, such as 0.05 (adjusted _P_-values), and those that attempt to control
the false discovery rate (FDR), that can be thought of as the proportion of the
positive outcomes that are false positives. There are variations in both
approaches, varying in power and conservativeness.

A correction needs to be applied to _P_-values or _LSD_. Different methods are
available for this. Methods differ in their assumptions and difficulty of
computation. Tukey's _HSD_ (honestly significant difference) and the Bonferroni
correction are well known approaches. Bonferroni's method is a "quick and dirty"
calculation but nowadays not recommended as it applies an unnecessarily strong
correction.

_Dunnet contrasts_ test all treatments individually against a control condition.
_Tukey contrasts_ targets the case of testing all possible pairs. It is also
possible to set-up contrasts of interest individually, or testing for other
differences, such as basing them on the average of related treatments.

In general, one should test for significant differences only those contrasts
selected _a priori_ as of interest. The fewer the pairs tested for significance,
the smaller a correction needs to be applied, thus increasing the power of
the test for the cases of interest. _Of course, for validity, which tests are
of interest should be decided independently of the data._

_Multiple comparisons are frequently applied as post-hoc tests (after a
different test of significance with a broader scope). As post-hoc tests, they
are seen as a way of investigating the source of an overall significant
difference, and multiple comparisons are applied only if the main effect of the
factor is in itself significant and the additional contrasts skipped otherwise.
In other cases multiple comparisons can be the primary statistical test
addressing the research hypotheses. Both approaches are valid as long as the
hypotheses have been set independently of the data being used in the test._

# Plot annotations for multiple comparison tests

The letter labels and labelled segments used to highlight pairwise comparisons
are a special case as they behave as data labels along the axis onto which an
explanatory factor has been mapped, usually _x_, but frequently as annotations
along the axis onto which a continuous numeric variable has been mapped, usually
_y_.

Layer function `stat_multcomp()` first fits a model (a linear model by default)
followed by a multiple comparisons' test with a user-controlled adjustment to
the _P_-values. The implementation makes use of function `glht()` from package
[multcomp](http://multcomp.r-forge.r-project.org/) making available all the
methods it supports for the adjustment of _P_-values for multiple comparisons.
Currently, `stat_multcomp()` only implements pairwise contrasts, as in this case
there are well established approaches to plot annotations. By passing a numeric
matrix as argument arbitrary sets of _pairwise_ contrasts can be specified.
_Contrasts that are not pairwise, i.e., that involve more than two levels of a
factor, are not yet supported.

The returned value and default geometry depend on the type of label, that can be
either bars (connecting segments) labelled with _P_-values or other parameters, 
or letters.

# Examples

Package [ggpmisc](https://docs.r4photobiology.info/ggpmisc/) imports and 
re-exports all definitions from [ggpp](https://docs.r4photobiology.info/ggpp/)
as well as from [ggplot2](https://ggplot2.tidyverse.org/), so it is enough to attach explicitly package [ggpmisc](https://docs.r4photobiology.info/ggpmisc/) . All three packages are available through CRAN.

```{r, message=FALSE}
library(ggpmisc)
theme_set(theme_bw())
```

## Using default labels

Tukey's method for all pairwise contrasts using _honestly significant differences_.

```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_summary(fun.data = mean_cl_normal, colour = "red") +
  stat_multcomp(label.y = 12, 
                size = 2.75, 
                vstep = 0.05) +
  expand_limits(y = 0)
```

Using Holm's method to adjust _P_-values.

```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_summary(fun.data = mean_cl_normal, colour = "red") +
  stat_multcomp(p.adjust.method = "holm",
                label.y = 12, 
                size = 2.75, 
                vstep = 0.05) +
  expand_limits(y = 0)
```
Dunnet's method for comparison of each treatment against a control, assumed to be the first level of the factor.

```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_summary(fun = mean, geom = "col", width = 0.5) +
  stat_summary(fun.data = mean_cl_normal, colour = "red") +
  stat_multcomp(label.y = 25, 
                size = 2.75,
                contrasts = "Dunnet")
```

A staircase of pairwise contrasts with _P_-values adjusted using Holm's method.

```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_summary(fun = mean, geom = "col", width = 0.5) +
  stat_summary(fun.data = mean_cl_normal, colour = "red") +
  stat_multcomp(label.y = c(18, 23, 24),
                size = 2.75,
                contrasts = rbind(c(0, 0, -1, 1),
                                  c(0, -1, 1, 0),
                                  c(-1, 1, 0, 0)))
```
Tukey's pairwise contrasts using defaults except for the position of the lowermost bar.

```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_summary(fun = mean, geom = "col", width = 0.5) +
  stat_summary(fun.data = mean_cl_normal, colour = "red") +
  stat_multcomp(size = 2.5, label.y = 25)
```

Tukey's pairwise contrasts using defaults and shown with letters.

```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_summary(fun = mean, geom = "col", width = 0.5) +
  stat_summary(fun.data = mean_cl_normal, colour = "red") +
  stat_multcomp(label.y = -1, 
                label.type = "letters")
```

```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_summary(fun = mean, geom = "col", width = 0.5) +
  stat_summary(fun.data = mean_cl_normal, colour = "red") +
  stat_multcomp(colour = "white", 
                label.type = "letters",
                adj.method.tag = 0)
```

```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_multcomp(size = 2.75) +
  stat_boxplot(width = 1/3)
```

```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_multcomp(size = 2.85,
                label.y = 11,
                geom = "text_pairwise",
                vstep = 0.07,
                p.digits = 2,
                contrasts = "Dunnet") +
  stat_boxplot(width = 1/3) +
  expand_limits(y = 0)
```

```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_multcomp(label.type = "letters",
                geom = "label",
                size = 2.75) +
  stat_boxplot(width = 1/3)
```

Numeric _P_-values shown for Dunnet contrasts.

```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_summary(fun = mean, geom = "col", width = 0.5) +
  stat_summary(fun.data = mean_cl_normal, colour = "red") +
  stat_multcomp(aes(x = stage(start = factor(cyl), 
                              after_stat = x.right.tip)),
                geom = "text",
                label.y = -1, 
                vstep = 0,
                size = 3,
                contrasts = "Dunnet")
```

## Using pre-built labels other than default


```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_multcomp(use_label(c("delta", "P")),
                size = 2.75,
                label.y = 11,
                vstep = 0.08,
                p.digits = 2,
                contrasts = "Dunnet") +
  stat_boxplot(width = 1/3) +
  expand_limits(y = 0)
```

```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_multcomp(use_label(c("t", "P")),
                          size = 2.75,
                label.y = 11,
                vstep = 0.08,
                contrasts = "Dunnet") +
  stat_boxplot(width = 1/3) +
  expand_limits(y = 0)
```

```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_summary(fun = mean, geom = "col", width = 0.5) +
  stat_summary(fun.data = mean_cl_normal, colour = "red") +
  stat_multcomp(aes(x = stage(start = factor(cyl), after_stat = x.right.tip),
                    label = after_stat(stars.label)),
                geom = "text",
                label.y = -1, 
                vstep = 0,
                contrasts = "Dunnet")
```

## Using other aesthetics than the default

```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_multcomp(aes(colour = after_stat(p.value) < 0.05),
                size = 2.75,
                label.y = 11,
                vstep = 0.08,
                p.digits = 2,
                contrasts = "Dunnet") +
  stat_boxplot(width = 1/3) +
  scale_colour_manual(values = c("grey50", "blue")) +
  expand_limits(y = 0)
```

```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_multcomp(aes(fill = after_stat(p.value) < 0.05),
                size = 2.75,
                label.y = 11,
                vstep = 0.08,
                p.digits = 2,
                contrasts = "Dunnet") +
  stat_boxplot(width = 1/3) +
  scale_fill_manual(values = c("grey90", "lightblue")) +
  expand_limits(y = 0)
```

```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_multcomp(aes(colour = factor(after_stat(letters.label))),
                label.type = "letters",
                size = 4) +
  stat_boxplot(width = 1/3) +
  expand_limits(y = 0)
```


```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
  stat_multcomp(aes(colour = factor(after_stat(letters.label))),
                label.type = "letters",
                geom = "point",
                size = 3,
                adj.method.tag = 0) +
  stat_boxplot(width = 1/3) +
  expand_limits(y = 0)
```

