---
title: "Non-linear model fits"
subtitle: "'ggpmisc' `r packageVersion('ggpmisc')`"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
---

```{r, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(fig.align = 'center', 
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = 'hold', fig.width = 7, fig.height = 5)

options(warnPartialMatchArgs = FALSE,
        tibble.print.max = 4,
        tibble.print.min = 4,
        dplyr.summarise.inform = FALSE)

eval_flag <- TRUE # evaluate all code chunks
```

## Aims of 'ggpmisc' and caveats

Package 'ggpmisc' makes it easier to add to plots created using 'ggplot2' annotations based on fitted models and other statistics. It does this by wrapping existing model fit and other functions from base R and various R packages. The same annotations can be produced by calling the model fit functions, extracting the desired estimates and adding them to plots. There are two advantages in wrapping these functions in an extension to package 'ggplot2': 1) it ensures the coupling of graphical elements and the annotations by building all elements of the plot using the same data and a consistent grammar and 2) it makes it easier to annotate plots to the casual user of R, already familiar with the grammar of graphics.

_To avoid confusion it is good to make clear what may seem obvious to some: if no plot is needed, then there is no reason to use this package. The values shown as annotations are not computed by 'ggpmisc' but instead by the usual model-fit and statistical functions from R and other packages. The same is true for model predictions, residuals, etc. that some of the functions in 'ggpmisc' display as lines, segments, or other graphical elements._

It is also important to remember that in most cases data analysis including exploratory and other stages should take place before annotated plots for publication are produced. Even though data analysis can benefit from combined numerical and graphical representation of the results, the use I envision for 'ggpmisc' is mainly for the production of plots for publication or communication. In any case, whether plots created with functions and methods from 'ggpmisc' are used for analysis or communication, it is crucial that users cite and refer both to 'ggpmisc' and to the underlying R and R packages used to fit models or to do other statistical computations.

```{r}
print(citation(package = "ggpmisc", auto = TRUE), bibtex = FALSE)
```

## Introduction

Layer functions in 'ggpmisc' can be used for fitting non-linear models, but
given the diversity of possible model formulas, automatic generation of
annotations for the model fit equations have not been implemented.

This article gives examples of non-linear model fits using both
`stat_poly_line()` together with `stat_poly_eq()`, and with the 'broom' based
statistics, `stat_fit_augnment`, `stat_fit_tidy()` and `stat_fit_glance`. In
earlier version of 'ggpmisc' only the 'broom'-based approach was available.
However, in 'ggpmisc' (>= 1.0.0) `stat_poly_eq()` and `stat_poly_line()` can
also be used to fit non-linear models.

## Preliminaries

We load the packages used in the examples. Those imported always, and those suggested, when available.

```{r, message=FALSE}
library(ggpmisc)
library(tibble)
library(dplyr)
library(quantreg)

eval_nlme <-  requireNamespace("nlme", quietly = TRUE)
if (eval_nlme) library(nlme)
eval_broom <-  requireNamespace("broom", quietly = TRUE)
if (eval_broom) library(broom)
eval_broom_mixed <-  requireNamespace("broom.mixed", quietly = TRUE)
if (eval_broom_mixed) library(broom.mixed)
eval_gginnards <-  requireNamespace("gginnards", quietly = TRUE)
if (eval_gginnards) library(gginnards)
```

As we will add text and labels onto the plotting area we change the default theme to an uncluttered one with a white canvas.

```{r}
old_theme <- theme_set(theme_bw())
```

## `nls()` with 'ggpmisc' native approach

Easiest is to use formatted `character` strings to be parsed into `plotmath`
expressions. We can in this case create the mapping to the `label` aesthetic
with function `use_label()` (shown below) of with function `f_use_label()` (not 
shown).

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  stat_poly_line(method = "nls", 
                 formula = micmen.formula,
                 se = FALSE) +
  stat_poly_eq(use_label("AIC", "BIC"),
               method = "nls", 
               formula = micmen.formula,
               label.x = "centre", label.y = "bottom") +
  labs(x = "C", y = "V")
```

No pre-formatted equation label is available, and we have to assemble it from
the numeric values of the fitted parameter estimates.

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  stat_poly_line(method = "nls", 
                 formula = micmen.formula,
                 se = FALSE) +
  stat_poly_eq(aes(label =
                     after_stat(
                       paste("V~`=`~frac(",
                             signif(b_0, digits = 2), "~C,",
                             signif(b_1, digits = 2), "+C)",
                             sep = ""))),
               parse = TRUE,
               method = "nls",
               output.type = "numeric",
               formula = micmen.formula,
               vstep = 0.1, size = 3,
               label.x = "centre", label.y = "bottom") +
  labs(x = "C", y = "V")
```

## `nls()` with 'broom'-based approach

The statistics based on methods from package 'broom' always return only numeric values and formatted strings need to be assembled in user's R code within a call to `aes()`. The examples below produce figures nearly identical to the two first examples in the previous section. 

```{r, eval=eval_broom}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  stat_smooth(method = "nls", 
              formula = micmen.formula,
              se = FALSE) +
  stat_fit_glance(aes(label = 
                        after_stat(
                          paste("AIC = ", signif(AIC, digits = 3), 
                                ", BIC = ", signif(BIC, digits = 3),
                                sep = ""))),
                  method = "nls", 
                  method.args = list(formula = micmen.formula),
                  label.x = "centre", label.y = "bottom") +
  labs(x = "C", y = "V")
```

```{r, eval=eval_broom}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  stat_smooth(method = "nls", 
              formula = micmen.formula,
              se = FALSE) +
  stat_fit_tidy(aes(label =
                      after_stat(
                        paste("V~`=`~frac(",
                              signif(Vm_estimate, digits = 2), "~C,",
                              signif(K_estimate, digits = 2), "+C)",
                              sep = ""))),
                parse = TRUE,
                method = "nls", 
                method.args = list(formula = micmen.formula),
                size = 3,
                label.x = "center",
                label.y = "bottom",
                vstep = 0.12) +
  labs(x = "C", y = "V")
```

