---
title: "Non-linear model fits"
subtitle: "'ggpmisc' `r packageVersion('ggpmisc')`"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
---

```{r, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(fig.align = 'center', 
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = 'hold', fig.width = 7, fig.height = 5)

options(warnPartialMatchArgs = FALSE,
        tibble.print.max = 4,
        tibble.print.min = 4,
        dplyr.summarise.inform = FALSE)

eval_flag <- TRUE # evaluate all code chunks
```

## Aims of 'ggpmisc' and caveats

Package 'ggpmisc' makes it easier to add to plots created using 'ggplot2' annotations based on fitted models and other statistics. It does this by wrapping existing model fit and other functions from base R and various R packages. The same annotations can be produced by calling the model fit functions, extracting the desired estimates and adding them to plots. There are two advantages in wrapping these functions in an extension to package 'ggplot2': 1) it ensures the coupling of graphical elements and the annotations by building all elements of the plot using the same data and a consistent grammar and 2) it makes it easier to annotate plots to the casual user of R, already familiar with the grammar of graphics.

_To avoid confusion it is good to make clear what may seem obvious to most readers: if no plot is needed, then there is no reason to use this package. The values shown as annotations are not computed by 'ggpmisc' but instead by the usual model-fit and statistical functions from R and other packages. The same is true for model predictions, residuals, etc. that some of the functions in 'ggpmisc' display as lines, segments, or other graphical elements._

It is also important to remember that in most cases data analysis including exploratory and other stages should take place before annotated plots for publication are produced. Even though data analysis can benefit from combined numerical and graphical representation of the results, the use I envision for 'ggpmisc' is mainly for the production of plots for publication or communication. In any case, whether plots created with functions and methods from 'ggpmisc' are used for analysis or communication, it is crucial that users cite and refer both to 'ggpmisc' and to the underlying R and R packages used to fit models or to do other statistical computations.

```{r}
print(citation(package = "ggpmisc", auto = TRUE), bibtex = FALSE)
```

## Introduction

Layer functions in 'ggpmisc' can be used for fitting non-linear models as well
as linear models, but given the diversity of possible model formulas, automatic
generation of annotations for the model fit equations have not been implemented.

Whether confidence bands can be plotted for the fitted model lines depends on
whether the `predict()` method for the fitted model object computes them. This
is not the case for `nls()` and `onls()` functions described in this article.
There are ways of calculating them, one being through bootstrapping, which is
computation intensive as implemented, for example, in R package 'nlraa'. These
approaches are currently not supported by the layer functions in 'ggpmisc'. In
the case of `nls()` and `onls()` informative messages are issued if `se = TRUE`
is passed to `stat_poly_eq()`, but in other cases, errors with criptic 
messages can be triggered in 'ggplot2' code.

This article gives examples of non-linear model fits using both
`stat_poly_line()` together with `stat_poly_eq()`, and with the 'broom' based
statistics, `stat_fit_augnment`, `stat_fit_tidy()` and `stat_fit_glance`. In
'ggpmisc' (< 1.0.0) only the 'broom'-based approach was available.
However, in 'ggpmisc' (>= 1.0.0) `stat_poly_eq()` and `stat_poly_line()` can
also be used to fit a broader range of models, including non-linear models.

## Preliminaries

We load the packages used in the examples. Those imported always, and those suggested, when available.

```{r, message=TRUE}
library(ggpmisc)
library(tibble)
library(dplyr)
library(quantreg)

eval_onls <-  requireNamespace("onls", quietly = TRUE)
if (eval_onls) library(onls) else message("Please, install 'onls'")
eval_nlme <-  requireNamespace("nlme", quietly = TRUE)
if (eval_nlme) library(nlme) else message("Please, install 'nlme'")
eval_broom <-  requireNamespace("broom", quietly = TRUE)
if (eval_broom) library(broom) else message("Please, install 'broom'")
eval_broom_mixed <-  requireNamespace("broom.mixed", quietly = TRUE)
if (eval_broom_mixed) library(broom.mixed) else message("Please, install 'broom.mixed'")
eval_gginnards <-  requireNamespace("gginnards", quietly = TRUE)
if (eval_gginnards) library(gginnards) else message("Please, install 'gginnards'")
```

As we will add text and labels onto the plotting area we change the default theme to an uncluttered one with a white canvas.

```{r}
old_theme <- theme_set(theme_bw())
```

## Non linear least-squares regression

The fitting of models that are non-linear in their parameters are fit using iterative approximation algorithms, as no analytical solution is available.
These methods are computation intensive, and require starting values for the
estimation of parameters. Geometrically, the deviations used to compute the
mean squares, are not different to those used for the analytical solution to 
estimation of parameters of linear models by ordinary least squares (OLS).

In R, `nls()` is the workhorse for non-linear least-squares regression. R also defines _self starting_ functions describing non-linear models. These objects include both the function to be fit, and a function implementing a method of obtaining starting values for the iterative algorithm. Starting values are rough approximations, not useful by themselves, but are good enough as starting points for the search for better estimates.

**Non linearity in the parameters** _In this context non-linearity does not refer to the shape of the fitted model when plotted. Polynomials are models linear in their parameters as these parameters are multiplier of the terms in the model equation. When the parameters being estimated appear in the model formula as power exponents, or similar, then we call such models non-linear in their parameters._

### `nls()` with 'ggpmisc' native approach

Function `SSmicmen()` is a self starting function for the Michaelis-Menthen equation for the rate of a chemical reaction $V$ as a function of the concentrtaion, $C$, of its precursor. It has two parameters, $V_m$, the asymptotic maximum $V$, and $K$, a reaction constant.


Easiest is to use formatted `character` strings to be parsed into `plotmath`
expressions. We can in this case create the mapping to the `label` aesthetic
with function `use_label()` (shown below) of with function `f_use_label()` (not 
shown).

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  stat_poly_line(method = "nls", 
                 formula = micmen.formula) +
  stat_poly_eq(use_label("AIC", "BIC"),
               method = "nls", 
               formula = micmen.formula,
               label.x = "centre", label.y = "bottom") +
  labs(x = "C", y = "V")
```

No pre-formatted equation label is available, and we have to assemble it from
the numeric values of the fitted parameter estimates.

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  stat_poly_line(method = "nls", 
                 formula = micmen.formula) +
  stat_poly_eq(aes(label =
                     after_stat(
                       paste("V~`=`~frac(",
                             signif(b_0, digits = 3), "~C,",
                             signif(b_1, digits = 3), "+C)",
                             sep = ""))),
               parse = TRUE,
               method = "nls",
               output.type = "numeric",
               formula = micmen.formula,
               vstep = 0.1, size = 3,
               label.x = "centre", label.y = "bottom") +
  labs(x = "C", y = "V")
```

Deviations can be highlighted.

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  stat_poly_line(method = "nls", 
                 formula = micmen.formula) +
  stat_fit_deviations(
               method = "nls",
               formula = micmen.formula) +
  labs(x = "C", y = "V")
```

Residuals can be plotted


```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  stat_fit_residuals(
               method = "nls",
               formula = micmen.formula) +
  labs(x = "C", y = "V residuals")
```

There is rarely a reason to do so, but `nls()` can be also used to fit linear models as it is a generally applicable procedure.

```{r}
poly.formula <- y ~ b0 + b1 * x + b2 * I(x^2)
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  stat_poly_line(method = "nls", 
                 method.args = list(start = list(b0 = 0, b1 = 1, b2 = 0)),
                 formula = y ~ b0 + b1 * x + b2 * I(x^2),
                 se = FALSE) +
  labs(x = "C", y = "V")
```


### `nls()` with 'broom'-based approach

The statistics based on methods from package 'broom' always return numeric values and formatted strings need to be assembled for all parameters in user's R code within a call to `aes()`. The examples below produce figures nearly identical to the two first examples in the previous section. 

```{r, eval=eval_broom}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  stat_poly_line(method = "nls", 
                 formula = micmen.formula) +
  stat_fit_glance(aes(label = 
                        after_stat(
                          paste("AIC = ", signif(AIC, digits = 3), 
                                ", BIC = ", signif(BIC, digits = 3),
                                sep = ""))),
                  method = "nls", 
                  method.args = list(formula = micmen.formula),
                  label.x = "centre", label.y = "bottom") +
  labs(x = "C", y = "V")
```

```{r, eval=eval_broom}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  stat_poly_line(method = "nls", 
                 formula = micmen.formula) +
  stat_fit_tidy(aes(label =
                      after_stat(
                        paste("V~`=`~frac(",
                              signif(Vm_estimate, digits = 3), "~C,",
                              signif(K_estimate, digits = 3), "+C)",
                              sep = ""))),
                parse = TRUE,
                method = "nls", 
                method.args = list(formula = micmen.formula),
                size = 3,
                label.x = "center",
                label.y = "bottom",
                vstep = 0.12) +
  labs(x = "C", y = "V")
```

## Orthogonal Nonlinear Least-Squares Regression

Orthogonal least-squares is similar in concept to model II and MA regression, but applicable to models that when plotted are curves, including models non-linear in their parameters. These methods are useful when explanatory variables are measured with error, and the intention is to describe a relationship. This is a common case in allometry (the estimation of the relative size of different parts of organs in an organism) and other data with similar properties. These and related methods, are usually described as for _errors in variables_ (EIV) problems.

R package 'onls' implements function `onls()` that works similarly to R's `nls()` but which does not make use of the self-starting estimates, which have always to be supplied explicitly.

The fit is done in two steps, NLS fit followed by the ONLS fit. The values from both fits are returned.

I use this data set here but in this context NLS is the correct approach. In fact only one of NLS or ONLS can be expected to suit a given data set. Anyway, it remains didactically useful to compare the application of both methods to the same data set.

### `onls()` with 'ggpmisc' native approach

Easiest is to use formatted `character` strings to be parsed into `plotmath`
expressions. We can in this case create the mapping to the `label` aesthetic
with function `use_label()` (shown below) of with function `f_use_label()` (not 
shown).

```{r, eval=eval_onls}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  stat_poly_line(method = "onls", 
                 method.args = list(start = list(Vm = 200, K = 0.1)),
                 formula = micmen.formula) +
  stat_poly_eq(use_label("AIC", "BIC"),
               method = "onls", 
               method.args = list(start = list(Vm = 200, K = 0.1)),
               formula = micmen.formula,
               label.x = "centre", label.y = "bottom") +
  labs(x = "C", y = "V")
```

No pre-formatted equation label is available, and we have to assemble it from
the numeric values of the fitted parameter estimates.

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  stat_poly_line(method = "onls", 
                 method.args = list(start = list(Vm = 200, K = 0.1)),
                 formula = micmen.formula) +
  stat_poly_eq(aes(label =
                     after_stat(
                       paste("V~`=`~frac(",
                             signif(b_0, digits = 3), "~C,",
                             signif(b_1, digits = 3), "+C)",
                             sep = ""))),
               parse = TRUE,
               method = "onls",
                 method.args = list(start = list(Vm = 200, K = 0.1)),
               output.type = "numeric",
               formula = micmen.formula,
               vstep = 0.1, size = 3,
               label.x = "centre", label.y = "bottom") +
  labs(x = "C", y = "V")
```

Deviations can be highlighted

So these deviations or residuals are **not** those whose squares have been minimized, but instead, computed as is OLS but relative to the curve fitted by ONOLS, i.e., using residuals orthogonal to the fitted curve.

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  stat_poly_line(method = "onls", 
                 method.args = list(start = list(Vm = 200, K = 0.1)),
                 formula = micmen.formula) +
  stat_fit_deviations(method = "onls",
                      method.args = list(start = list(Vm = 200, K = 0.1)),
                      formula = micmen.formula) +
  labs(x = "C", y = "V")
```

The same residuals as highlighted above can be plotted. However, _these are not the orthogonal residuals used to fit the model_.

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  stat_fit_residuals(
               method = "onls",
               method.args = list(start = list(Vm = 200, K = 0.1)),
               formula = micmen.formula) +
  labs(x = "C", y = "V residuals")
```

Function  `onls()` can also be used to fit linear models using the orthogonal least-squares approach.

```{r}
poly.formula <- y ~ b0 + b1 * x + b2 * I(x^2)
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  stat_poly_line(method = "onls", 
                 method.args = 
                   list(start = list(b0 = 0, b1 = 1, b2 = 0)),
                 formula = poly.formula,
                 se = FALSE) +
  labs(x = "C", y = "V")
```

### `onls()` with 'broom'-based approach

The statistics based on methods from package 'broom' always return only numeric values and formatted strings need to be assembled in user's R code within a call to `aes()`. The examples below produce figures nearly identical to the two first examples in the previous section. 

```{r, eval=eval_broom}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  stat_poly_line(method = "onls", 
                 method.args = list(start = list(Vm = 200, K = 0.1)),
                 formula = micmen.formula) +
  stat_fit_glance(aes(label = 
                        after_stat(
                          paste("AIC = ", signif(AIC, digits = 3), 
                                ", BIC = ", signif(BIC, digits = 3),
                                sep = ""))),
                  method = "onls", 
                  method.args = list(formula = micmen.formula,
                                     start = list(Vm = 200, K = 0.1)),
                  label.x = "centre", label.y = "bottom") +
  labs(x = "C", y = "V")
```

```{r, eval=eval_broom}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  geom_point() +
  stat_poly_line(method = "onls", 
                 method.args = list(start = list(Vm = 200, K = 0.1)),
                 formula = micmen.formula) +
  stat_fit_tidy(aes(label =
                      after_stat(
                        paste("V~`=`~frac(",
                              signif(Vm_estimate, digits = 3), "~C,",
                              signif(K_estimate, digits = 3), "+C)",
                              sep = ""))),
                parse = TRUE,
                method = "onls", 
                method.args = list(formula = micmen.formula,
                                   start = list(Vm = 200, K = 0.1)),
                size = 3,
                label.x = "center",
                label.y = "bottom",
                vstep = 0.12) +
  labs(x = "C", y = "V")
```

## Comparison of ONLS to NLS

```{r}
micmen.formula <- y ~ SSmicmen(x, Vm, K) 
ggplot(Puromycin, aes(conc, rate, colour = state)) +
  stat_poly_line(method = "nls", 
                 formula = micmen.formula,
                 linetype = "dashed",
                 limit.to = seq(0, 1.2, length.out = 80)) +
  stat_poly_line(method = "onls", 
                 method.args = list(start = list(Vm = 200, K = 0.1)),
                 formula = micmen.formula,
                 se = FALSE,
                 limit.to = seq(0, 1.2, length.out = 80)) +
  labs(x = "C", y = "V")
```

## Limitations and caveats

Non-linear models are fit by approximation using an iterative approach. Convergence depends on the "goodness" of the guesses for parameter values passed as a named list to `start`. Solutions may be not unique and local optima can derail the search for the global best fit, and it is recommended, when not using self starting functions to repeat the fit with different starting values. When some parameter values can lead to computation errors, boundaries must be set so that the search avoids these "forbidden" conditions. In addition, different algorithms used for minimization differ in computational efficiency but also in the problems for which they may fail completly. In case of failure, it can be worthwhile to try a different `algorithm`.

Even if these methods are supported, model fitting can easily fail. Thus, if errors are encountered, it is necessary to run `nls()` or `onls()` on their own, to diagnose the reason of the failure to converge on a solution, arithmetic errors, or problems.

As the stats fit the model one data group at a time, when debugging, it is necessary to subset the data in the same way as `ggplot()` does.

```{r}
fm1 <- nls(rate ~ SSmicmen(conc, Vm, K),
           data = subset(Puromycin, state == "treated"))
summary(fm1)
coef(fm1)
predict(fm1)
```


```{r}
fm2 <- onls(rate ~ SSmicmen(conc, Vm, K),
            start = list(Vm = 200, K = 0.1),
            data = subset(Puromycin, state == "treated"))
summary(fm2)
coef(fm2)
predict(fm2)
```
