---
title: "Custom Polynomial Models"
subtitle: "'ggpmisc' `r packageVersion('ggpmisc')`"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
---

```{r, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(fig.align = 'center', 
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = 'hold', fig.width = 7, fig.height = 7)

options(warnPartialMatchArgs = FALSE,
        tibble.print.max = 4,
        tibble.print.min = 4,
        dplyr.summarise.inform = FALSE)

eval_flag <- TRUE # evaluate all code chunks
```

## Aims of 'ggpmisc' and caveats

Package 'ggpmisc' makes it easier to add to plots created using 'ggplot2' annotations based on fitted models and other statistics. It does this by wrapping existing model fit and other functions. The same annotations can be produced by calling the model fit functions, extracting the desired estimates and adding them to plots. There are two advantages in wrapping these functions in an extension to package 'ggplot2': 1) we ensure the coupling of graphical elements and the annotations by building all elements of the plot using the same data and a consistent grammar and 2) we make it easier to annotate plots to the casual user of R, already familiar with the grammar of graphics.

_To avoid confusion it is good to make clear what may seem obvious to some: if no plot is needed, then there is no reason to use this package. The values shown as annotations are not computed by 'ggpmisc' but instead by the usual model-fit and statistical functions from R and R packages. The same is true for model predictions, residuals, etc. that some of the functions in 'ggpmisc' display as lines, segments, or other graphical elements._

It is also important to remember that in most cases data analysis including exploratory and other stages should take place before annotated plots for publication are produced. Even though data analysis can benefit from combined numerical and graphical representation of the results, the use I envision for 'ggpmisc' is mainly for the production of plots for publication or communication. In case case, whether used for analysis or communication, it is crucial that users cite and refer both to 'ggpmisc' and to the underlying R and R packages when publishing plots created with functions and methods from 'ggpmisc'.

```{r}
print(citation(package = "ggpmisc", auto = TRUE), bibtex = FALSE)
```

## Introduction

The first chapter gives examples of the use of the *stats* from 'ggpmisc' for plotting predictions and annotating ggplots using robust and resistant regression .  
The second chapter gives examples of the use of the *stats* from 'ggpmisc' for plotting predictions and annotating ggplots with fitted linear, cubic and polynomial splines.
The third chapter gives examples of the use of the *stats* from 'ggpmisc' for plotting predictions and annotating ggplots with functions wrapping model fit methods together with model selection and other code that can automate the choice of the model formula, model fit method, etc., on a per data group basis.

### Fitted models in `stat_poly_line()` and `stat_poly_eq()`

Both functions are designed for flexibility while still attempting to keep their implementation and use rather straightforward.

_Function `stat_poly_eq()` requires that the model `formula` is a regular polynomial (with no missing intermediate power terms), with the exception that the intercept can be missing._ Function `stat_poly_line()` has fewer requirements for the argument passed to `formula`. 

Both statistics:

  - Accept any R function or its name as argument for `method`,
  - as long as the fitted model object belongs to classes `"lm"`, `"rlm"`, `"lqs"`, `"lts"` or `"gls"`, as tested with `inherits()`, and additionally models of other classes as long as query methods are available.
  - Their implementation tolerates some missing methods and fields in model fit objects and their summary, i.e., returning `NA_real_` or empty `character` strings depending on the case.
  - The formula used to construct the equation is, whenever possible, extracted from the fitted model object. Only exceptionally, with this approach fails, the formula passed as argument for `formula` in the call is used with a warning.

This design makes it possible to use as `method` existing model fit functions that return an object of one of the classes specifically supported, wrapper functions that call such functions, or in some other way return an object of a _known_ class such as `"lm"` or a class with suitable methods implemented.

To showcase this, I explore the use of contributed and user defined model fit functions through examples. Several of the examples use facets or grouping to demonstrate how the models fitted depend on the data. The approach can be also useful in the absence of facets and/or groups, for example for dynamic reports or dashboards.

Two frequently used model fit functions that return model fit objects that require special handling to extract values are supported in 'ggpmisc' by specific statistics: `stat_ma_line()` and `stat_ma_eq()` for mayor axis regression with package 'lmodel2', and `stat_quant_line()`, `stat_quant_band()` and `stat_quant_equation()` for quantile regression with package 'quantreg'. Their use is documented in the User Guide and their help pages.

### Preliminaries

Other packages are loaded in the section they are used.

```{r, message=FALSE}
library(ggpmisc)
```

Attaching package 'ggpmisc' also attaches package 'ggpp' as it provides several of the geometries used by default in the statistics described below. Package 'ggpp' can be loaded and attached on its own, and has [separate documentation](https://docs.r4photobiology.info/ggpp/).

As we will use text and labels on the plotting area we change the default theme to an uncluttered one.

```{r}
old_theme <- theme_set(theme_bw())
```

## Robust and resistant methods

We consider cases in which fitting linear models with _ordinary least squares_ (OLS) as criterion is unsuitable. 

In the case of outliers, many methods are available, and each fit function usually also supports several variations within a class of methods. In _robust regression_ the weight of extreme observations in the computation of the penalty function is decreased compared to OLS. The effect of such decrease can be described as weights ranging continuously between 0 and 1, either actually used in the fit algorithm or computed _a posteriori_ by comparison to OLS. In _resistant regression_, some observations are ignored altogether, which is equivalent to discrete weights taking one of two values, 0 or 1. Even if these approaches work automatically, similarly as when discarding observations manually, it is important to investigate the origin of outliers.

**When using robust and resistant methods it is crucial to report the results of analyses clearly showing the weights used for different observations.** One of the goals of 'ggpmisc' is to make this step easy. The `data` objects returned by the stats ftpm 'ggpmisc' contain, whenever they are available, both _prior_ `weights` and `robustness.weights`. The former being those passed by mapping a variable to the `weight` aesthetic and used as argument for the `weights` parameter of model fit function, and the latter being those actually used in the final fit, possibly computed.

### Robust regression ('MASS')

```{r}
print(citation(package = "MASS", auto = TRUE), bibtex = FALSE)
```

```{r}
library(MASS)
```

Package 'MASS' provides function `rlm()` for "Robust Fitting of Linear Models". The returned value is an `"rlm"` object. Class `"rlm"` is derived from `"lm"` so the stats work mostly as expected. However, we need to be aware that no $R^2$ or $P$ estimates are available and the corresponding labels are empty strings.

The first example plot is in the style I recommend for use in reporting as it makes the weights visible, later plots from this same subset of the data are mostly useful in teaching and data exploration.

```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  stat_poly_line(method = "rlm") +
  stat_poly_eq(use_label("eq", "n", "method"), 
               label.x = "right",
               method = "rlm") +
  stat_fit_deviations(method = "rlm",
                      mapping = aes(colour = after_stat(robustness.weights)),
                      show.legend = TRUE,
                      geom = "point") +
  scale_color_gradient(low = "red", high = "blue", limits = c(0, 1)) +
  theme(legend.position = "top")
```

Fitted values and deviations highlighted.

```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  stat_poly_line(method = "rlm") +
  stat_fit_fitted(method = "rlm", shape = "cross") +
  stat_fit_deviations(method = "rlm", linewidth = 0.3) +
  stat_fit_deviations(method = "rlm",
                      mapping = aes(colour = after_stat(robustness.weights)),
                      show.legend = TRUE,
                      geom = "point")+
  scale_color_gradient(low = "red", high = "blue", limits = c(0, 1)) +
  theme(legend.position = "top")
```

Residuals plotted.

```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  stat_fit_residuals(method = "rlm",
                     resid.type = "response",
                     mapping = aes(colour = after_stat(robustness.weights)),
                     show.legend = TRUE,
                     geom = "point") +
  labs(y = "Response residuals") +
  scale_color_gradient(low = "red", high = "blue", limits = c(0, 1)) +
  theme(legend.position = "top")
```

```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  stat_fit_residuals(method = "rlm",
                     resid.type = "pearson",
                     mapping = aes(colour = after_stat(robustness.weights)),
                     show.legend = TRUE,
                     geom = "point") +
  labs(y = "Pearson residuals") +
  scale_color_gradient(low = "red", high = "blue", limits = c(0, 1)) +
  theme(legend.position = "top")
```

We are not constrained to a first degree polynomial. This plot follows the recommended style for reporting.

```{r, fig.width = 7, fig.height = 5}
my.formula <- y ~ poly(x, 2, raw = TRUE)
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  stat_poly_eq(use_label("eq", "n", "method"), 
               formula = my.formula,
               label.x = "right",
               method = "rlm") +
  stat_poly_line(method = "rlm", 
                 formula = my.formula,
                 colour = "black") +
  stat_fit_deviations(mapping = aes(colour = after_stat(robustness.weights)),
                      method = "rlm",
                      formula = my.formula,
                      show.legend = TRUE,
                      geom = "point") +
  scale_color_gradient(low = "red", high = "blue", limits = c(0, 1)) +
  theme(legend.position = "top")
```

We can the same approach with the whole data set using facets.

```{r}
#| fig.asp: 1.25
ggplot(mpg, aes(displ, hwy)) +
  stat_poly_line(method = "rlm") +
  stat_poly_eq(method = "rlm", mapping = use_label("eq", "n"),
               label.x = "right") +
    stat_fit_deviations(method = "rlm",
                      mapping = aes(colour = after_stat(robustness.weights)),
                      show.legend = TRUE,
                      geom = "point") +
  facet_wrap(~class, ncol = 2) +
  scale_color_gradient(low = "red", high = "blue", limits = c(0, 1)) +
  theme(legend.position = "top")
```

### Robust regression ('robustbase')

```{r}
print(citation(package = "robustbase", auto = TRUE), bibtex = FALSE)
```

```{r}
library(robustbase)
```

Package 'robustbase' provides function `lmrob()` for "MM-type Estimators for Linear Regression". The returned value is a `"lmrob"` object. Class `"lmrob"` is derived from `"lm"` so the stats work mostly as expected. However, we need to be aware that no $P$ estimate is available and the corresponding labels are empty strings.

A few examples are given, as the only change to the code from the previous section is in the argument passed to parameter `method`. The first plot is what I consider suitable for use in reports. Using defaults, `lmrob` seems to be more aggresive against "outliers" than `rlm`, at least for these data.

```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  stat_poly_line(method = "lmrob") +
  stat_poly_eq(use_label("eq", "n", "method"), 
               label.x = "right",
               method = "lmrob") +
  stat_fit_deviations(method = "lmrob",
                      mapping = aes(colour = after_stat(robustness.weights)),
                      show.legend = TRUE,
                      geom = "point") +
  scale_color_gradient(low = "red", high = "blue", limits = c(0, 1)) +
  theme(legend.position = "top")
```

Fitted values and deviations highlighted.

```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  stat_poly_line(method = "lmrob") +
  stat_fit_fitted(method = "lmrob", shape = "cross") +
  stat_fit_deviations(method = "lmrob", linewidth = 0.3) +
  stat_fit_deviations(method = "lmrob",
                      mapping = aes(colour = after_stat(robustness.weights)),
                      show.legend = TRUE,
                      geom = "point")+
  scale_color_gradient(low = "red", high = "blue", limits = c(0, 1)) +
  theme(legend.position = "top")
```

Residuals plotted.


```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  stat_fit_residuals(method = "lmrob",
                     resid.type = "response",
                     mapping = aes(colour = after_stat(robustness.weights)),
                     show.legend = TRUE,
                     geom = "point") +
  labs(y = "Response residuals") +
  scale_color_gradient(low = "red", high = "blue", limits = c(0, 1)) +
  theme(legend.position = "top")
```

### Resistant regression ('robustbase')

```{r}
print(citation(package = "robustbase", auto = TRUE), bibtex = FALSE)
```

```{r}
library(robustbase)
```

Package 'robustbase' provides function `ltsReg()` for "least trimmed squares" regression. The returned value is an `"lts"` object. The usual accessors functions are available for Class `"lts"`.

Package 'MASS' provides function `lqs()` for "Resistant Regression". The recommended `method` for this function is `"lts"`. Thus, using function `ltsReg()` from package 'robustbase' is preferable as not all the usual accessors are available when using `lqs()`. If needed, `lqs()` is also, partly supported by 'ggpmisc', but no example is shown.

A few examples are given, as the only change to the code from the previous sections is in the argument passed to parameter `method`. The first plot is what I consider suitable for use in reports.

```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  stat_poly_line(method = "lts") +
  stat_poly_eq(use_label("eq", "n", "method"), 
               label.x = "right",
               method = "lts") +
  stat_fit_deviations(method = "lts",
                      mapping = aes(colour = after_stat(robustness.weights)),
                      show.legend = TRUE,
                      geom = "point") +
  scale_color_gradient(low = "red", high = "blue", limits = c(0, 1)) +
  theme(legend.position = "top")
```

Fitted values and deviations highlighted.


```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  stat_poly_line(method = "lts") +
  stat_fit_fitted(method = "lts", shape = "cross") +
  stat_fit_deviations(method = "lts", linewidth = 0.3) +
  stat_fit_deviations(method = "lts",
                      mapping = aes(colour = after_stat(robustness.weights)),
                      show.legend = TRUE,
                      geom = "point")+
  scale_color_gradient(low = "red", high = "blue", limits = c(0, 1)) +
  theme(legend.position = "top")
```

Residuals plotted.

```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  stat_fit_residuals(method = "lts",
                     mapping = aes(colour = after_stat(robustness.weights)),
                     show.legend = TRUE,
                     geom = "point") +
  labs(y = "Response residuals") +
  scale_color_gradient(low = "red", high = "blue", limits = c(0, 1)) +
  theme(legend.position = "top")
```

In the examples above the argument passed was `method = "lts"`, which is a convenience short notation equivalent to `method = robustbase::ltsReg`.

## Use of variance covariates

_Generalised least squares_ (GLS) is an extension to OLS that makes it possible to describe how the variance of the response _y_ varies. GLS supports description of heterogeneous variance by means of variance covariates and correlation structures.

### General Least Squares ('nlme')

```{r}
print(citation(package = "nlme", auto = TRUE), bibtex = FALSE)
```

```{r}
library(nlme)
```

In 'ggpmisc' (>= 0.6.2) function `gls()` from package 'nlme' is supported natively by functions `stat_poly_line()` and `stat_poly_eq()`. In this example, using a variance covariate. _Currently, support is limited in 'ggpmisc' to the prediction line,
and labels.

```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
   geom_point() +
   stat_poly_line(method = "gls", 
                  method.args = list(weights = nlme::varPower()), 
                  se = FALSE) +
   stat_poly_eq(method = "gls", 
                method.args = list(weights = nlme::varPower()),
                mapping = use_label("eq", "n", "method"),
   label.x = "right")
```

Fitted values and deviations highlighted.

```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
   stat_poly_line(method = "gls", 
                  method.args = list(weights = nlme::varPower()), 
                  se = FALSE) +
  stat_fit_fitted(method = "gls", 
                  method.args = list(weights = nlme::varPower()),
                  shape = "cross") +
  stat_fit_deviations(method = "gls", 
                  method.args = list(weights = nlme::varPower()),
                  linewidth = 0.3) +
  geom_point()
```

Residuals plotted.

```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  stat_fit_residuals(method = "gls", 
                     method.args = list(weights = nlme::varPower()),
                     resid.type = "response") +
  labs(y = "Response residuals")
```

```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  stat_fit_residuals(method = "gls", 
                     method.args = list(weights = nlme::varPower()),
                     resid.type = "pearson") +
  labs(y = "Pearson residuals")
```

```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  stat_fit_residuals(method = "gls", 
                     method.args = list(weights = nlme::varPower()),
                     resid.type = "normalized") +
  labs(y = "Normalized residuals")
```

## Splines

Splines are used when a flexible curve is required to efficiently describe a relationship between variables. Splines are composed of segments connected at knots. The simplest possible are linear splines where each segment is a straight line. In higher degree splines the segments are polynomials and conditions are imposed to ensure continuity at knots even for their derivative, to ensure smooth transition at each not. Except for linear splines the fitted parameters for each segment and not locations are too complex to be informative.

Even when model fit equations are not useful or are not generated automatically, other labels constructed by `stat_poly_eq()` remain useful. This stat not only returns ready made labels but also numeric values for parameter estimates.

### Linear Splines ('lspline')

```{r}
print(citation(package = "lspline", auto = TRUE), bibtex = FALSE)
```

```{r}
library(lspline)
```

Package 'lspline' provides functions that can be used to fit linear splines with function `lm()`. These functions are called from the model formula, and do not fit the change points where the knots are located. These functions, `lspline()`, `qlspline()` amd `elspline()` can be used directly in the model formula passed to `stat_poly_line()`. They can be used also with `stat_poly_eq()` but the equation label returned is incorrect, with the terms for the different segments displayed as the terms for the powers of a polynomial. Thus, it is necessary to construct in the call to `aes()` the equations for the different segments. The other labels are correctly generated.

```{r, fig.width = 7, fig.height = 5}
cp <- 4.1 # a guess
my.formula <- y ~ lspline::lspline(x, knots = cp)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  stat_poly_eq(aes(label = 
                     after_stat(
                       paste("\"L : \"*y~`=`~", round(b_0, 2),
                           round(b_1, 2), "~x*\"; R: \"*y~`=`~", 
                           round(b_0 + cp * b_1, 2), 
                           round(b_2, 2), "~x",
                           sep = ""))), 
               parse = TRUE, output.type = "numeric",
               label.x = "right",
               formula = my.formula,
               method = "lm") +
  stat_poly_eq(use_label("R2", "P", "AIC"), 
               label.x = "right",
               label.y = 0.87,
               formula = my.formula,
               method = "lm") +
  stat_poly_line(method = "lm", 
                 formula = my.formula)
```

Plotting fitted values and observations.

```{r, fig.width = 7, fig.height = 5}
ggplot(mpg, aes(displ, hwy)) +
  stat_poly_line(method = "lm", 
                 formula = my.formula) +
  geom_point() +
  stat_fit_fitted(method = "lm", formula = my.formula, 
                  shape = "cross", alpha = 0.5)
```

Adding deviations.

```{r, fig.width = 7, fig.height = 5}
ggplot(mpg, aes(displ, hwy)) +
  stat_poly_line(method = "lm", 
                 formula = my.formula) +
  geom_point() +
  stat_fit_deviations(method = "lm", 
                      formula = my.formula, 
                      linewidth = 0.25) +
  stat_fit_fitted(method = "lm", formula = my.formula, 
                  shape = "cross", alpha = 0.5)
```

Plotting residuals.

```{r, fig.width = 7, fig.height = 5}
ggplot(mpg, aes(displ, hwy)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  stat_fit_residuals(method = "lm", 
                     formula = my.formula) +
  labs(y = "Response residuals")
```

It is possible to use `rlm()` or `lmrob()` instead of `lm()`, i.e., a robust approach instead of OLS. Here we use `method = "lmrob"`, combining the approach from the example of robust regression with the linear spline fitting above.

```{r, fig.width = 7, fig.height = 5}
cp <- 4.1
my.formula <- y ~ lspline(x, knots = cp)

ggplot(mpg, aes(displ, hwy)) +
  stat_fit_deviations(formula = my.formula,
                      method = "lmrob",
                      mapping = aes(colour = after_stat(robustness.weights)),
                      show.legend = TRUE,
                      geom = "point") +
  stat_poly_eq(aes(label =
                     after_stat(
                     paste("\"L : \"*y~`=`~", round(b_0, 2),
                           round(b_1, 2), "~x",
                           "*\"; R: \"*y~`=`~", 
                           round(b_0 + cp * b_1, 2), 
                           round(b_2, 2), "~x",
                           sep = ""))), 
               parse = TRUE, 
               output.type = "numeric",
               label.x = "right",
               formula = my.formula,
               method = "lmrob") +
  stat_poly_eq(use_label("R2", "n", "method"), 
               label.x = "right",
               label.y = 0.87,
               formula = my.formula,
               method = "lmrob") +
  stat_poly_line(method = "lmrob", 
                 formula = my.formula) +
  scale_color_gradient(low = "red", high = "blue", limits = c(0, 1)) +
  theme(legend.position = "top")
```

Plotting residuals.

```{r, fig.width = 7, fig.height = 5}
ggplot(mpg, aes(displ, hwy)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  stat_fit_residuals(method = "lmrob", 
                     formula = my.formula,
                     mapping = aes(colour = after_stat(robustness.weights)),
                     show.legend = TRUE) +
  scale_color_gradient(name = "Weights", 
                       low = "red", high = "blue", 
                       limits = c(0, 1),
                       guide = "colourbar")
```

### Linear Splines ('segmented')

Package 'segmented' has a design that makes it easy to use. The fitted model
used has a structure consistent with that returned by `lm()`. Any differences
are smoothed by query/extract functions that are consistent with those for
`"lm"` objects. As segmented linear models are not polynomials, no `eq.label`
is automatically generated. Another difference is that the location of change
points is estimated and returned. Starting from 'ggpmisc' (>= 0.6.3) this
information is returned by `stat_poly_eq()` and can be used to in user-assembled
labels as shown in the examples below.

```{r}
print(citation(package = "segmented", auto = TRUE), bibtex = FALSE)
```

```{r}
library(segmented)
```

Function `segreg()` makes it possible to fit LM and GLM models that are
segmented. Differently to 'lspline', the location of the change points or spline
knots is estimated during model fitting in addition to estimating segment slopes
and intercept. As 'lspline' it uses a model formula to set the model including
control of the segmentation through `seg()`. Differently to `lspline()`,
`segreg()` is a model fit function that returns a fitted model object with
additional information on change points and slopes.

The model `formula = y ~ seg(x, npsi = 1)` requests a fit with a single change
point, which being the default is equivalent to `formula = y ~ seg(x)`.

```{r, fig.width = 7, fig.height = 5}
my.formula <- y ~ seg(x, npsi = 1)

ggplot(mpg, aes(displ, hwy)) +
  stat_poly_line(method = "segreg", 
                 formula = my.formula)+
  geom_point() +
  stat_poly_eq(aes(label = 
                     after_stat(
                       paste("x[cp]~`=`~", 
                             round(knots[[1]][1], 2), " %+-% ",
                             round(knots.se[[1]][1], 2), "*\"; L: \"*y~`=`~", 
                             round(b_0, 2), # "+",
                             round(b_1, 2), "~x*\"; R: \"*y~`=`~", 
                             round(b_0 + knots[[1]][1] * (b_1 - b_2), 2),"+",
                             round(b_2, 2), "~x",
                             sep = ""))), 
               parse = TRUE, 
               output.type = "numeric",
               label.x = "right",
               formula = my.formula,
               method = "segreg") +
  stat_poly_eq(use_label("R2", "P", "AIC"), 
               label.x = "right",
               label.y = 0.87,
               formula = my.formula,
               method = "segreg")
```

In the next example the slope of the second segment is fixed at zero to keep it
horizontal using `formula = y ~ seg(x, npsi = 1, est = c(1, 0))` where 1
indicates a fitted slope and 0 a slope fixed at this value. Because of how the
parameters are returned, the code to assemble the equation needs to be
rewritten.

```{r, fig.width = 7, fig.height = 5}
my.formula <- y ~ seg(x, npsi = 1, est = c(1, 0))

ggplot(mpg, aes(displ, hwy)) +
  stat_poly_line(method = "segreg", 
                 formula = my.formula) +
  geom_point() +
  stat_poly_eq(aes(label = 
                     paste("x[cp]~`=`~",
                           round(after_stat(knots)[[1]][1], 2),
                           " %+-% ",
                           round(after_stat(knots.se)[[1]][1], 2),
                           "*\"; L: \"*y~`=`~", 
                           round(after_stat(b_1), 2), # "+",
                           round(after_stat(b_2), 2), "~x",
                           "*\"; R: \"*y~`=`~", 
                           round(after_stat(b_1) + 
                                   after_stat(knots)[[1]][1] *
                                   after_stat(b_2),
                                 2),
                           sep = "")), 
               parse = TRUE, output.type = "numeric",
               label.x = "right",
               formula = my.formula,
               method = "segreg") +
  stat_poly_eq(use_label("R2", "P", "AIC"), 
               label.x = "right",
               label.y = 0.87,
               formula = my.formula,
               method = "segreg")
```

Highlighting fitted values.

```{r, fig.width = 7, fig.height = 5}
ggplot(mpg, aes(displ, hwy)) +
  stat_poly_line(method = "segreg", 
                 formula = my.formula) +
  geom_point() +
  stat_fit_fitted(method = "segreg", 
                 formula = my.formula,
                 shape = "cross")
```

Plot of residuals.

```{r, fig.width = 7, fig.height = 5}
ggplot(mpg, aes(displ, hwy)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  stat_fit_residuals(method = "segreg", 
                      formula = my.formula)
```

### Natural cubic splines ('splines')

**Note:** The lines for cubic and other smoothing splines can be also added to plots
with `stat_smooth()` from package 'ggplot2'.

```{r}
print(citation(package = "splines", auto = TRUE), bibtex = FALSE)
```

```{r}
library(splines)
```

Standard package 'splines', part of the R distribution, provides functions that can be used to fit splines with function `lm()`. These functions are called from the model formula. Functions, `bs()` amd `ns()` can be used
directly in the model formula passed to `stat_poly_line()`. They can be used with `stat_poly_eq()` but the equation label returned is incorrect. The other labels are correctly generated.

```{r, fig.width = 7, fig.height = 5}
cp <- 4.1 # a guess
my.formula <- y ~ splines::ns(x, knots = cp)

ggplot(mpg, aes(displ, hwy)) +
  stat_poly_line(method = "lm", 
                 formula = my.formula) +
  geom_point() +
  stat_poly_eq(use_label("R2", "P", "AIC"), 
               label.x = "right",
               formula = my.formula,
               method = "lm")
```

Plotting fitted values and observations.

```{r, fig.width = 7, fig.height = 5}
ggplot(mpg, aes(displ, hwy)) +
  stat_poly_line(method = "lm", 
                      formula = my.formula) +
  geom_point() +
  stat_fit_fitted(method = "lm", formula = my.formula, 
                  shape = "cross", alpha = 0.5)
```

Plotting residuals.

```{r, fig.width = 7, fig.height = 5}
ggplot(mpg, aes(displ, hwy)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  stat_fit_residuals(method = "lm", 
                     formula = my.formula) +
  labs(y = "Residuals")
```

It is possible to use `rlm()` or `lmrob()` instead of `lm()`.

```{r, fig.width = 7, fig.height = 5}
cp <- 4.1
my.formula <- y ~ splines::ns(x, knots = cp)

ggplot(mpg, aes(displ, hwy)) +
  stat_fit_deviations(formula = my.formula,
                      method = "lmrob",
                      mapping = aes(colour = 
                                      after_stat(robustness.weights)),
                      show.legend = TRUE,
                      geom = "point") +
  stat_poly_eq(use_label("R2", "n", "method"), 
               label.x = "right",
               formula = my.formula,
               method = "lmrob") +
  stat_poly_line(method = "lmrob", 
                 formula = my.formula) +
  scale_color_gradient(low = "red", high = "blue", limits = c(0, 1)) +
  theme(legend.position = "top")
```

Plotting residuals.

```{r, fig.width = 7, fig.height = 5}
ggplot(mpg, aes(displ, hwy)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  stat_fit_residuals(method = "rlm", 
                     formula = my.formula,
                     mapping = aes(colour = after_stat(robustness.weights)),
                     show.legend = TRUE) +
  scale_color_gradient(name = "Weights", 
                       low = "red", high = "blue", 
                       limits = c(0, 1),
                       guide = "colourbar") +
  labs(y = "Residuals")
```

### Polynomial splines ('splines')

**Note:** The lines for cubic and other smoothing splines can be also added to plots
with `stat_smooth()` from package 'ggplot2'.

```{r}
print(citation(package = "splines", auto = TRUE), bibtex = FALSE)
```

```{r}
library(splines)
```

Standard package 'splines', part of the R distribution, provides functions that can be used to fit splines with function `lm()`. These functions are called from the model formula. Functions, `bs()` amd `ns()` can be used
directly in the model formula passed to `stat_poly_line()`. They can be used with `stat_poly_eq()` but the equation label returned is incorrect. The other labels are correctly generated.

```{r, fig.width = 7, fig.height = 5}
cp <- 4.1 # a guess
my.formula <- y ~ splines::bs(x, knots = cp)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  stat_poly_eq(use_label("R2", "P", "AIC"), 
               label.x = "right",
               formula = my.formula,
               method = "lm") +
  stat_poly_line(method = "lm", 
                 formula = my.formula)
```


Plotting residuals.

```{r, fig.width = 7, fig.height = 5}
ggplot(mpg, aes(displ, hwy)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  stat_fit_residuals(method = "lm", formula = my.formula) +
  labs(y = "Residuals")
```

It is posisble to use `rlm()` or `lmrob()` instead of `lm()`.

```{r, fig.width = 7, fig.height = 5}
cp <- 4.1
my.formula <- y ~ splines::bs(x, knots = cp)

ggplot(mpg, aes(displ, hwy)) +
  stat_fit_deviations(formula = my.formula,
                      method = "lmrob",
                      mapping = aes(colour = after_stat(robustness.weights)),
                      show.legend = TRUE,
                      geom = "point") +
  stat_poly_eq(use_label("R2", "n", "method"), 
               label.x = "right",
               formula = my.formula,
               method = "lmrob") +
  stat_poly_line(method = "lmrob", 
                 formula = my.formula) +
  scale_color_gradient(low = "red", high = "blue", limits = c(0, 1)) +
  theme(legend.position = "top")
```


Plotting residuals.

```{r, fig.width = 7, fig.height = 5}
ggplot(mpg, aes(displ, hwy)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  stat_fit_residuals(method = "lmrob", 
                     formula = my.formula,
                     mapping = aes(colour = after_stat(robustness.weights)),
                     show.legend = TRUE) +
  scale_color_gradient(name = "Weights", 
                       low = "red", high = "blue", 
                       limits = c(0, 1),
                       guide = "colourbar") +
  labs(y = "Residuals")
```

## User defined functions

As mentioned above, layer functions `stat_poly_line()` and `stat_poly_eq()` support the use of several different model fit functions, as long as the estimates and predictions can be extracted. In the case of `stat_poly_eq()`, the model formula embedded in this object must be a polynomial. The variables mapped to _x_ and _y_ aesthetics must be continuous numeric, times or dates. Models are fitted separately to each group and plot panel, similarly as by `ggplot2::stat_smooth()`. 'ggpmisc' provides additional flexibility by delaying some decisions until after the models have been fit to the observations.

**`stat_poly_line()` and `stat_poly_eq()` adjust their behaviour based on the class of the model object returned by the method, not the name of the method passed in the call as argument.** Thus, the class of the model fit object returned can vary in response to data, and thus also be different for different panels or groups in the same plot. 

**`stat_poly_eq()` builds the fitted model equation based on the model formula extracted from the model fit object, which can differ from that passed as argument to `formula` when calling the statistic.** Thus, the model formula returned by the function passed as argument for `method` can vary by group and/or panel. As the examples below demonstrate, this approach can be very useful. In addition, `NA` as return value from the function passed as argument to `method` can be used to signal to `stat_poly_line()` and `stat_poly_eq()` that no model has been fit for a group or panel.

Below I use `lm()` and `"lm"` model fit objects in most examples, but the same approaches can be applied to objects of other classes, including `"rlm"`, `"lmrob"`, '"lqs"`, `"lts"`, and `"gls"`.

### Fit or not

In the first example we fit a model only if a minimum number of distinct _x_ values are present in `data`. We define a function that instead of fitting the model, returns `NA` when the condition is not fulfilled. Here `"x"` is the aesthetic corresponding to the explanatory variable, and has to be replaced by `"y"` if `orientation = y` is passed as argument to the statistic.

```{r}
fit_or_not <- function(formula, data, ..., orientation = "x") {
  if (length(unique(data[[orientation]])) > 5) {
    lm(formula = formula, data = data, ...)
  } else {
    NA_real_
  }
}
```

Instead of using `lm()` as method, we pass our wrapper function `fit_or_not()` as method. As the function returns either an `"lm"` object from the wrapped call to `lm()` or `NA`, the call to the statistics can make use of all other features as needed.

```{r}
#| fig.asp: 1.25

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  stat_poly_line(method = "fit_or_not") +
  stat_poly_eq(method = fit_or_not, 
               mapping = use_label("eq", "R2"),
               label.x = "right") +
  stat_panel_counts(label.x = "left", label.y = "bottom") +
  theme(legend.position = "bottom") +
  facet_wrap(~class, ncol = 2)
```

```{r}
#| fig.asp: 1.25

ggplot(mpg, aes(displ, hwy)) +
  stat_fit_residuals(method = "fit_or_not") +
  facet_wrap(~class, ncol = 2)
```

### Degree of polynomial

Not fitting a model, is a drastic solution. An alternative is implemented next, to plot the mean when the slope of the linear regression is not significantly different from zero. In the example below instead of using `stats::lm()` as method, we define a different wrapper function that tests for the significance of the slope in linear regression, and if not significant, fits the mean instead. _This works because the model formula is extracted from the fitted model rather than using the argument passed by the user in the call to the statistics._


_Fitting different models to different panels is supported._ User-defined method functions are required to return an object that inherits from class `"lm"` or of another supported method. The function is applied per group and panel, and the model `formula` fitted and can differ among them, as it is retrieved from this object to construct the equation label.

In the example below instead of using `stats::lm()` as method, a wrapper function is defined and used, which tests for the significance of the slope in linear regression, and if not significant, fits the mean instead. 

```{r}
poly_or_mean <- function(formula, data, ...) {
   fm <- lm(formula = formula, data = data, ...)
   if (anova(fm)[["Pr(>F)"]][1] > 0.1) {
      lm(formula = y ~ 1, data = data, ...)
   } else {
      fm
   }
}
```

Function `poly_or_mean()`, just defined, or its name is paased as the argument for `method`.

```{r}
#| fig.asp: 1.25

ggplot(mpg, aes(displ, hwy)) +
   geom_point() +
   stat_poly_line(method = "poly_or_mean") +
   stat_poly_eq(method = poly_or_mean,
               mapping = use_label("eq", "R2"),
               label.x = "right") +
   facet_wrap(~class, ncol = 2)
```

```{r}
#| fig.asp: 1

ggplot(mpg, aes(displ, hwy, color = class)) +
   geom_point() +
   stat_poly_line(method = "poly_or_mean") +
   stat_poly_eq(method = poly_or_mean,
               mapping = use_label("eq"),
               label.x = "right") +
   theme(legend.position = "top")
```

A different approach to selecting the degree of a polynomial is to use stepwise selection based on AIC. In this case the `formula` passed as argument is the "upper" degree of the polynomial. All lower degree polynomials are fitted and the one with lowest AIC used.

```{r}
best_poly_lm <- function(formula, data, ...) {
  poly.term <- as.character(formula)[3]
  degree <- as.numeric(gsub("poly[(]x, |, raw = TRUE|[)]", "", poly.term))
  fms <- list()
  AICs <- numeric(degree)
  for (d in 1:degree) {
    # we need to define the formula with the value of degree replaced
    working.formula <- as.formula(bquote(y ~ poly(x, degree = .(d), raw = TRUE)))
    fms[[d]] <- lm(formula = working.formula, data = data, ...)
    AICs[d] <- AIC(fms[[d]])
  }
  fms[[which.min(AICs)[1]]] # if there is a tie, we take the simplest
}
```

We then use our function `best_poly_lm()` as the method.

```{r}
#| fig.asp: 1.25

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  stat_poly_line(formula = y ~ poly(x, 3, raw = TRUE), 
                 method = "best_poly_lm") +
  stat_poly_eq(formula = y ~ poly(x, 3, raw = TRUE), 
               method = "best_poly_lm",
               mapping = use_label("eq"),
               label.x = "right") +
  expand_limits(y = c(0, 65)) +
  facet_wrap(~class, ncol = 2)
```

```{r}
#| fig.asp: 1

ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_point() +
  stat_poly_line(formula = y ~ poly(x, 3, raw = TRUE), 
                 method = "best_poly_lm") +
  stat_poly_eq(formula = y ~ poly(x, 3, raw = TRUE), 
               method = "best_poly_lm",
               mapping = use_label("eq", "AIC"),
               label.x = "right",
               vstep = 0.035) +
  expand_limits(y = 45) +
  theme(legend.position = "top")
```

**Note:** Other approaches are possible because additional arbitrary arguments can be passed by name to the method function. Thus a more flexible approach
would be to use R's function `step()` in the wrapper.

### Functions that re-fit a fitted model

The examples in this section are for package 'segmented', however, a similar
approach can be useful with any model fitting that involves calls to more than
one function in sequence, such as computing starting values for a numerical
method, or modifying the model fit object.

### Segmented ('segmented')

Package 'segmented' implements methods for fitting models with change points. The location of the change points themselves are fitted with estimates and their standard deviations computed.

```{r}
print(citation(package = "segmented", auto = TRUE), bibtex = FALSE)
```

```{r}
library(segmented)
```

The code here, is equivalent to that shown above based on model-fit function
`segreg()` also from package 'segmented'. The difference is that when using
`segmented()` the fitting is done in two steps. In this case there is no need 
to use this approach, but it can be useful when segmentation is to be applied 
only to some plot panels or applied conditionally in single panel plots. Another
use is when fitting segmented linear regressions based on model-fit functions
other than `lm()` or `glm()`. However, in this second case, some of the
necessary model-fit query functions can be unavailable and may need to be
defined by the user. There are also model fit methods that do not compute or
provide the estimates required by the segmentation computations.

Package 'segmented' provides several specializations of method `segmented()`.
These methods are dispatched according to the model fit object passed to
parameter `obj`. As this package provides all the expected query functions for
the object `segmented()` methods return, their use is not too difficult. It is
possible to write a wrapper function to pass as argument to `stat_poly_line()`
and `stat_poly_eq()` and in addition, if used, construct the fitted equation
labels within a call to `aes()`. This function, that I named lm2segmented.lm,
calls first `lm()` and then calls `segmented()` with the fitted model as
argument. It accepts additional arguments that make it possible to control the
number and initial values of the knots. As in the case of other model fit
functions, models with multiple explanatory variables are rarely used with
'ggplot2' stats.

```{r, fig.width = 7, fig.height = 5}
lm2segmented.lm <- 
  function(formula,
           data, 
           seg.Z = ~ x, # explanatory variable used for segmentation
           psi = median(data$x), # starting value(s) for knot(s)
           npsi = 1, # number of knots if psi is not given
           fixed.psi = NULL, # breakpoint not to be modified by fitting
           control = seg.control(), 
           ...) {
    obj <- lm(formula = formula, data = data, ...)
    segmented(obj = obj, 
              seg.Z = seg.Z, 
              psi = psi, 
              npsi = npsi, 
              fixed.psi = fixed.psi, 
              control = control, 
              model = TRUE, 
              keep.class = FALSE)
  }
```

```{r, fig.width = 7, fig.height = 5}
# using defaults, one breakpoint
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  stat_poly_line(method = lm2segmented.lm)
```

```{r, fig.width = 7, fig.height = 5}
ggplot(mpg, aes(displ, hwy)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  stat_fit_residuals(method = lm2segmented.lm) +
  labs(y = "Residuals")
```

```{r, fig.width = 7, fig.height = 5}
# requesting two breakpoints with starting values
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  stat_poly_line(method = lm2segmented.lm,
                 method.args = list(psi = c(3, 5)))
```

A more complex example with labels, assembled within the call to `aes()`. In addition to the intercept and slope for the leftmost segment, it returns the slope for each segment to the right of the first one, but not their intercepts. As `segment()` fits and returns the position of the knots, which are also returned by
`stat_poly_eq()` in 'ggpmisc' (>= 0.6.3). A label is added with the knot position, and two regression equations, one for each segment of the linear spline.

```{r, fig.width = 7, fig.height = 5}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  stat_poly_eq(aes(label =
                     after_stat(
                       paste("italic(x)[cp]~`=`~", 
                             round(knots[[1]][1], 2), "*\"; L: \"*y~`=`~", 
                             round(b_0, 2), #"+",
                             round(b_1, 2), "~x*\"; R: \"*y~`=`~", 
                             round(b_0 + knots[[1]][1] * (b_1 - b_2), 2), "+",
                             round(b_2, 2), "~x",
                             sep = ""))), 
               parse = TRUE, output.type = "numeric",
               label.x = "right",
               formula = y ~ x,
               method = lm2segmented.lm) +
  stat_poly_eq(use_label("R2", "P", "AIC"), 
               label.x = "right",
               label.y = 0.87,
               formula = y ~ x,
               method = lm2segmented.lm,
               method.args = list(psi = c(3, 5))) +
  stat_poly_line(method = lm2segmented.lm, 
                 formula = y ~ x,)
```

## Work-arounds

Some model-fit functions return objects for which no `predict()` method exist. In such cases, as users we can define such a method. More difficult situations can also be faced. 

### Errors in variables ('refitME')

In this section I show a situation I faced when trying to use package 'refitME'. The functions exported by this package fail when called from within a function, so they do not currently work if passed as an argument for `method`. Below, a temporary work-around based on a solution provided by the maintainer of 'refitME' is shown.

```{r}
print(citation(package = "refitME", auto = TRUE), bibtex = FALSE)
```

```{r}
library(refitME)
```

As 'refitME' functions return a model fit object of the same class as that received as input, a model fitted with `lm()` could be, in principle, refit in a user-defined wrapper function as those above, calling first `lm()` and re-fitting the obtained model with a `refitME::refitME()` as shown in the code chunk below. However, at the moment the functions in package 'refitME' fail when called from within another function.

```{r, eval=FALSE}
lm_EIV <- function(formula, data, ..., sigma.sq.u) {
  fm <- lm(formula = formula, data = data, ...)
  MCEMfit_glm(mod = fm, family = "gaussian", sigma.sq.u = sigma.sq.u)
}
```

After defining such a function, we could then use our function `lm_EIV()` as the method.

```{r, eval=FALSE}
ggplot(mpg, aes(displ, hwy)) +
   geom_point() +
   stat_poly_line(method = "lm_EIV", method.args = c(sigma.sq.u = 0)) +
   stat_poly_eq(method = "lm_EIV", method.args = c(sigma.sq.u = 0),
                mapping = use_label("eq"),
   label.x = "right") +
   theme(legend.position = "bottom") +
   facet_wrap(~class, ncol = 2)
```

However, it prints one warning like the one below for each call of the statistics.

```
Warning: Computation failed in `stat_poly_eq()`.
Caused by error in `names(new.dat)[c(1, which(names(mod$data) %in% names.w))] <- names(mod$data)[c(1, which(names(mod$data) %in% names.w))]`:
! replacement has length zero```
```

I have reported the problem through a [GitHub issue](https://github.com/JakubStats/refitME/issues/5) to Jakub Stoklosa, the maintainer of 'refitME', and he is looking for a permanent solution. Meanwhile, he has provided a temporary fix that I copy below with a few small edits to the ggplot code. Many thanks to Jakub Stoklosa for his very fast answer!

```{r}
rm(list = ls(pattern = "*"))

lm_EIV <- function(formula, data, ..., sigma.sq.u) {
  assign("my.data", value = data, envir = globalenv())
  fm <- lm(formula = formula, data = my.data, ...)
  fit <- refitME(fm, sigma.sq.u)
  
  fit$coefficients <- Re(fit$coefficients)
  fit$fitted.values <- Re(fit$fitted.values)
  fit$residuals <- Re(fit$residuals)
  fit$effects <- Re(fit$effects)
  fit$linear.predictors <- Re(fit$linear.predictors)
  fit$se <- Re(fit$se)
  
  if ("qr" %in% names(fit)) {
  fit$qr$qr <- Re(fit$qr$qr)
  fit$qr$qraux <- Re(fit$qr$qraux)
  }
  
  class(fit) <- c("lm_EIV", "lm")
  return(fit)
}

predict.lm_EIV <- function(object, newdata = NULL, ...) {
  preds <- NextMethod("predict", object, newdata = newdata, ...)
  
  if (is.numeric(preds)) {
    return(Re(preds))
  } else if (is.matrix(preds)) {
    return(Re(preds))
  } else {
    return(preds) # Leave unchanged if not numeric/matrix
  }
}
```

Here the limitation of 'ggplot2' and 'ggpmisc' in that the fit is recomputed in each layer causes a further difficulty, as with most algorithms using random numbers, the successive refits do not necessarily produce the exact same results. In this case, the equation and line are not matched, and the mismatch can be large for large values of `sigma.sq.u` and low replication, especially in the presence of outliers. 

This is something that can most likely be fixed in 'ggpmisc' by setting the seed of the RNG. To check that work-around is working as expected, we set `sigma.sq.u = 0` effectively using an OLS approach and expecting values nearly identical as with `lm()`.

```{r, message=FALSE}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  stat_poly_line(aes(y = hwy),
                 method = "lm_EIV",
                 method.args = list(sigma.sq.u = 0)) +
  stat_poly_eq(aes(y = hwy, 
                   label = 
                     after_stat(
                       paste(eq.label, rr.label, sep = "~~"))),
               method = "lm_EIV",
               method.args = list(sigma.sq.u = 0),
               label.x = "right") +
  facet_wrap(~class, ncol = 2)
```

We do not know the measurement error for the measurement of `displ`, the displacement volume of the engines. For this example we make up a value: `sigma.sq.u = 0.01` expecting estimates different from above and with `lm()`.
 
```{r, message=FALSE}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  stat_poly_line(aes(y = hwy),
                 method = "lm_EIV",
                 method.args = list(sigma.sq.u = 0.01)) +
  stat_poly_eq(aes(y = hwy, 
                   label = 
                     after_stat(
                       paste(eq.label, rr.label, sep = "~~"))),
               method = "lm_EIV",
               method.args = list(sigma.sq.u = 0.01),
               label.x = "right") +
  facet_wrap(~class, ncol = 2)
```

## New data for prediction

Like in `stat_smooth()` from 'ggplot2', parameter `fullrange` in `stat_poly_line()`, `stat_quant_line()`, `stat_quant_band()` and `stat_ma_line()` can be used to switch between prediction based on the range of the explanatory variable data or based on the limits of the scale of the corresponding aesthetic. Meanwhile, `n` controls the number of equally spaced points at which the prediction will be computed. In most use cases these two alternatives are enough.

However, in some cases, finer control is useful. In plots with facets, with major axis (MA) regression and possibly other model fits, it can be useful to constrain the range of the prediction to that of the response variable. Furthermore, in MA regression, the fitted line is the same for `y ~ x` and `x ~ y` , thus simultaneously restricting the range based on both `x` and `y` seems logically preferable. Another case that cannot be automated with `stat_mooth()` are predictions covering user-defined ranges or for specific values of the explanatory variable.

In 'ggpmisc' (>= 1.0.0) the new `limit.to` parameter provides the necessary control. The `character` strings `"x"`, `"y"`, `"xy"` and `"none"` make it possible for user to set which data range to use as limit.

For these examples I expanded the `x` scale limits, so that they are clearly broader than the default ones. These are rather contrived examples demonstrating how fine control works.

The first example, with `limit.to = "none"` produces the same plot as using `fullrange = TRUE`.

```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  geom_point() +
  stat_poly_line(limit.to = "none") +
  expand_limits(x = c(-2:10))
```

With `limit.to = "y"` the prediction is computed as in the example above, but predicted values aoutside the range of `y` are not passed to the geom, and, thus, not plotted.

```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  geom_point() +
  stat_poly_line(limit.to = "y") +
  expand_limits(x = c(-2:10))
```

In addition to character strings as described above, `limit.to` also accepts as argument a numeric vector to be used as new data. This makes it possible to obtain a prediction for a range that is neither that of the scale or that of the data. A use case is when an unrestricted range could interfere with a plot annotation.

```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  geom_point() +
  stat_poly_line(limit.to = seq(from = 2, to = 7, length.out = 80)) +
  expand_limits(x = c(1:8))
```

Although rarely useful when continuous variables are mapped to both `x` and `y`, it is also possible to compute predictions for discrete values of the explanatory variable. In this example the predictions extend past the range of the data, and the range of the limits of the scales are automatically extended for them to fit. Instead of the default `geom = "smooth"`, `geom = "pointrange"` displays the individual predictions. 

```{r, fig.width = 7, fig.height = 5}
ggplot(subset(mpg, class == "suv"), aes(displ, hwy)) +
  geom_point() +
  stat_poly_line(limit.to = 2:7,
                 geom = "pointrange",
                 colour = "blue")
```

## Transformations and `eq.label`

In 'ggplot2' setting transformations in _x_ and _y_ scales results in statistics receiving transformed data as input. The data "seen" by statistics is the same as if the transformation had been applied within a call to `aes()`. However, the scales
automatically adjust breaks and tick labels to display the values in the untransformed scale. When the transformation is applied in the model `formula`
definition, the transformation is applied later, but still before the model is fit. all these cases, the model fit is based on transformed values and thus the
estimates refer to them. This is why, a linear fit remains linear in spite of the transformations. 

Only `eq.label` is affected by transformations in the sense that the _x_ and _y_ in the default equation refer to transformed values, rather than those shown in the tick labels of the axes, i.e., the default model equation label is incorrect. The correction to apply is to override the _x_ and/or _y_ symbols in the equation label by an expression that includes the transformation used. In these examples we
fit polynomials by OLS using `lm()`, the default method.

```{r, fig.width = 7, fig.height = 5}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  stat_poly_line() +
  stat_poly_eq(mapping = use_label("eq", "R2", "n", "P"),
             eq.with.lhs = "log[10]~hwy~`=`~",
             eq.x.rhs = "~displ") +
  scale_y_log10() + 
  expand_limits(y = 50)
```

```{r, fig.width = 7, fig.height = 5}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  stat_poly_line(formula = y ~ poly(x, 3, raw = TRUE)) +
  stat_poly_eq(mapping = use_label("eq"),
               formula = y ~ poly(x, 3, raw = TRUE),
               eq.with.lhs = "hwy~`=`~",
               eq.x.rhs = "~(log[10]~displ)") +
  scale_x_log10(name = "displ") + 
  expand_limits(y = 50)
```
